---
title: "Tryggve2 - schizophrenia - Sweden (part 2)"
author: "Kaarina Kowalec, kaarina.kowalec@ki.se"
date: "November 2018"
output: 
  html_document: 
    fig_caption: yes
    theme: simplex
    highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Motivation and Overview

NOTE: NOT RUN

Examine variables associated with poor outcomes in schizophrenia, using first, the Swedish National Registers. In this report, I will present some initial coding for the analyses.

**Overview**

1. Import data
2. Create a dataframe with all necessary variables including exposures and outcomes
3. Analyses

**Exposures or Predictors**
1) Family of origin:
+ parental ages at birth ##LY
+ parental SES-education-occupation-income
+ birth order
+ family composition (number of parents/partners, parental criminality-incarceration, divorce, siblings, out-of-home placement)

2) Childhood trauma: ##LY
+ parental death
+ severe medical illness of parents/sibs

3) Education: done-kk
+ educational attainment
+ standardized testing in grade 9

4) Prior personal medical history:
+ comorbid disorders and medication use ##kk
+ prior personal medical history (somatic disorders) #kk
+ childhood hospitalization

5) Family history of psychiatric disorder ##kk
+ MD, BIP, psychosis, suicide in first-degree relatives

6) Genetic risk factors  NOT NOW
+ GRS of MDD, SCZ, BIP, IQ etc.

**Outcomes**
+ the number of psychiatric hospitalizations and outpatient contacts 
+ the number of somatic healthcare utilization
+ indication of 2nd or 3rd line treatments (pharmacological augmentation, ECT, TMS, DBS, clozapine) 
+ suicide or suicide attempt

**TO DOs**

cut down this doc into 4 separate RMD (eventually):
outcome
exposure/risk factors
basics
analysis

***

### Step 0: Load R packages
```{r R packages}
library(dplyr)
library(data.table)
library(ggplot2)
library(lubridate)

Sys.setenv(TZ="Europe/Stockholm")
```

### Step 1: Load data
REQUIRE EDITING on the file path
```{r specify file path}
# edit the file path where you have stored the part 1 data
data.f <- "/Users/luyi/CloudStation/WORK_2018/MyProjects_with_Pat/GAPS_MDD/DATA/FREED/DATA/scz.part1.RData"

# load the R data
load(data.f) 
```

Per-diagnoses to per-individual 
```{r collapse to per-individual}
# 1. before collapsing the records, apply 2 criteria: 
#    1) refine ICD codes
    # ICD-8: 295, 2950-2954, 2956-2959, i.e. to remove 2955 (latent SCZ)
    # ICD-9: 295A-295E, 295G, 295H, 295W, 295X, i.e. just drop 295F (latent SCZ, other codes do not exit)
    # ICD-10: F20 (SCZ), F25 (SAD), F231 (Acute polymorphic psychotic disorder with symptoms of schizophrenia) & F232 (Acute schizophrenia-like psychotic disorder)
#    2) ≥2 diagnoses 

# select specific ICD codes
scz.v2 <- scz %>% 
  filter(
    (icd.version!=10 & grepl("^295$|^2950|^2951|^2952|^2953|^2954|^2956|^2957|^2958|^2959|^295A|^295B|^295C|^295D|^295E|^295G|^295H|^295W|^295X", diagnosis)) | 
    (icd.version==10 & grepl("^F20$|^F200|^F201|^F202|^F203|^F204|^F205|^F206|^F208|^F209|^F25$|^F250|^F251|^F252|^F258|^F259|^F231|^F232", diagnosis)) # apply the criteria #1
  )

# collapse to per-individual record and not per-admission
scz.perind <- scz.v2 %>% 
  arrange(id,scz_date) %>% 
  group_by(id) %>%
  slice(1) %>%
  filter(nsczcontacts >= 2) %>% # apply the criteria #2
  select(-admit.date:-icd.version, -scz_date) %>%
  ungroup()

# make list of scz cases and date of first contact for comparison with psych disorders below
first.scz.date <- scz.perind %>%
  select(id, first_scz_date, dob)
  
  
```

### Step 2: Exposure-Prior psychiatric history
Need to start with making a list of any one with a psych disorder in Sweden because will need
for family history of psychiatric disorders.

We are only selecting those with a date (1) before first treatment contact and (2) before age 15 years (indicating childhood)

+ comorbid psych disorders/conditions: psych (adhd, asd, anx, bip, ed, mdd, ocd, ptsd, scz, suds, suicides)
+ medication use (wait on this-not sure which drugs to focus on)

1. ADHD
```{r ADHD}
# first extract all individuals with ADHD diagnosis 
adhd <- hdr %>% 
  filter(  (icd.version==9 & grepl("^314|^314A|^314B|^314C|^314J|^314W|^314X", diagnosis)) # no icd codes for ADHD prior to ICD version 9
           | (icd.version==10 & grepl("^F90|^F900|^F901|^F908", diagnosis)) 
  ) %>%
  mutate(adhd = 1,
         admit.date = ymd(admit.date)) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  add_count(id) %>%  # record the number of diagnoses per individual into variable "n"
  rename(nadhdcontacts = n) %>% # rename n 
  mutate(first.adhd.date = first(admit.date)) %>% # record the first ADHD diagnosis date 
  slice(1) %>% 
  select(id,adhd,nadhdcontacts, first.adhd.date)

# then merge with individuals with SCZ 
scz.adhd <- inner_join(first.scz.date,adhd, by="id") %>% # inner join together with scz cases
  mutate(age.first.adhd = as.integer((first.adhd.date - dob)/365.25),
         adhd.childhood = ifelse(age.first.adhd <= 15, 1, 0), # only those with adhd dx before age 15 years
         adhd.pre.scz = ifelse(first.adhd.date < first_scz_date, 1, 0)  #  only those with adhd dx before scz contact
         ) %>%
  select(id,nadhdcontacts, adhd.childhood,adhd.pre.scz)

```
2. ASD
```{r ASD}
# first extract all individuals with ASD diagnosis 
asd <- hdr %>% 
  filter(  (icd.version==9 & grepl("^299A", diagnosis)) # no icd codes for asd prior to version 9
           | (icd.version==10 & grepl("^F840|^F841|^F845", diagnosis)) 
  ) %>%
  mutate(asd = 1,
         admit.date = ymd(admit.date)) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  add_count(id) %>%  # record the number of diagnoses per individual into variable "n"
  rename(nasdcontacts = n) %>% # rename n 
  mutate(first.asd.date = first(admit.date)) %>% # record the first diagnosis date
  slice(1) %>%
  select(id,asd,nasdcontacts,first.asd.date)

# then merge with individuals with SCZ 
scz.asd <- inner_join(first.scz.date,asd, by="id") %>% # inner join together with scz cases
  mutate(age.first.asd = as.integer((first.asd.date - dob)/365.25),
         asd.childhood = ifelse(age.first.asd <= 15, 1, 0), # only those with dx before age 15 years
         asd.pre.scz = ifelse(first.asd.date < first_scz_date, 1, 0)  #  only those with dx before scz contact
         ) %>%
  select(id, nasdcontacts, asd.childhood, asd.pre.scz)

```
3. Anxiety
```{r Anxiety}
# first extract all individuals with anxiety diagnosis 
anx <- hdr %>% 
  filter(  (icd.version==8 & grepl("^300|^3000|^3002", diagnosis)) |
             (icd.version==9 & grepl("^300|^300A|^300C", diagnosis)) |
             (icd.version==10 & grepl("^F40|^F400|^F401|^F402|^F408|^F409|^F41|^F410|^F411|^F412|^F413|^F418|^F419", diagnosis)) 
  ) %>%
  mutate(anx = 1,
         admit.date = ymd(admit.date)) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  add_count(id) %>%  # record the number of diagnoses per individual into variable "n"
  rename(nanxcontacts = n) %>% # rename n 
  mutate(first.anx.date = first(admit.date)) %>%
  slice(1) %>%
  select(id, anx, nanxcontacts, first.anx.date)

# then merge with individuals with SCZ 
scz.anx <- inner_join(first.scz.date,anx, by="id") %>%  # inner join together with scz cases
  mutate(age.first.anx = as.integer((first.anx.date - dob)/365.25),
         anx.childhood = ifelse(age.first.anx <= 15, 1, 0), # only those with anx dx before age 15 years
         anx.pre.scz = ifelse(first.anx.date < first_scz_date, 1, 0)  #  only those with anx dx before scz contact
  ) %>%
  select(id, nanxcontacts, anx.childhood, anx.pre.scz)

```
4. Bipolar
```{r Bipolar}
# first extract all individuals with bipolar diagnosis 
bip <- hdr %>%
  filter( ((icd.version==8) & grepl('^2961|^2963|^2968',diagnosis)) |
          ((icd.version==9) & grepl('^296|^269A|^269C|^2691|^296D|^2694|^296E|^2965|^296F|^2966|^296G|^2697|^269H|^2691|^26989|^269W|^26999|^269X',diagnosis)) | 
            ((icd.version==10) & grepl('^F30$|^F301|^F302|^F308|^F309|^F31|^F310|^F311|^F312|^F313|^F314|^F315|^F316|^F317|^F318|^F319',diagnosis)) ) %>%
  mutate(bip = 1,
         admit.date = ymd(admit.date)) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  add_count(id) %>%  # record the number of diagnoses per individual into variable "n"
  rename(nbipcontacts = n) %>% # rename n 
  mutate(first.bip.date = first(admit.date)) %>%
  slice(1) %>%
  select(id, bip, nbipcontacts, first.bip.date)

scz.bip <- inner_join(first.scz.date,bip, by="id") %>%# inner join together with scz cases
  mutate(age.first.bip = as.integer((first.bip.date - dob)/365.25),
         bip.childhood = ifelse(age.first.bip <= 15, 1, 0), # only those with bip dx before age 15 years
         bip.pre.scz = ifelse(first.bip.date < first_scz_date, 1, 0)  #  only those with bip dx before scz contact
  ) %>%
  select(id, nbipcontacts, bip.childhood, bip.pre.scz)
```
5. Eating disorder
```{r ED}
# first extract all individuals with eating disorder diagnosis 
ed <- hdr %>%
  filter(  (icd.version==8 & grepl("^7840|^3065", diagnosis))
           | (icd.version==9 & grepl("^307B|^307F", diagnosis))
           | (icd.version==10 & grepl("^F50|^F500|^F501|^F502|^F503|^F509", diagnosis))) %>%
  mutate(ed = 1,
         admit.date = ymd(admit.date)) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  add_count(id) %>%  # record the number of diagnoses per individual into variable "n"
  rename(nedcontacts = n) %>% # rename n 
  mutate(first.ed.date = first(admit.date)) %>%
  slice(1) %>%
  select(id, ed, nedcontacts, first.ed.date)

scz.ed <- inner_join(first.scz.date,ed, by="id") %>% # inner join together with scz cases
  mutate(age.first.ed = as.integer((first.ed.date - dob)/365.25),
         ed.childhood = ifelse(age.first.ed <= 15, 1, 0), # only those with ed dx before age 15 years
         ed.pre.scz = ifelse(first.ed.date < first_scz_date, 1, 0)  #  only those with ed dx before scz contact
  ) %>%
  select(id, nedcontacts, ed.childhood, ed.pre.scz)
```
6. Major depression
```{r MDD}
# first extract all individuals with depression diagnosis 
mdd <- hdr %>% 
  filter(  (icd.version==8 & grepl("^2960|^3004", diagnosis)) |
            (icd.version==9 & grepl("^296B|^311", diagnosis)) |
             (icd.version==10 & grepl("^F32|^F320|^F321|^F322|^F323|^F328|^F329|^F33|^F330|^F331|^F332|^F333|^F334|^F338|^F339", diagnosis)) 
  ) %>%
  mutate(mdd = 1,
       admit.date = ymd(admit.date)) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  add_count(id) %>%  # record the number of diagnoses per individual into variable "n"
  rename(nmddcontacts = n) %>% # rename n 
  mutate(first.mdd.date = first(admit.date)) %>%
  slice(1) %>%
  select(id, mdd, nmddcontacts, first.mdd.date)

scz.mdd <- inner_join(first.scz.date,mdd, by="id") %>% # inner join together with scz cases
  mutate(age.first.mdd = as.integer((first.mdd.date - dob)/365.25),
         mdd.childhood = ifelse(age.first.mdd <= 15, 1, 0), # only those with mdd dx before age 15 years
         mdd.pre.scz = ifelse(first.mdd.date < first_scz_date, 1, 0)  #  only those with mdd dx before scz contact
  ) %>%
  select(id, nmddcontacts, mdd.childhood, mdd.pre.scz)

```
7. Obsessive-Compulsive Disorder
```{r OCD}
# first extract all individuals with OCD diagnosis 
ocd <- hdr %>%
   filter((icd.version==8 & grepl("^3003", diagnosis))
      | (icd.version==9 & grepl("^300D", diagnosis))
      | (icd.version==10 & grepl("^F42|^F420|^F421|^F422|^F428|^F429", diagnosis))) %>%
  mutate(ocd = 1,
         admit.date = ymd(admit.date)) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  add_count(id) %>%  # record the number of diagnoses per individual into variable "n"
  rename(nmddcontacts = n) %>% # rename n 
  mutate(first.ocd.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,ocd,first.ocd.date)

ocd <- inner_join(first.scz.date,ocd) # inner join together with scz cases

ocd <- ocd %>%
  mutate(age.first.ocd = as.numeric((first.ocd.date - dob)/365.25),
         ocd.childhood = ifelse(age.first.ocd <= 15, 1, 0), # only those with ocd dx before age 15 years
         ocd.pre.scz = ifelse(first.ocd.date < first_scz_date, 1, 0)  #  only those with ocd dx before scz contact
  ) %>%
  select(id,ocd.childhood,ocd.pre.scz)


#######################
# PTSD ################
#######################
ptsd <- hdr %>% 
  filter(  (icd.version==8 & grepl("^307", diagnosis)) |
             (icd.version==9 & grepl("^308|^308A|^308B|^308C|^308D|^308E|^308X|^309|^309A|^309B|^309C|^309D|^309E|^309W|^309X", diagnosis)) |
             (icd.version==10 & grepl("^F43|^F430|^F431|^F432|^F438|^F439", diagnosis)) 
  ) %>%
  mutate(ptsd = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.ptsd.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,ptsd,first.ptsd.date)

ptsd <- inner_join(first.scz.date,ptsd) # inner join together with scz cases

ptsd <- ptsd %>%
  mutate(age.first.ptsd = as.numeric((first.ptsd.date - dob)/365.25),
         ptsd.childhood = ifelse(age.first.ptsd <= 15, 1, 0), # only those with ptsd dx before age 15 years
         ptsd.pre.scz = ifelse(first.ptsd.date < first_scz_date, 1, 0)  #  only those with ptsd dx before scz contact
  ) %>%
  select(id,ptsd.childhood,ptsd.pre.scz)


#######################
# Substance abuse #####
#######################
suds <- hdr %>%
  filter(  (icd.version==8 & grepl("^303|^3030|^3031|^3032|^3039|^304|^3040|^3041|^3042|^3043|^3044|^3045|^3046|^3047|^3048|^3049", diagnosis))
           | (icd.version==9 & grepl("^303|^303A|^303X|^304|^304A|^304B|^304C|^304D|^304D|^304E|^304F|^304G|^304H|^304W|^304X|^305A|^305X", diagnosis))
           | (icd.version==10 & grepl("^F10|^F100|^F101|^F102|^F103|^F104|^F105|^F106|^F107|^F108|^F109|^F11|^F110|^F111|^F112|^F113|^F114|^F115|^F116|^F117|^F118|^F119|
                               F19|^F190|^F191|^F192|^F192|^F194|^F195|^F196|^F197|^F198|^F199", diagnosis)) 
  ) %>%
  mutate(suds = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.suds.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,suds,first.suds.date)

suds <- inner_join(first.scz.date,suds) # inner join together with scz cases

suds <- suds %>%
  mutate(age.first.suds = as.numeric((first.suds.date - dob)/365.25),
         suds.childhood = ifelse(age.first.suds <= 15, 1, 0), # only those with suds dx before age 15 years
         suds.pre.scz = ifelse(first.suds.date < first_scz_date, 1, 0)  #  only those with suds dx before scz contact
  ) %>%
  select(id,suds.childhood,suds.pre.scz)


#######################
# Suicides ############
#######################
sui <- hdr %>%
   filter((icd.version==8 & grepl("^E950|^E951|^E952$|^E9520|^E9521|^E9529|^E953|^E954|^E955|^E956|^E957|^E958|^E959", diagnosis))
      | (icd.version==9 & grepl("^E95A|^E95B|^E95C|^E95D|^E95E|^E95F|^E95G|^E95H|^E95W|^E95X", diagnosis))
      | (icd.version==10)) %>%
  mutate(sui = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.sui.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,sui,first.sui.date)
sui <- inner_join(first.scz.date,sui) # inner join together with scz cases

sui <- sui %>%
  mutate(age.first.sui = as.numeric((first.sui.date - dob)/365.25),
         sui.childhood = ifelse(age.first.sui <= 15, 1, 0), # only those with sui dx before age 15 years
         sui.pre.scz = ifelse(first.sui.date < first_scz_date, 1, 0)  #  only those with sui dx before scz contact
  ) %>%
  select(id,sui.childhood,sui.pre.scz)


#########################################
###### MERGE ALL psych DISORDERS ########
#########################################

psych.comorbid <- full_join(bp,anx)
psych.comorbid <- full_join(psych.comorbid,ptsd)
psych.comorbid <- full_join(psych.comorbid,adhd)
psych.comorbid <- full_join(psych.comorbid,ed)
psych.comorbid <- full_join(psych.comorbid,sui)
psych.comorbid <- full_join(psych.comorbid,suds)
psych.comorbid <- full_join(psych.comorbid,ocd)
psych.comorbid <- full_join(psych.comorbid,asd)
psych.comorbid <- full_join(psych.comorbid,mdd)

rm(bp,anx,ptsd,adhd,ed,sui,suds,ocd,asd,mdd) # remove original dataframes
 
scz <- left_join(scz,psych.comorbid)
rm(psych.comorbid)

scz <- scz %>%
  mutate(bip.childhood=ifelse(is.na(bip.childhood),0,bip.childhood),
         bip.pre.scz=ifelse(is.na(bip.pre.scz),0,bip.pre.scz),
         anx.childhood=ifelse(is.na(anx.childhood),0,anx.childhood),
         anx.pre.scz=ifelse(is.na(anx.pre.scz),0,anx.pre.scz),
         ptsd.childhood=ifelse(is.na(ptsd.childhood),0,ptsd.childhood),
         ptsd.pre.scz=ifelse(is.na(ptsd.pre.scz),0,ptsd.pre.scz),
         adhd.childhood=ifelse(is.na(adhd.childhood),0,adhd.childhood),
         adhd.pre.scz=ifelse(is.na(adhd.pre.scz),0,adhd.pre.scz),
         ed.childhood=ifelse(is.na(ed.childhood),0,ed.childhood),
         ed.pre.scz=ifelse(is.na(ed.pre.scz),0,ed.pre.scz),
        sui.childhood=ifelse(is.na(sui.childhood),0,sui.childhood),
         sui.pre.scz=ifelse(is.na(sui.pre.scz),0,sui.pre.scz),
         suds.childhood=ifelse(is.na(suds.childhood),0,suds.childhood),
         suds.pre.scz=ifelse(is.na(suds.pre.scz),0,suds.pre.scz),
         ocd.childhood=ifelse(is.na(ocd.childhood),0,ocd.childhood),
         ocd.pre.scz=ifelse(is.na(ocd.pre.scz),0,ocd.pre.scz),
        asd.childhood=ifelse(is.na(asd.childhood),0,asd.childhood),
         asd.pre.scz=ifelse(is.na(asd.pre.scz),0,asd.pre.scz),
         mdd.childhood=ifelse(is.na(mdd.childhood),0,mdd.childhood),
         mdd.pre.scz=ifelse(is.na(mdd.pre.scz),0,mdd.pre.scz))
  
 
# Medication use  --- wait until decide on specific drugs

```
