---
title: "Tryggve2 - major depression - Sweden (part 3)"
author: "Lu Yi, lu.yi@ki.se"
date: "December 2018"
output: 
  html_document: 
    fig_caption: yes
    theme: simplex
    highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Motivation and Overview

NOTE: NOT RUN

Examine variables associated with poor outcomes in schizophrenia, using first, the Swedish National Registers. In this report, I will present some initial coding for the analyses.

**Overview**

1. Import data
2. Create a dataframe with all necessary variables including exposures and outcomes
3. Analyses

**Exposures or Predictors**
1) Family of origin:
+ parental ages at birth ##LY
+ parental SES-education-occupation-income
+ birth order
+ family composition (number of parents/partners, parental criminality-incarceration, divorce, siblings, out-of-home placement)

2) Childhood trauma: ##LY
+ parental death
+ severe medical illness of parents/sibs

3) Education: done-kk
+ educational attainment
+ standardized testing in grade 9

4) Prior personal medical history:
+ comorbid disorders and medication use ##kk
+ prior personal medical history (somatic disorders) #kk
+ childhood hospitalization

5) Family history of psychiatric disorder ##kk
+ MD, BIP, psychosis, suicide in first-degree relatives

6) Genetic risk factors  NOT NOW
+ GRS of MDD, SCZ, BIP, IQ etc.

**Outcomes**
+ the number of psychiatric hospitalizations and outpatient contacts 
+ the number of somatic healthcare utilization
+ indication of 2nd or 3rd line treatments (pharmacological augmentation, ECT, TMS, DBS, clozapine) 
+ suicide or suicide attempt

**TO DOs**

cut down this doc into 4 separate RMD (eventually):
outcome
exposure/risk factors
basics
analysis

***

### Step 0: Load R packages
```{r R packages}
library(tidyverse)
#library(dplyr)
library(data.table)
#library(ggplot2)
library(lubridate)
library(stringr)
library(skimr)

Sys.setenv(TZ="Europe/Stockholm")
```

### Step 1: Load data
REQUIRE EDITING on the file path
```{r specify file path}
# edit the file path where you have stored the part 1 data
data.f <- "/Users/luyi/CloudStation/WORK_2018/MyProjects_with_Pat/GAPS_MDD/DATA/FREED/DATA/mdd.part2.RData"

# load the R data
load(data.f) 

# load two more datasets here <- TO DO: later need to update in PART 1
# 1. Cause of Death register with deathcause
deathcause.f <- "/Volumes/projects/S3/S3_Research/tryggve2/tryggve_psych_v_deathcause.tsv"

deathcause <- fread(death.f, sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "dod"="X_DODSDAT",
         "icd.version"="ICD_NR",
         "death.cause"="CAUSE",
         "death.cause.nr"="CAUSENR")

# 2. suicide attempts identified through NPR 
sui.attempt.f <- "/Volumes/projects/S3/S3_Research/tryggve2/tryggve_psych_v_npr_sui.tsv"

sui.attempt <- fread(sui.attempt.f, sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "admit.date"="DATUM", # date of admission
         "inpatient"="SOURCE", # inpatient (1) or outpatient (2)
         "icd.version"="ICD", # ICD version - 8, 9 or 10
         "diagnosis"="DIA", # ICD diagnostic code
         "dx.code.num"="X_DID") %>% # primary vs. other ICD dx codes, i.e. H01 = primary, all others = secondary 
  mutate(inpatient = ifelse(inpatient == 2, 0, 1)) #recode such that inpatient == 1 and outpatient == 0

```

### Step 2: Outcome 1 & 2: Suicide - attempts or completed
Suicide attempt must have occurred after the first MDD date and only need first one (for now)
```{r suicide attempt}
# attempts from hospital discharge register
sui.a <- sui.attempt %>%
   filter((icd.version==8 & grepl("^E950|^E951|^E952$|^E9520|^E9521|^E9529|^E953|^E954|^E955|^E956|^E957|^E958|^E959", diagnosis))
      | (icd.version==9 & grepl("^E95A|^E95B|^E95C|^E95D|^E95E|^E95F|^E95G|^E95H|^E95W|^E95X", diagnosis))
      | (icd.version==10 & grepl("^X60|^X61|^X62|^X63|^X64|^X65|^X66|^X67|^X68|^X69|^X70|^X71|^X72|^X73|^X74|^X75|^X76|^X77|^X78|^X79|^X80|^X81|X82|^X83|^X84", diagnosis)) ) %>%
  mutate(sui = 1,
         admit.date = ymd(admit.date)) %>%
  arrange(id, admit.date)

sui.a <- inner_join(first.mdd.date,sui.a, by="id") # inner join together with mdd cases to get date of first.mdd.admit

sui.a <- sui.a %>%
  filter(admit.date > first_mdd_date) %>% # remove sui attempts before mdd contact   
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(date.sui.attempt = first(admit.date),
         sui.attempt = sui, sui=NULL) %>%
  slice(1) %>%
  select(id, date.sui.attempt, sui.attempt)
```

```{r completed suicide}
# completed suicide from cause of death register
sui.c <- deathcause %>%
   filter((icd.version==8 & grepl("^E950|^E951|^E952$|^E9520|^E9521|^E9529|^E953|^E954|^E955|^E956|^E957|^E958|^E959", death.cause))
      | (icd.version==9 & grepl("^E95A|^E95B|^E95C|^E95D|^E95E|^E95F|^E95G|^E95H|^E95W|^E95X", death.cause))
      | (icd.version==10 & grepl("^X60|^X61|^X62|^X63|^X64|^X65|^X66|^X67|^X68|^X69|^X70|^X71|^X72|^X73|^X74|^X75|^X76|^X77|^X78|^X79|^X80|^X81|X82|^X83|^X84", death.cause)) ) %>%
  mutate(sui.death = 1,
         death.date = ymd(dod)) %>%
  arrange(id, death.date)

sui.c <- inner_join(first.mdd.date,sui.c, by="id") # inner join together with mdd cases to get date of first.mdd.admit

sui.c <- sui.c %>%
  filter(death.date > first_mdd_date) %>% # remove sui death before mdd contact **there are a few samples, mdd_date same as death date
  arrange(id, death.date) %>%
  group_by(id) %>%
  mutate(date.sui.death = first(death.date)) %>%
  slice(1) %>%
  select(id, date.sui.death, sui.death)

```

```{r merge suicide data with MDD ind}

# merge suicide attempt and completed suicide with original mdd dataframe

mdd.perind <- left_join(mdd.perind, sui.a, by="id")
mdd.perind <- left_join(mdd.perind, sui.c, by="id")

# also wanted derive a variable indicating whether had any suicide attempt or completed suicide

rm(sui.a, sui.c)

```

### Step 3: Outcome 3: Premature mortality TO DO 
+ Mortality which occurs by age 56 years or earlier (any causes) (premature death=56 years as defined by Fazel, Lichtenstein, et al, Lancet Psychiatry 2014)
+ Could also take life expectancy in Sweden for both males and females in 2016 (82.2 years), less 25 years (expected number of years less for someone w SCZ) = 82-25 = 57 years

?? what is the shortened life expectancy for MDD

...

### Check missing data in all outcomes
```{r missingness}
# check missing data on each variable
mdd.perind %>% purrr::map_df(~ sum(is.na(.))) %>% gather(variable, N_missing)

```

### SAVE ALL THE DATA 
```{r save data}
save.image(file = "/Users/luyi/CloudStation/WORK_2018/MyProjects_with_Pat/GAPS_MDD/DATA/FREED/DATA/mdd.part2.RData",compress = TRUE)
```
