---
title: "Tryggve2 - Schizophrenia - Sweden - "
author: "Kaarina Kowalec, kaarina.kowalec@ki.se"
date: "November 2018"
output: 
  html_document: 
    fig_caption: yes
    theme: simplex
    highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Motivation and Overview

Examine variables associated with poor outcomes in schizophrenia, using first, the Swedish National Registers. In this report, I will present some initial coding for the analyses.

**Overview**

1. Import data to create list of SCZ cases and other exposures/outcomes
2. Generate list of SCZ cases then export list of id's to pull out full hospital discharge register for only those cases (limit file sizes)
3. Generate exposures
4. Generate outcomes
5. Analyses

**Exposures or Predictors**
1) Family of origin:
+ parental ages at birth ##LY
+ parental SES-education-occupation-income
+ birth order
+ family composition (number of parents/partners, parental criminality-incarceration, divorce, siblings, out-of-home placement)

2) Childhood trauma: ##LY
+ parental death
+ severe medical illness of parents/sibs

3) Education: done-kk
+ educational attainment
+ standardized testing in grade 9

4) Prior personal medical history (kk done, need childhood hospitalisation):
+ comorbid disorders and medication use
+ prior personal medical history (somatic disorders)
+ childhood hospitalization

5) Family history of psychiatric disorder ##kk
+ MD, BIP, psychosis, suicide in first-degree relatives

6) Genetic risk factors  NOT NOW
+ GRS of MDD, SCZ, BIP, IQ etc.

**Outcomes**
+ the number of psychiatric hospitalizations and outpatient contacts 
+ the number of somatic healthcare utilization
+ indication of 2nd or 3rd line treatments (pharmacological augmentation, ECT, TMS, DBS, clozapine) 
+ suicide or suicide attempt

**TO DOs**

cut down this doc into 4 separate RMD (eventually):
outcome
exposure/risk factors
basics
analysis

***

### Step 1: Load data & packages

1. Hospital discharge data (hdr): all in-/out-patient hospitalisation codes, multiple lines per person, contains date of admission/discharge
2. Population register (pop): birth date, sex, birth country and Swedish birth county 
3. Death register (death): date of death, causes of death
4. Migration register (migrate): date of emigration/immigration
5. Multi-generation register (family): personnummer of index case and personnummer of child/parents/sibs
6. Medical birth register (birth): containing birth outcomes
7. Drug register (drug): All prescriptions redeemed between 7/2005-12/2014, with ATC codes/drug names)
8. Education (education): from LISA register; Svensk utbildningsnomenklatur (SUN) or Swedish education nomenclature
9. Grade 9 grades (grades): year 9 grades from the School register
10. Civil register (civil): civil status (may or may not use)

```{r warning=FALSE, message=FALSE}
library(dplyr)
library(data.table)
library(ggplot2)
library(lubridate)
library(stringr)

Sys.setenv(TZ="Europe/Stockholm")

# connect to meb server

hdr <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/tryggve_psych_v_npr.tsv", sep="\t") %>%
  as_tibble() %>%
  rename(id=LOPNR,
         admit.date=DATUM, # date of admission
         inpatient=SOURCE, # inpatient (1) or outpatient (2)
         icd.version=ICD, # icd version - 8, 9 or 10
         diagnosis=DIA,# ICD diagnostic code
         #id.admit.num=X_PID, # per individual ID per admission
         dx.code.num=X_DID) %>% # primary vs. other ICD dx codes, i.e. H01 = primary, all others = secondary
         #care.code=MVO) %>%  # MVO of 9 = psych care, all others will not be used
  mutate(inpatient = ifelse(inpatient == 2, 0, 1)) #recode such that inpatient == 1 and outpatient == 0

pop <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/tryggve_psych_v_indiv.tsv", sep="\t") %>%
  as_tibble() %>%
  rename(id=LOPNR,
         sex=KON, # 1 = male, 2 = female
         birth.country=FODELSEGRUPP, # country of birth 
         swe.birth.county=FODELSELAN,# swedish county of birth
         dob=FODELSEDATUM) %>% # date of birth (mmyyyy)
  mutate(swe.birth.county = ifelse(swe.birth.county == 2, 1, swe.birth.county), # account for recategorisation as per Amir Sariaslan
    swe.birth.county = ifelse(swe.birth.county == 11, 12, swe.birth.county),
    swe.birth.county = ifelse(swe.birth.county %in% c(15, 16), 14, swe.birth.county))

pop$birth.country <- as.factor(pop$birth.country)
levels(pop$birth.country) <-
  c(
    "Sweden",
    "Nordic not Sweden",
    "EU28 not Nordic",
    "Europe not EU28 not Nordic",
    "Africa",
    "North America",
    "South America",
    "Asia",
    "Oceania",
    "Soviet union"
  )

# urban vs rural birth?
levels(pop$swe.birth.county) <-
  c("Stockholm County",
    "Uppsala",
    "Sodermanland",
    "Vastergotland",
    "Jonkoping",
    "Kronoberg",
    "Kalmar",
    "Gotland",
    "Blekinge",
    "Skane",
    "Halland",
    "Vastra Gotaland",
    "Varmland",
    "Orebro",
    "Vastmanland",
    "Dalarna",
    "Gavleborg",
    "Vasternorrland",
    "Jamtland",
    "Vasterbotten",
    "Norrbotten"
  )

death <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/tryggve_psych_v_death.tsv", sep="\t") %>%
  as_tibble() %>%
  rename(id=LOPNR,
         dod=X_DODSDATUM)

migrate <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/tryggve_psych_v_migrate.tsv", sep="\t") %>%
  as_tibble() %>%
  rename(id=LOPNR,
         migrate.date=MDATUM,
         migrate.type=MTYP) %>% # (E=emigration, I=immigration)
  filter(migrate.type == "E") # only want emigrations

family <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/tryggve_psych_v_parent.tsv", sep="\t") %>%
  as_tibble() %>%
  rename(id=LOPNR,
         mom.id=LOPNRMOR,
         dad.id=LOPNRFAR)

child <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/tryggve_psych_v_child.tsv", sep="\t") %>%
  as_tibble() %>%
  rename(id=LOPNR,
         kid.id=LOPNRBARN) 

birth <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/tryggve_psych_v_mbr.tsv", sep="\t") %>%
  as_tibble() %>%
  rename(mom.id=LOPNRMOR,
         kid.id=LOPNRBARN,
         mom.age.at.birth=MALDER)

education <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/tryggve_psych_v_education.tsv", sep="\t") %>%
  as_tibble() %>%
  rename(id=LOPNR,
         edlevel=SUN2000NIVA) 

grades <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/tryggve_psych_v_grades.tsv", sep="\t") %>%
  as_tibble() %>%
  select(LOPNR, AVGAR, MEDELBETYG, MERITVARDE) %>% # select only useful variables
  rename(id=LOPNR,
         grad.yr=AVGAR, # graduation year
         avg.grade.fr88=MEDELBETYG, # MEDELBETYG	Average grade (1988-1997, min-max=0,0-5,0)
         merit.grade.fr98=MERITVARDE) #  MERITVARDE	Merit rating (1998-, best 16 subjects, min-max=0-320)

### LEAVE FOR NOW ####
civil <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/tryggve_psych_v_civil.tsv",sep="\t") %>%
  as_tibble() %>%
  rename(id=LOPNR,
         year=AR,
         civil.status=CIVIL) %>%
  mutate(civil.status == case_when(civil.status == 1 ~ "Unmarried",
                                   civil.status == 2 ~ "Married man",
                                   civil.status == 3 ~ "Married woman, not cohabiting",
                                   civil.status == 4 | civil.status == "RP" ~ "Divorced",
                                   civil.status == 5 ~ "Widow/Widower",
                                   civil.status == 7 ~ "Married woman, cohabiting",
                                   civil.status == 8 ~ "Child, <18 years",
                                   civil.status == 9 ~ "Fosterchild, <18 years",
                                   civil.status == "EP" ~ "Surviving partner",
                                   civil.status == "G" ~ "Married",
                                   civil.status == "OG" ~ "Registered partner",
                                   civil.status == "SP" ~ "Divorced partner",
                                   civil.status == "─" ~ "Widow/Widower",
                                   TRUE ~ NA))


######################################
######### ONLY AVAILABLE IN SWEDEN ###
######################################

# drug register not need by other countries 
drug <- fread("~/drug.tsv", sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "num.pkg"="ANTAL",
         "atc"="ATC",
         "ddd"="DDDFORP",
         "dose"="DOSER", #free text
         "disp.date"="EDATUM", # (yyyy-mm-dd)
         "pkg.size"="FORPSTL",
         "admin.route"="LFORMGRUPP", #  +formula of medication, grouped: 7=cutaneous/transdermal, 14=rectal, 12=unspecified, 9=oral/fast
         "drug.prod"="PRODUKTNAMN", # product name
         "strength_alf"="STYRKA_ALF",
         "strength_enh"="STYRKA_ENH",
         "strength_num"="STYRKNUMERIC")

```
### Step 2: Create data frame with SCZ cases, date of first and second SCZ admission
```{r message=FALSE}

scz <- hdr %>%
  filter((icd.version != 10 & substr(diagnosis,1,3) == "295") | # use ?grep and refine codes over time for SCZ but start broad for now
          (icd.version == 10 & substr(diagnosis,1,3) == "F20") |
           (icd.version == 10 & substr(diagnosis,1,3) == "F25")) %>%
  mutate(scz = 1, # add variable to indicate those with SCZ
         scz_date = ymd(as.character(admit.date))) %>% # date of admission with SCZ as discharge dx
  add_count(id) %>%  # add number of admissions with SCZ discharge dx
  mutate(nsczcontacts = n, n=NULL) # n=null removes the original n variable

scz <- scz %>%
  arrange(id, scz_date) %>% # order by ID and then scz_date
  group_by(id) %>% # per person
  mutate(first_scz_date = first(scz_date), # get date of first and second contact points for scz
         second_scz_date = nth(scz_date, 2)) %>%
  ungroup()

## tally up the number of scz hospitalisations (inpatiet and outpatient)

sczinpt <- scz %>%
  filter(inpatient == 1) %>% # filter for inpatient record
  add_count(id) %>% # add number of inpatient records per person
  rename(nsczinpt = n) %>%  # rename variable
  arrange(id) %>% # order by id
  group_by(id) %>% 
  slice(1) %>% # take the first row
  select(id,nsczinpt) 

sczoutpt <- scz %>%
  filter(inpatient == 0) %>% # filter for outpatient record
  add_count(id) %>% # add number of outpatient records pp
  rename(nsczoutpt = n) %>% # rename variable
  arrange(id) %>% # order by id
  group_by(id) %>%
  slice(1) %>% # take first row
  select(id,nsczoutpt)

scz <- scz %>%
  arrange(id,scz_date) %>% # collapse to per-individual record and not per-admission
  group_by(id) %>%
  slice(1) %>%
  select(id, scz, first_scz_date, second_scz_date, nsczcontacts) %>%
  ungroup()

scz <- left_join(scz, sczinpt) # join together w inpatient
scz <- left_join(scz, sczoutpt, by = "id") # join together w outpatient

rm(sczinpt,sczoutpt) # remove un-needed dataframes 

# fix variables with NA
# some of the n_inpatient / n_outpatient = NA because the individual has all inpatient records and no outpatients records or vice-versa for scz

scz <- scz %>%
  mutate(nsczinpt = ifelse(is.na(nsczinpt), 0, nsczinpt),
         nsczoutpt = ifelse(is.na(nsczoutpt), 0, nsczoutpt))


# number of SCZ cases = 79,044
# number of SCZ with >= 2 SCZ admissions = 60,625 
# number of SCZ with >= 2 SCZ admissions with a date past January 1, 1973 = 60,037

```



### Step 3: Demographics: sex, death, emigration
```{r}
## Merge in population data

scz <- left_join(scz,pop)

###############################
# DOB #########################
###############################

# add 15 to date of birth
scz$dob <- as.character(scz$dob)
scz$dob <- sapply(scz$dob, paste0, "15")

scz <- scz %>%
  mutate(dob = ymd(dob))

rm(pop) # remove pop register

###############################
# Death #######################
###############################

death <- death %>%
  mutate(dod = ymd(as.character(dod)))

scz <- left_join(scz, death) # join in data with main population
rm(death) # remove death reg

###############################
# Date of Emigration ##########
###############################

# select only those born in respective countries? for discussion later on
# add "15" to date
migrate$migrate.date <- as.character(migrate$migrate.date)
migrate$migrate.date <- sapply(migrate$migrate.date, paste0, "15")

migrate <- migrate %>% 
  mutate(migrate.date = ymd(as.character(migrate.date))) %>% # make date of emigration into ymd format
  arrange(id, migrate.date) %>% # order by id then migration date
  group_by(id) %>% # group by person
  mutate(first.emigrate.date = first(migrate.date)) %>% # add first emigration date (i.e. date they left the country)
  ungroup()

migrate <- migrate %>% 
  arrange(id) %>% # order by ID
  group_by(id) %>%
  slice(1) %>% # select first observation only per individual
  select(id, first.emigrate.date)

scz <- left_join(scz, migrate)

rm(migrate) # remove migration register

###############################
# Age at 1st, 2nd SCZ contact #
###############################

scz <- scz %>%
  mutate(age.first.scz = as.numeric((first_scz_date - dob)/365.25))


```



### Step 4: Exposure-Prior psychiatric history
Need to start with making a list of any one with a psych disorder in Sweden because will need
for family history of psychiatric disorders.

We are only selecting those with a date (1) before first treatment contact and (2) before age 15 years (indicating childhood)

+ comorbid psych disorders/conditions: psych (adhd, asd, anx, bip, ed, mdd, ocd, ptsd, suds, suicides)
+ medication use (wait on this-not sure which drugs to focus on)
```{r}
# make list of scz cases and date of first contact for comparison with psych disorders below
first.scz.date <- scz %>%
  select(id, first_scz_date, dob)

#######################  
# ADHD ################
#######################
adhd <- hdr %>% 
  filter(  (icd.version==9 & grepl("^314|^314A|^314B|^314C|^314J|^314W|^314X", diagnosis)) # no icd codes for ADHD prior to ICD version 9
           | (icd.version==10 & grepl("^F90|^F900|^F901|^F908", diagnosis)) 
  ) %>%
  mutate(adhd = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.adhd.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,adhd,first.adhd.date)

adhd <- inner_join(first.scz.date,adhd) # inner join together with scz cases

adhd <- adhd %>%
  mutate(age.first.adhd = as.numeric((first.adhd.date - dob)/365.25),
         adhd.childhood = ifelse(age.first.adhd <= 15, 1, 0), # only those with adhd dx before age 15 years
         adhd.pre.scz = ifelse(first.adhd.date < first_scz_date, 1, 0)  #  only those with adhd dx before scz contact
) %>%
  select(id,adhd.childhood,adhd.pre.scz)

#######################
# ASD #################
#######################
asd <- hdr %>% 
  filter(  (icd.version==9 & grepl("^299A", diagnosis)) # no icd codes for asd prior to version 9
           | (icd.version==10 & grepl("^F840|^F841|^F845", diagnosis)) 
  ) %>%
  mutate(asd = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.asd.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,asd,first.asd.date)

asd <- inner_join(first.scz.date,asd) # inner join together with scz cases

asd <- asd %>%
  mutate(age.first.asd = as.numeric((first.asd.date - dob)/365.25),
         asd.childhood = ifelse(age.first.asd <= 15, 1, 0), # only those with asd dx before age 15 years
         asd.pre.scz = ifelse(first.asd.date < first_scz_date, 1, 0)  #  only those with asd dx before scz contact
) %>%
  select(id,asd.childhood,asd.pre.scz)

#######################
# Anxiety #############
#######################
anx <- hdr %>% 
  filter(  (icd.version==8 & grepl("^300|^3000|^3002", diagnosis)) |
             (icd.version==9 & grepl("^300|^300A|^300C", diagnosis)) |
             (icd.version==10 & grepl("^F40|^F400|^F401|^F402|^F408|^F409|^F41|^F410|^F411|^F412|^F413|^F418|^F419", diagnosis)) 
  ) %>%
  mutate(anx = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.anx.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,anx,first.anx.date)

anx <- inner_join(first.scz.date,anx) # inner join together with scz cases

anx <- anx %>%
  mutate(age.first.anx = as.numeric((first.anx.date - dob)/365.25),
         anx.childhood = ifelse(age.first.anx <= 15, 1, 0), # only those with anx dx before age 15 years
         anx.pre.scz = ifelse(first.anx.date < first_scz_date, 1, 0)  #  only those with anx dx before scz contact
  ) %>%
  select(id,anx.childhood,anx.pre.scz)


#######################
# BIP #################
#######################
bp <- hdr %>%
  filter( ((icd.version==8) & grepl('^2961|^2963|^2968',diagnosis)) |
          ((icd.version==9) & grepl('^296|^269A|^269C|^2691|^296D|^2694|^296E|^2965|^296F|^2966|^296G|^2697|^269H|^2691|^26989|^269W|^26999|^269X',diagnosis)) | 
            ((icd.version==10) & grepl('^F30$|^F301|^F302|^F308|^F309|^F31|^F310|^F311|^F312|^F313|^F314|^F315|^F316|^F317|^F318|^F319',diagnosis)) ) %>%
  mutate(bip = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.bip.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,bip,first.bip.date)

bp <- inner_join(first.scz.date,bp) # inner join together with scz cases

bp <- bp %>%
  mutate(age.first.bip = as.numeric((first.bip.date - dob)/365.25),
         bip.childhood = ifelse(age.first.bip <= 15, 1, 0), # only those with bip dx before age 15 years
         bip.pre.scz = ifelse(first.bip.date < first_scz_date, 1, 0)  #  only those with bip dx before scz contact
  ) %>%
  select(id,bip.childhood,bip.pre.scz)



#######################
# Eating Disorders ####
#######################
ed <- hdr %>%
  filter(  (icd.version==8 & grepl("^7840|^3065", diagnosis))
           | (icd.version==9 & grepl("^307B|^307F", diagnosis))
           | (icd.version==10 & grepl("^F50|^F500|^F501|^F502|^F503|^F509", diagnosis)) 
  ) %>%
  mutate(ed = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.ed.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,ed,first.ed.date)

ed <- inner_join(first.scz.date,ed) # inner join together with scz cases

ed <- ed %>%
  mutate(age.first.ed = as.numeric((first.ed.date - dob)/365.25),
         ed.childhood = ifelse(age.first.ed <= 15, 1, 0), # only those with ed dx before age 15 years
         ed.pre.scz = ifelse(first.ed.date < first_scz_date, 1, 0)  #  only those with ed dx before scz contact
  ) %>%
  select(id,ed.childhood,ed.pre.scz)


#######################
# MDD #################
#######################

mdd <- hdr %>% 
  filter(  (icd.version==8 & grepl("^2960|^2968|^2969|^2980|^3004", diagnosis)) |
            (icd.version==9 & grepl("^296C|^296D|^311", diagnosis)) |
             (icd.version==10 & grepl("^F32|^F320|^F321|^F322|^F323|^F328|^F329|^F33|^F330|^F331|^F332|^F333|^F334|^F338|^F339", diagnosis)) 
  ) %>%
  mutate(mdd = 1,
       admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.mdd.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,mdd,first.mdd.date)

mdd <- inner_join(first.scz.date,mdd) # inner join together with scz cases

mdd <- mdd %>%
  mutate(age.first.mdd = as.numeric((first.mdd.date - dob)/365.25),
         mdd.childhood = ifelse(age.first.mdd <= 15, 1, 0), # only those with mdd dx before age 15 years
         mdd.pre.scz = ifelse(first.mdd.date < first_scz_date, 1, 0)  #  only those with mdd dx before scz contact
  ) %>%
  select(id,mdd.childhood,mdd.pre.scz)


#######################
# OCD #################
#######################
ocd <- hdr %>%
   filter((icd.version==8 & grepl("^3003", diagnosis))
      | (icd.version==9 & grepl("^300D", diagnosis))
      | (icd.version==10 & grepl("^F42|^F420|^F421|^F422|^F428|^F429", diagnosis))) %>%
  mutate(ocd = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.ocd.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,ocd,first.ocd.date)

ocd <- inner_join(first.scz.date,ocd) # inner join together with scz cases

ocd <- ocd %>%
  mutate(age.first.ocd = as.numeric((first.ocd.date - dob)/365.25),
         ocd.childhood = ifelse(age.first.ocd <= 15, 1, 0), # only those with ocd dx before age 15 years
         ocd.pre.scz = ifelse(first.ocd.date < first_scz_date, 1, 0)  #  only those with ocd dx before scz contact
  ) %>%
  select(id,ocd.childhood,ocd.pre.scz)


#######################
# PTSD ################
#######################
ptsd <- hdr %>% 
  filter(  (icd.version==8 & grepl("^307", diagnosis)) |
             (icd.version==9 & grepl("^308|^308A|^308B|^308C|^308D|^308E|^308X|^309|^309A|^309B|^309C|^309D|^309E|^309W|^309X", diagnosis)) |
             (icd.version==10 & grepl("^F43|^F430|^F431|^F432|^F438|^F439", diagnosis)) 
  ) %>%
  mutate(ptsd = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.ptsd.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,ptsd,first.ptsd.date)

ptsd <- inner_join(first.scz.date,ptsd) # inner join together with scz cases

ptsd <- ptsd %>%
  mutate(age.first.ptsd = as.numeric((first.ptsd.date - dob)/365.25),
         ptsd.childhood = ifelse(age.first.ptsd <= 15, 1, 0), # only those with ptsd dx before age 15 years
         ptsd.pre.scz = ifelse(first.ptsd.date < first_scz_date, 1, 0)  #  only those with ptsd dx before scz contact
  ) %>%
  select(id,ptsd.childhood,ptsd.pre.scz)


#######################
# Substance abuse #####
#######################
suds <- hdr %>%
  filter(  (icd.version==8 & grepl("^303|^3030|^3031|^3032|^3039|^304|^3040|^3041|^3042|^3043|^3044|^3045|^3046|^3047|^3048|^3049", diagnosis))
           | (icd.version==9 & grepl("^303|^303A|^303X|^304|^304A|^304B|^304C|^304D|^304D|^304E|^304F|^304G|^304H|^304W|^304X|^305A|^305X", diagnosis))
           | (icd.version==10 & grepl("^F10|^F100|^F101|^F102|^F103|^F104|^F105|^F106|^F107|^F108|^F109|^F11|^F110|^F111|^F112|^F113|^F114|^F115|^F116|^F117|^F118|^F119|
                               F19|^F190|^F191|^F192|^F192|^F194|^F195|^F196|^F197|^F198|^F199", diagnosis)) 
  ) %>%
  mutate(suds = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.suds.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,suds,first.suds.date)

suds <- inner_join(first.scz.date,suds) # inner join together with scz cases

suds <- suds %>%
  mutate(age.first.suds = as.numeric((first.suds.date - dob)/365.25),
         suds.childhood = ifelse(age.first.suds <= 15, 1, 0), # only those with suds dx before age 15 years
         suds.pre.scz = ifelse(first.suds.date < first_scz_date, 1, 0)  #  only those with suds dx before scz contact
  ) %>%
  select(id,suds.childhood,suds.pre.scz)


#######################
# Suicides ############
#######################
sui <- hdr %>%
   filter((icd.version==8 & grepl("^E950|^E951|^E952$|^E9520|^E9521|^E9529|^E953|^E954|^E955|^E956|^E957|^E958|^E959", diagnosis))
      | (icd.version==9 & grepl("^E95A|^E95B|^E95C|^E95D|^E95E|^E95F|^E95G|^E95H|^E95W|^E95X", diagnosis))
      | (icd.version==10)) %>%
  mutate(sui = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.sui.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,sui,first.sui.date)
sui <- inner_join(first.scz.date,sui) # inner join together with scz cases

sui <- sui %>%
  mutate(age.first.sui = as.numeric((first.sui.date - dob)/365.25),
         sui.childhood = ifelse(age.first.sui <= 15, 1, 0), # only those with sui dx before age 15 years
         sui.pre.scz = ifelse(first.sui.date < first_scz_date, 1, 0)  #  only those with sui dx before scz contact
  ) %>%
  select(id,sui.childhood,sui.pre.scz)


#########################################
###### MERGE ALL psych DISORDERS ########
#########################################

psych.comorbid <- full_join(bp,anx)
psych.comorbid <- full_join(psych.comorbid,ptsd)
psych.comorbid <- full_join(psych.comorbid,adhd)
psych.comorbid <- full_join(psych.comorbid,ed)
psych.comorbid <- full_join(psych.comorbid,sui)
psych.comorbid <- full_join(psych.comorbid,suds)
psych.comorbid <- full_join(psych.comorbid,ocd)
psych.comorbid <- full_join(psych.comorbid,asd)
psych.comorbid <- full_join(psych.comorbid,mdd)

rm(bp,anx,ptsd,adhd,ed,sui,suds,ocd,asd,mdd) # remove original dataframes
 
scz <- left_join(scz,psych.comorbid)
rm(psych.comorbid)

scz <- scz %>%
  mutate(bip.childhood=ifelse(is.na(bip.childhood),0,bip.childhood),
         bip.pre.scz=ifelse(is.na(bip.pre.scz),0,bip.pre.scz),
         anx.childhood=ifelse(is.na(anx.childhood),0,anx.childhood),
         anx.pre.scz=ifelse(is.na(anx.pre.scz),0,anx.pre.scz),
         ptsd.childhood=ifelse(is.na(ptsd.childhood),0,ptsd.childhood),
         ptsd.pre.scz=ifelse(is.na(ptsd.pre.scz),0,ptsd.pre.scz),
         adhd.childhood=ifelse(is.na(adhd.childhood),0,adhd.childhood),
         adhd.pre.scz=ifelse(is.na(adhd.pre.scz),0,adhd.pre.scz),
         ed.childhood=ifelse(is.na(ed.childhood),0,ed.childhood),
         ed.pre.scz=ifelse(is.na(ed.pre.scz),0,ed.pre.scz),
        sui.childhood=ifelse(is.na(sui.childhood),0,sui.childhood),
         sui.pre.scz=ifelse(is.na(sui.pre.scz),0,sui.pre.scz),
         suds.childhood=ifelse(is.na(suds.childhood),0,suds.childhood),
         suds.pre.scz=ifelse(is.na(suds.pre.scz),0,suds.pre.scz),
         ocd.childhood=ifelse(is.na(ocd.childhood),0,ocd.childhood),
         ocd.pre.scz=ifelse(is.na(ocd.pre.scz),0,ocd.pre.scz),
        asd.childhood=ifelse(is.na(asd.childhood),0,asd.childhood),
         asd.pre.scz=ifelse(is.na(asd.pre.scz),0,asd.pre.scz),
         mdd.childhood=ifelse(is.na(mdd.childhood),0,mdd.childhood),
         mdd.pre.scz=ifelse(is.na(mdd.pre.scz),0,mdd.pre.scz))
  
 
# Medication use  --- wait until decide on specific drugs

```


### Step 5: Exposure-Family history of psychiatric disorder
To get at family history of psychiatric disorders, we need a list of all individuals with any of these disorders first (n=10 disorders):
  + ADHD
  + ASD
  + Anxiety
  + Bipolar
  + ED
  + MDD
  + OCD
  + PTSD
  + SUDS
  + Suicide attempts
```{r}
first.scz.date <- first.scz.date %>%
  mutate(scz=1)

# Create 4 new lists : 1) mother, 2) father, 3) child, 4) sibs
# create list of 'family' and 'child' with index ID being those w SCZ
scz.family <- inner_join(first.scz.date, family, by = "id")
kid <- inner_join(first.scz.date, child, by = "id")
 
# inner join w scz file with each of the 4 lists and see if any of these have SCZ/SAD; make variable say scz: yes or no but it must be dx before index case SCZ!

### MOM
mother <- scz.family %>%
  select(mom.id) %>%
  distinct(mom.id)

mother.scz <- inner_join(mother, first.scz.date, by = c("mom.id" = "id"))
mother.scz <- mother.scz %>%
  select(-dob) %>%
  rename(mom.scz = scz,
        first_scz_date.mom = first_scz_date)

mother.scz <- left_join(mother.scz, family, by = "mom.id") # get link between mother child
mother.scz <- left_join(mother.scz, first.scz.date, by = "id") # get info from child scz

mother.scz <- mother.scz %>%
  mutate(mom.scz = ifelse(first_scz_date.mom < first_scz_date, 1, 0)) %>% #  only those with scz dx in mom before scz dx in index case
  arrange(mom.id,desc(mom.scz)) %>%
  group_by(mom.id) %>%
  slice(1) %>%
  select(mom.scz,mom.id)
  
### DAD
father <- scz.family %>%
  select(dad.id) %>%
  distinct(dad.id)

father.scz <- inner_join(father, first.scz.date, by = c("dad.id" = "id"))

father.scz <- father.scz %>%
  select(-dob) %>%
  rename(dad.scz = scz,
        first_scz_date.dad= first_scz_date)

father.scz <- left_join(father.scz, family, by = "dad.id") # get link between father child
father.scz <- left_join(father.scz, first.scz.date, by = "id") # get info from child scz

father.scz <- father.scz %>%
  mutate(dad.scz = ifelse(first_scz_date.dad < first_scz_date, 1, 0)) %>% #  only those with scz dx in dad before scz dx in index case
  arrange(dad.id,desc(dad.scz)) %>%
  group_by(dad.id) %>%
  slice(1) %>%
  select(dad.scz,dad.id)

## KIDS
kids.scz <- kid %>%
  select(kid.id) %>%
  distinct(kid.id)

kids.scz <- inner_join(kids.scz, first.scz.date, by = c("kid.id" = "id"))

kids.scz <- kids.scz %>%
  select(-dob) %>%
  rename(kid.scz = scz,
        first_scz_date.kid = first_scz_date)

kids.scz <- left_join(kids.scz, child, by = "kid.id") # get link between kid and index
kids.scz <- left_join(kids.scz, first.scz.date, by = "id") # get info from child scz

kids.scz <- kids.scz %>%
  mutate(kid.scz = ifelse(first_scz_date.kid < first_scz_date, 1, 0)) %>% #  only those with scz dx in kid before scz dx in index case
  arrange(kid.id,desc(kid.scz)) %>%
  group_by(kid.id) %>%
  slice(1) %>%
  select(kid.scz,kid.id)

## SIBS
sibs.scz <- scz.family %>% # take only relationships with mom and dad LOPNRs included
  filter(!is.na(dad.id) & !is.na(mom.id)) %>%
  add_count(dad.id, mom.id) %>%
  filter(n > 1) %>%
  rename(sib.aff.scz = n)

sibs.scz <- sibs.scz %>%
  group_by(dad.id, mom.id) %>%
  arrange(first_scz_date) %>%
  mutate(sib1.scz.date = first(first_scz_date)) %>%
  ungroup() %>%
  mutate(sibs.scz = ifelse(sib1.scz.date < first_scz_date, 1, 0)) %>% #  only those with scz dx in sibs before scz dx in index case)
  select(id, sibs.scz)

# join back with kids and parents file 
scz.family <- left_join(scz.family, mother.scz, by = "mom.id")
scz.family <- left_join(scz.family, father.scz, by = "dad.id")
kid <- left_join(kid, kids.scz, by = "kid.id")

kid <- kid %>%
  mutate(fecund = 1) %>% # might be interested at some point as to whether or not someone had a kid - so add another variable
  select(id, kid.scz, fecund) %>%
  arrange(id, desc(kid.scz)) %>% # remove multiple entries in kids (i.e. >1 kid)
  group_by(id) %>%
  slice(1)

scz.family$mom.scz <- as.integer(scz.family$mom.scz)
scz.family$dad.scz <- as.integer(scz.family$dad.scz)

# add variable that indicates family history yes or no
scz.family <- scz.family %>%
  select(id,mom.scz,dad.scz) %>%
  mutate(fhx = case_when(mom.scz==1 | dad.scz==1 ~ 1,
                         mom.scz==0 | dad.scz==0 ~ 0,
                         TRUE ~ NA_real_)) %>%
  select(-mom.scz, -dad.scz)

kid <- kid %>%
  rename(fhx=kid.scz)

# merge parents and kids file
fhx <- full_join(scz.family, kid)
fhx <- full_join(fhx, sibs.scz)

fhx <- fhx %>%
  mutate(fhx.scz = case_when((fhx == 1 | sibs.scz == 1) ~ 1, 
                                 fhx == 0 | sibs.scz == 0 ~ 0,
                                 TRUE ~ NA_real_))

fhx <- fhx %>%
  select(id,fhx.scz,fecund)

# filter duplicates
fhx <- fhx %>%
  arrange(id,desc(fhx.scz)) %>%
  group_by(id) %>%
  slice(1)

# join scz with fhx

scz <- left_join(scz, fhx, by = "id")

# remove unneeded dataframes
rm(father,father.scz,mother,mother.scz,fhx,scz.family,sibs.scz,kid,kids.scz)




#########################################################################
############ Family history of psychiatric disorder  ####################
#########################################################################

#################################
#### CREATE FAMILY MATRIX #######
#################################

# create list of 'family' and 'child' with index ID being those w SCZ
scz.family <- inner_join(first.scz.date, family, by = "id") # list of index ids with SCZ, and their mom + dad
scz.kid <- inner_join(first.scz.date, child, by = "id") # list of index ids with SCZ and their offspring

# Create 4 new lists : 1) mother, 2) father, 3) child, 4) sibs
# then inner join w each psych disorder, make variable say PSYCH.DX: yes or no but it must be dx before index case SCZ!

mother <- scz.family %>%
  arrange(mom.id) %>%
  group_by(mom.id) %>% 
  slice(1) %>%
  select(-dad.id,-dob,-scz)

father <- scz.family %>%
  arrange(dad.id) %>%
  group_by(dad.id) %>% 
  slice(1) %>%
  select(-mom.id,-dob,-scz)

kid <- scz.kid %>%
  arrange(kid.id) %>%
  group_by(kid.id) %>% 
  slice(1) %>%
  select(-dob, -scz)

sibs <- family %>%
  filter(!is.na(dad.id) & !is.na(mom.id)) %>%
  add_count(dad.id, mom.id) %>%
  filter(n > 1) %>%
  rename(sib.id=id)

sibs.scz <- scz.family %>% # take only relationships with mom and dad LOPNRs included
  filter(!is.na(dad.id) & !is.na(mom.id)) %>%
  add_count(dad.id, mom.id) %>%
  filter(n > 1) %>%
  rename(sib.aff.scz = n)

sibs.scz <- left_join(sibs.scz,sibs)

sibs.scz <- sibs.scz %>%
  filter(id != sib.id) # remove duplicates

###############################
###### DISORDER SPECIFIC ###### 
###############################

#######################  
# ADHD ################
#######################
adhd <- hdr %>% 
  filter(  (icd.version==9 & grepl("^314|^314A|^314B|^314C|^314J|^314W|^314X", diagnosis)) # no icd codes for ADHD prior to ICD version 9
           | (icd.version==10 & grepl("^F90|^F900|^F901|^F908", diagnosis)) 
  ) %>%
  mutate(adhd = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.adhd.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,adhd,first.adhd.date)

### MOM
mother.adhd <- inner_join(mother, adhd, by = c("mom.id"="id"))
mother.adhd <- mother.adhd %>%
  mutate(mom.adhd = ifelse(first.adhd.date < first_scz_date, 1, 0)) %>% #  only those with adhd dx in mom before scz dx in index case
  arrange(mom.id,desc(mom.adhd)) %>%
  group_by(mom.id) %>%
  slice(1) %>%
  select(mom.adhd,mom.id)

### DAD
father.adhd <- inner_join(father, adhd, by = c("dad.id"="id"))
father.adhd <- father.adhd %>%
  mutate(dad.adhd = ifelse(first.adhd.date < first_scz_date, 1, 0)) %>% #  only those with adhd dx in dad before scz dx in index case
  arrange(dad.id,desc(dad.adhd)) %>%
  group_by(dad.id) %>%
  slice(1) %>%
  select(dad.adhd,dad.id)

## KIDS
kids.adhd <- inner_join(kid, adhd, by = c("kid.id" = "id"))
kids.adhd <- kids.adhd %>%
  mutate(kids.adhd = ifelse(first.adhd.date < first_scz_date, 1, 0)) %>% #  only those with adhd dx in dad before scz dx in index case
  arrange(kid.id,desc(kids.adhd)) %>%
  group_by(kid.id) %>%
  slice(1) %>%
  select(kids.adhd,kid.id)

## SIBS
adhd.sibs <- inner_join(sibs.scz, adhd, by = c("sib.id"="id"))
adhd.sibs <- adhd.sibs %>%
  group_by(id) %>%
  mutate(sibs.adhd = ifelse(first.adhd.date < first_scz_date, 1, 0)) %>% #  only those with adhd dx in sibs before scz dx in index case)
  ungroup() %>%
  select(id, sibs.adhd)

# join back with kids and parents file 
adhd.family <- left_join(scz.family, mother.adhd, by = "mom.id")
adhd.family <- left_join(adhd.family, father.adhd, by = "dad.id")
adhd.kid <- left_join(kid, kids.adhd, by = "kid.id")

adhd.kid <- adhd.kid %>%
  select(id, kid.id, kids.adhd) %>%
  arrange(id, desc(kids.adhd)) %>% # remove multiple entries in kids (i.e. >1 kid)
  group_by(id) %>%
  slice(1)

adhd.family$mom.adhd <- as.integer(adhd.family$mom.adhd)
adhd.family$dad.adhd <- as.integer(adhd.family$dad.adhd)

# add variable that indicates family history yes or no
adhd.family <- adhd.family %>%
  select(id,mom.adhd,dad.adhd) %>%
  mutate(fhx = case_when(mom.adhd==1 | dad.adhd==1 ~ 1,
                         mom.adhd==0 | dad.adhd==0 ~ 0,
                         TRUE ~ NA_real_)) %>%
  select(-mom.adhd, -dad.adhd)

adhd.kid <- adhd.kid %>%
  rename(fhx=kids.adhd)

# merge parents and kids file
adhd.fhx <- full_join(adhd.family, adhd.kid)
adhd.fhx <- full_join(adhd.fhx, adhd.sibs)

adhd.fhx <- adhd.fhx %>%
  mutate(fhx.adhd = case_when((fhx == 1 | sibs.adhd == 1) ~ 1, 
                                 (fhx == 0 | sibs.adhd == 0)~ 0,
                                 TRUE ~ NA_real_)) %>%
  select(id,fhx.adhd) %>%
  arrange(id,desc(fhx.adhd)) %>%
  group_by(id) %>%
  slice(1)

# join scz with fhx
scz <- left_join(scz, adhd.fhx, by = "id")

# remove unneeded dataframes
rm(adhd,adhd.family,adhd.kid,father.adhd,adhd.fhx,kids.adhd,mother.adhd,adhd.sibs)





#######################
# ASD #################
#######################
asd <- hdr %>% 
  filter(  (icd.version==9 & grepl("^299A", diagnosis)) # no icd codes for asd prior to version 9
           | (icd.version==10 & grepl("^F840|^F841|^F845", diagnosis)) 
  ) %>%
  mutate(asd = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.asd.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,asd,first.asd.date)

### MOM
mother.asd <- inner_join(mother, asd, by = c("mom.id"="id"))
mother.asd <- mother.asd %>%
  mutate(mom.asd = ifelse(first.asd.date < first_scz_date, 1, 0)) %>% #  only those with asd dx in mom before scz dx in index case
  arrange(mom.id,desc(mom.asd)) %>%
  group_by(mom.id) %>%
  slice(1) %>%
  select(mom.asd,mom.id)

### DAD
father.asd <- inner_join(father, asd, by = c("dad.id"="id"))
father.asd <- father.asd %>%
  mutate(dad.asd = ifelse(first.asd.date < first_scz_date, 1, 0)) %>% #  only those with asd dx in dad before scz dx in index case
  arrange(dad.id,desc(dad.asd)) %>%
  group_by(dad.id) %>%
  slice(1) %>%
  select(dad.asd,dad.id)

## KIDS
kids.asd <- inner_join(kid, asd, by = c("kid.id" = "id"))
kids.asd <- kids.asd %>%
  mutate(kids.asd = ifelse(first.asd.date < first_scz_date, 1, 0)) %>% #  only those with asd dx in dad before scz dx in index case
  arrange(kid.id,desc(kids.asd)) %>%
  group_by(kid.id) %>%
  slice(1) %>%
  select(kids.asd,kid.id)

## SIBS
asd.sibs <- inner_join(sibs.scz, asd, by = c("sib.id"="id"))
asd.sibs <- asd.sibs %>%
  group_by(id) %>%
  mutate(sibs.asd = ifelse(first.asd.date < first_scz_date, 1, 0)) %>% #  only those with asd dx in sibs before scz dx in index case)
  ungroup() %>%
  select(id, sibs.asd)

# join back with kids and parents file 
asd.family <- left_join(scz.family, mother.asd, by = "mom.id")
asd.family <- left_join(asd.family, father.asd, by = "dad.id")
asd.kid <- left_join(kid, kids.asd, by = "kid.id")

asd.kid <- asd.kid %>%
  select(id, kid.id, kids.asd) %>%
  arrange(id, desc(kids.asd)) %>% # remove multiple entries in kids (i.e. >1 kid)
  group_by(id) %>%
  slice(1)

asd.family$mom.asd <- as.integer(asd.family$mom.asd)
asd.family$dad.asd <- as.integer(asd.family$dad.asd)

# add variable that indicates family history yes or no
asd.family <- asd.family %>%
  select(id,mom.asd,dad.asd) %>%
  mutate(fhx = case_when(mom.asd==1 | dad.asd==1 ~ 1,
                         mom.asd==0 | dad.asd==0 ~ 0,
                         TRUE ~ NA_real_)) %>%
  select(-mom.asd, -dad.asd)

asd.kid <- asd.kid %>%
  rename(fhx=kids.asd)

# merge parents and kids file
asd.fhx <- full_join(asd.family, asd.kid)
asd.fhx <- full_join(asd.fhx, asd.sibs)

asd.fhx <- asd.fhx %>%
  mutate(fhx.asd = case_when((fhx == 1 | sibs.asd == 1) ~ 1, 
                              (fhx == 0 | sibs.asd == 0)~ 0,
                              TRUE ~ NA_real_)) %>%
  select(id,fhx.asd) %>%
  arrange(id,desc(fhx.asd)) %>%
  group_by(id) %>%
  slice(1)

# join scz with fhx
scz <- left_join(scz, asd.fhx, by = "id")

# remove unneeded dataframes
rm(asd,asd.family,asd.kid,father.asd,asd.fhx,kids.asd,mother.asd,asd.sibs)




#######################
# Anxiety #############
#######################
anx <- hdr %>% 
  filter(  (icd.version==8 & grepl("^300|^3000|^3002", diagnosis)) |
             (icd.version==9 & grepl("^300|^300A|^300C", diagnosis)) |
             (icd.version==10 & grepl("^F40|^F400|^F401|^F402|^F408|^F409|^F41|^F410|^F411|^F412|^F413|^F418|^F419", diagnosis)) 
  ) %>%
  mutate(anx = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.anx.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,anx,first.anx.date)

### MOM
mother.anx <- inner_join(mother, anx, by = c("mom.id"="id"))
mother.anx <- mother.anx %>%
  mutate(mom.anx = ifelse(first.anx.date < first_scz_date, 1, 0)) %>% #  only those with anx dx in mom before scz dx in index case
  arrange(mom.id,desc(mom.anx)) %>%
  group_by(mom.id) %>%
  slice(1) %>%
  select(mom.anx,mom.id)

### DAD
father.anx <- inner_join(father, anx, by = c("dad.id"="id"))
father.anx <- father.anx %>%
  mutate(dad.anx = ifelse(first.anx.date < first_scz_date, 1, 0)) %>% #  only those with anx dx in dad before scz dx in index case
  arrange(dad.id,desc(dad.anx)) %>%
  group_by(dad.id) %>%
  slice(1) %>%
  select(dad.anx,dad.id)

## KIDS
kids.anx <- inner_join(kid, anx, by = c("kid.id" = "id"))
kids.anx <- kids.anx %>%
  mutate(kids.anx = ifelse(first.anx.date < first_scz_date, 1, 0)) %>% #  only those with anx dx in dad before scz dx in index case
  arrange(kid.id,desc(kids.anx)) %>%
  group_by(kid.id) %>%
  slice(1) %>%
  select(kids.anx,kid.id)

## SIBS
anx.sibs <- inner_join(sibs.scz, anx, by = c("sib.id"="id"))
anx.sibs <- anx.sibs %>%
  group_by(id) %>%
  mutate(sibs.anx = ifelse(first.anx.date < first_scz_date, 1, 0)) %>% #  only those with anx dx in sibs before scz dx in index case)
  ungroup() %>%
  select(id, sibs.anx)

# join back with kids and parents file 
anx.family <- left_join(scz.family, mother.anx, by = "mom.id")
anx.family <- left_join(anx.family, father.anx, by = "dad.id")
anx.kid <- left_join(kid, kids.anx, by = "kid.id")

anx.kid <- anx.kid %>%
  select(id, kid.id, kids.anx) %>%
  arrange(id, desc(kids.anx)) %>% # remove multiple entries in kids (i.e. >1 kid)
  group_by(id) %>%
  slice(1)

anx.family$mom.anx <- as.integer(anx.family$mom.anx)
anx.family$dad.anx <- as.integer(anx.family$dad.anx)

# add variable that indicates family history yes or no
anx.family <- anx.family %>%
  select(id,mom.anx,dad.anx) %>%
  mutate(fhx = case_when(mom.anx==1 | dad.anx==1 ~ 1,
                         mom.anx==0 | dad.anx==0 ~ 0,
                         TRUE ~ NA_real_)) %>%
  select(-mom.anx, -dad.anx)

anx.kid <- anx.kid %>%
  rename(fhx=kids.anx)

# merge parents and kids file
anx.fhx <- full_join(anx.family, anx.kid)
anx.fhx <- full_join(anx.fhx, anx.sibs)

anx.fhx <- anx.fhx %>%
  mutate(fhx.anx = case_when((fhx == 1 | sibs.anx == 1) ~ 1, 
                              (fhx == 0 | sibs.anx == 0)~ 0,
                              TRUE ~ NA_real_)) %>%
  select(id,fhx.anx) %>%
  arrange(id,desc(fhx.anx)) %>%
  group_by(id) %>%
  slice(1)

# join scz with fhx
scz <- left_join(scz, anx.fhx, by = "id")

# remove unneeded dataframes
rm(anx,anx.family,anx.kid,father.anx,anx.fhx,kids.anx,mother.anx,anx.sibs)

#######################
# BIP #################
#######################
bp <- hdr %>%
  filter( ((icd.version==8) & grepl('^2961|^2963|^2968',diagnosis)) |
          ((icd.version==9) & grepl('^296|^269A|^269C|^2691|^296D|^2694|^296E|^2965|^296F|^2966|^296G|^2697|^269H|^2691|^26989|^269W|^26999|^269X',diagnosis)) | 
            ((icd.version==10) & grepl('^F30$|^F301|^F302|^F308|^F309|^F31|^F310|^F311|^F312|^F313|^F314|^F315|^F316|^F317|^F318|^F319',diagnosis)) ) %>%
  mutate(bip = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.bip.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,bip,first.bip.date)


### MOM
mother.bp <- inner_join(mother, bp, by = c("mom.id"="id"))
mother.bp <- mother.bp %>%
  mutate(mom.bp = ifelse(first.bip.date < first_scz_date, 1, 0)) %>% #  only those with bp dx in mom before scz dx in index case
  arrange(mom.id,desc(mom.bp)) %>%
  group_by(mom.id) %>%
  slice(1) %>%
  select(mom.bp,mom.id)

### DAD
father.bp <- inner_join(father, bp, by = c("dad.id"="id"))
father.bp <- father.bp %>%
  mutate(dad.bp = ifelse(first.bip.date < first_scz_date, 1, 0)) %>% #  only those with bp dx in dad before scz dx in index case
  arrange(dad.id,desc(dad.bp)) %>%
  group_by(dad.id) %>%
  slice(1) %>%
  select(dad.bp,dad.id)

## KIDS
kids.bp <- inner_join(kid, bp, by = c("kid.id" = "id"))
kids.bp <- kids.bp %>%
  mutate(kids.bp = ifelse(first.bip.date < first_scz_date, 1, 0)) %>% #  only those with bp dx in dad before scz dx in index case
  arrange(kid.id,desc(kids.bp)) %>%
  group_by(kid.id) %>%
  slice(1) %>%
  select(kids.bp,kid.id)

## SIBS
bp.sibs <- inner_join(sibs.scz, bp, by = c("sib.id"="id"))
bp.sibs <- bp.sibs %>%
  group_by(id) %>%
  mutate(sibs.bp = ifelse(first.bip.date < first_scz_date, 1, 0)) %>% #  only those with bip dx in sibs before scz dx in index case)
  ungroup() %>%
  select(id, sibs.bp)

# join back with kids and parents file 
bp.family <- left_join(scz.family, mother.bp, by = "mom.id")
bp.family <- left_join(bp.family, father.bp, by = "dad.id")
bp.kid <- left_join(kid, kids.bp, by = "kid.id")

bp.kid <- bp.kid %>%
  select(id, kid.id, kids.bp) %>%
  arrange(id, desc(kids.bp)) %>% # remove multiple entries in kids (i.e. >1 kid)
  group_by(id) %>%
  slice(1)

bp.family$mom.bp <- as.integer(bp.family$mom.bp)
bp.family$dad.bp <- as.integer(bp.family$dad.bp)

# add variable that indicates family history yes or no
bp.family <- bp.family %>%
  select(id,mom.bp,dad.bp) %>%
  mutate(fhx = case_when(mom.bp==1 | dad.bp==1 ~ 1,
                         mom.bp==0 | dad.bp==0 ~ 0,
                         TRUE ~ NA_real_)) %>%
  select(-mom.bp, -dad.bp)

bp.kid <- bp.kid %>%
  rename(fhx=kids.bp)

# merge parents and kids file
bp.fhx <- full_join(bp.family, bp.kid)
bp.fhx <- full_join(bp.fhx, bp.sibs)

bp.fhx <- bp.fhx %>%
  mutate(fhx.bp = case_when((fhx == 1 | sibs.bp == 1) ~ 1, 
                              (fhx == 0 | sibs.bp == 0)~ 0,
                              TRUE ~ NA_real_)) %>%
  select(id,fhx.bp) %>%
  arrange(id,desc(fhx.bp)) %>%
  group_by(id) %>%
  slice(1)

# join scz with fhx
scz <- left_join(scz, bp.fhx, by = "id")

# remove unneeded dataframes
rm(bp,bp.family,bp.kid,father.bp,bp.fhx,kids.bp,mother.bp,bp.sibs)

#######################
# Eating Disorders ####
#######################
ed <- hdr %>%
  filter(  (icd.version==8 & grepl("^7840|^3065", diagnosis))
           | (icd.version==9 & grepl("^307B|^307F", diagnosis))
           | (icd.version==10 & grepl("^F50|^F500|^F501|^F502|^F503|^F509", diagnosis)) 
  ) %>%
  mutate(ed = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.ed.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,ed,first.ed.date)


### MOM
mother.ed <- inner_join(mother, ed, by = c("mom.id"="id"))
mother.ed <- mother.ed %>%
  mutate(mom.ed = ifelse(first.ed.date < first_scz_date, 1, 0)) %>% #  only those with ed dx in mom before scz dx in index case
  arrange(mom.id,desc(mom.ed)) %>%
  group_by(mom.id) %>%
  slice(1) %>%
  select(mom.ed,mom.id)

### DAD
father.ed <- inner_join(father, ed, by = c("dad.id"="id"))
father.ed <- father.ed %>%
  mutate(dad.ed = ifelse(first.ed.date < first_scz_date, 1, 0)) %>% #  only those with ed dx in dad before scz dx in index case
  arrange(dad.id,desc(dad.ed)) %>%
  group_by(dad.id) %>%
  slice(1) %>%
  select(dad.ed,dad.id)

## KIDS
kids.ed <- inner_join(kid, ed, by = c("kid.id" = "id"))
kids.ed <- kids.ed %>%
  mutate(kids.ed = ifelse(first.ed.date < first_scz_date, 1, 0)) %>% #  only those with ed dx in dad before scz dx in index case
  arrange(kid.id,desc(kids.ed)) %>%
  group_by(kid.id) %>%
  slice(1) %>%
  select(kids.ed,kid.id)

## SIBS
ed.sibs <- inner_join(sibs.scz, ed, by = c("sib.id"="id"))
ed.sibs <- ed.sibs %>%
  group_by(id) %>%
  mutate(sibs.ed = ifelse(first.ed.date < first_scz_date, 1, 0)) %>% #  only those with ED dx in sibs before scz dx in index case)
  ungroup() %>%
  select(id, sibs.ed)

# join back with kids and parents file 
ed.family <- left_join(scz.family, mother.ed, by = "mom.id")
ed.family <- left_join(ed.family, father.ed, by = "dad.id")
ed.kid <- left_join(kid, kids.ed, by = "kid.id")

ed.kid <- ed.kid %>%
  select(id, kid.id, kids.ed) %>%
  arrange(id, desc(kids.ed)) %>% # remove multiple entries in kids (i.e. >1 kid)
  group_by(id) %>%
  slice(1)

ed.family$mom.ed <- as.integer(ed.family$mom.ed)
ed.family$dad.ed <- as.integer(ed.family$dad.ed)

# add variable that indicates family history yes or no
ed.family <- ed.family %>%
  select(id,mom.ed,dad.ed) %>%
  mutate(fhx = case_when(mom.ed==1 | dad.ed==1 ~ 1,
                         mom.ed==0 | dad.ed==0 ~ 0,
                         TRUE ~ NA_real_)) %>%
  select(-mom.ed, -dad.ed)

ed.kid <- ed.kid %>%
  rename(fhx=kids.ed)

# merge parents and kids file
ed.fhx <- full_join(ed.family, ed.kid)
ed.fhx <- full_join(ed.fhx, ed.sibs)

ed.fhx <- ed.fhx %>%
  mutate(fhx.ed = case_when((fhx == 1 | sibs.ed == 1) ~ 1, 
                              (fhx == 0 | sibs.ed == 0)~ 0,
                              TRUE ~ NA_real_)) %>%
  select(id,fhx.ed) %>%
  arrange(id,desc(fhx.ed)) %>%
  group_by(id) %>%
  slice(1)

# join scz with fhx
scz <- left_join(scz, ed.fhx, by = "id")

# remove unneeded dataframes
rm(ed,ed.family,ed.kid,father.ed,ed.fhx,kids.ed,mother.ed,ed.sibs)


#######################
# MDD #################
#######################

mdd <- hdr %>% 
  filter(  (icd.version==8 & grepl("^2960|^2968|^2969|^2980|^3004", diagnosis)) |
            (icd.version==9 & grepl("^296C|^296D|^311", diagnosis)) |
             (icd.version==10 & grepl("^F32|^F320|^F321|^F322|^F323|^F328|^F329|^F33|^F330|^F331|^F332|^F333|^F334|^F338|^F339", diagnosis)) 
  ) %>%
  mutate(mdd = 1,
       admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.mdd.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,mdd,first.mdd.date)


### MOM
mother.mdd <- inner_join(mother, mdd, by = c("mom.id"="id"))
mother.mdd <- mother.mdd %>%
  mutate(mom.mdd = ifelse(first.mdd.date < first_scz_date, 1, 0)) %>% #  only those with mdd dx in mom before scz dx in index case
  arrange(mom.id,desc(mom.mdd)) %>%
  group_by(mom.id) %>%
  slice(1) %>%
  select(mom.mdd,mom.id)

### DAD
father.mdd <- inner_join(father, mdd, by = c("dad.id"="id"))
father.mdd <- father.mdd %>%
  mutate(dad.mdd = ifelse(first.mdd.date < first_scz_date, 1, 0)) %>% #  only those with mdd dx in dad before scz dx in index case
  arrange(dad.id,desc(dad.mdd)) %>%
  group_by(dad.id) %>%
  slice(1) %>%
  select(dad.mdd,dad.id)

## KIDS
kids.mdd <- inner_join(kid, mdd, by = c("kid.id" = "id"))
kids.mdd <- kids.mdd %>%
  mutate(kids.mdd = ifelse(first.mdd.date < first_scz_date, 1, 0)) %>% #  only those with mdd dx in dad before scz dx in index case
  arrange(kid.id,desc(kids.mdd)) %>%
  group_by(kid.id) %>%
  slice(1) %>%
  select(kids.mdd,kid.id)

## SIBS
mdd.sibs <- inner_join(sibs.scz, mdd, by = c("sib.id"="id"))
mdd.sibs <- mdd.sibs %>%
  group_by(id) %>%
  mutate(sibs.mdd = ifelse(first.mdd.date < first_scz_date, 1, 0)) %>% #  only those with mdd dx in sibs before scz dx in index case)
  ungroup() %>%
  select(id, sibs.mdd)

# join back with kids and parents file 
mdd.family <- left_join(scz.family, mother.mdd, by = "mom.id")
mdd.family <- left_join(mdd.family, father.mdd, by = "dad.id")
mdd.kid <- left_join(kid, kids.mdd, by = "kid.id")

mdd.kid <- mdd.kid %>%
  select(id, kid.id, kids.mdd) %>%
  arrange(id, desc(kids.mdd)) %>% # remove multiple entries in kids (i.e. >1 kid)
  group_by(id) %>%
  slice(1)

mdd.family$mom.mdd <- as.integer(mdd.family$mom.mdd)
mdd.family$dad.mdd <- as.integer(mdd.family$dad.mdd)

# add variable that indicates family history yes or no
mdd.family <- mdd.family %>%
  select(id,mom.mdd,dad.mdd) %>%
  mutate(fhx = case_when(mom.mdd==1 | dad.mdd==1 ~ 1,
                         mom.mdd==0 | dad.mdd==0 ~ 0,
                         TRUE ~ NA_real_)) %>%
  select(-mom.mdd, -dad.mdd)

mdd.kid <- mdd.kid %>%
  rename(fhx=kids.mdd)

# merge parents and kids file
mdd.fhx <- full_join(mdd.family, mdd.kid)
mdd.fhx <- full_join(mdd.fhx, mdd.sibs)

mdd.fhx <- mdd.fhx %>%
  mutate(fhx.mdd = case_when((fhx == 1 | sibs.mdd == 1) ~ 1, 
                              (fhx == 0 | sibs.mdd == 0)~ 0,
                              TRUE ~ NA_real_)) %>%
  select(id,fhx.mdd) %>%
  arrange(id,desc(fhx.mdd)) %>%
  group_by(id) %>%
  slice(1)

# join scz with fhx
scz <- left_join(scz, mdd.fhx, by = "id")

# remove unneeded dataframes
rm(mdd,mdd.family,mdd.kid,father.mdd,mdd.fhx,kids.mdd,mother.mdd,mdd.sibs)


#######################
# OCD #################
#######################
ocd <- hdr %>%
   filter((icd.version==8 & grepl("^3003", diagnosis))
      | (icd.version==9 & grepl("^300D", diagnosis))
      | (icd.version==10 & grepl("^F42|^F420|^F421|^F422|^F428|^F429", diagnosis))) %>%
  mutate(ocd = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.ocd.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,ocd,first.ocd.date)


### MOM
mother.ocd <- inner_join(mother, ocd, by = c("mom.id"="id"))
mother.ocd <- mother.ocd %>%
  mutate(mom.ocd = ifelse(first.ocd.date < first_scz_date, 1, 0)) %>% #  only those with ocd dx in mom before scz dx in index case
  arrange(mom.id,desc(mom.ocd)) %>%
  group_by(mom.id) %>%
  slice(1) %>%
  select(mom.ocd,mom.id)

### DAD
father.ocd <- inner_join(father, ocd, by = c("dad.id"="id"))
father.ocd <- father.ocd %>%
  mutate(dad.ocd = ifelse(first.ocd.date < first_scz_date, 1, 0)) %>% #  only those with ocd dx in dad before scz dx in index case
  arrange(dad.id,desc(dad.ocd)) %>%
  group_by(dad.id) %>%
  slice(1) %>%
  select(dad.ocd,dad.id)

## KIDS
kids.ocd <- inner_join(kid, ocd, by = c("kid.id" = "id"))
kids.ocd <- kids.ocd %>%
  mutate(kids.ocd = ifelse(first.ocd.date < first_scz_date, 1, 0)) %>% #  only those with ocd dx in dad before scz dx in index case
  arrange(kid.id,desc(kids.ocd)) %>%
  group_by(kid.id) %>%
  slice(1) %>%
  select(kids.ocd,kid.id)

## SIBS
ocd.sibs <- inner_join(sibs.scz, ocd, by = c("sib.id"="id"))
ocd.sibs <- ocd.sibs %>%
  group_by(id) %>%
  mutate(sibs.ocd = ifelse(first.ocd.date < first_scz_date, 1, 0)) %>% #  only those with ocd dx in sibs before scz dx in index case)
  ungroup() %>%
  select(id, sibs.ocd)

# join back with kids and parents file 
ocd.family <- left_join(scz.family, mother.ocd, by = "mom.id")
ocd.family <- left_join(ocd.family, father.ocd, by = "dad.id")
ocd.kid <- left_join(kid, kids.ocd, by = "kid.id")

ocd.kid <- ocd.kid %>%
  select(id, kid.id, kids.ocd) %>%
  arrange(id, desc(kids.ocd)) %>% # remove multiple entries in kids (i.e. >1 kid)
  group_by(id) %>%
  slice(1)

ocd.family$mom.ocd <- as.integer(ocd.family$mom.ocd)
ocd.family$dad.ocd <- as.integer(ocd.family$dad.ocd)

# add variable that indicates family history yes or no
ocd.family <- ocd.family %>%
  select(id,mom.ocd,dad.ocd) %>%
  mutate(fhx = case_when(mom.ocd==1 | dad.ocd==1 ~ 1,
                         mom.ocd==0 | dad.ocd==0 ~ 0,
                         TRUE ~ NA_real_)) %>%
  select(-mom.ocd, -dad.ocd)

ocd.kid <- ocd.kid %>%
  rename(fhx=kids.ocd)

# merge parents and kids file
ocd.fhx <- full_join(ocd.family, ocd.kid)
ocd.fhx <- full_join(ocd.fhx, ocd.sibs)

ocd.fhx <- ocd.fhx %>%
  mutate(fhx.ocd = case_when((fhx == 1 | sibs.ocd == 1) ~ 1, 
                              (fhx == 0 | sibs.ocd == 0)~ 0,
                              TRUE ~ NA_real_)) %>%
  select(id,fhx.ocd) %>%
  arrange(id,desc(fhx.ocd)) %>%
  group_by(id) %>%
  slice(1)

# join scz with fhx
scz <- left_join(scz, ocd.fhx, by = "id")

# remove unneeded dataframes
rm(ocd,ocd.family,ocd.kid,father.ocd,ocd.fhx,kids.ocd,mother.ocd,ocd.sibs)


#######################
# PTSD ################
#######################
ptsd <- hdr %>% 
  filter(  (icd.version==8 & grepl("^307", diagnosis)) |
             (icd.version==9 & grepl("^308|^308A|^308B|^308C|^308D|^308E|^308X|^309|^309A|^309B|^309C|^309D|^309E|^309W|^309X", diagnosis)) |
             (icd.version==10 & grepl("^F43|^F430|^F431|^F432|^F438|^F439", diagnosis)) 
  ) %>%
  mutate(ptsd = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.ptsd.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,ptsd,first.ptsd.date)

### MOM
mother.ptsd <- inner_join(mother, ptsd, by = c("mom.id"="id"))
mother.ptsd <- mother.ptsd %>%
  mutate(mom.ptsd = ifelse(first.ptsd.date < first_scz_date, 1, 0)) %>% #  only those with ptsd dx in mom before scz dx in index case
  arrange(mom.id,desc(mom.ptsd)) %>%
  group_by(mom.id) %>%
  slice(1) %>%
  select(mom.ptsd,mom.id)

### DAD
father.ptsd <- inner_join(father, ptsd, by = c("dad.id"="id"))
father.ptsd <- father.ptsd %>%
  mutate(dad.ptsd = ifelse(first.ptsd.date < first_scz_date, 1, 0)) %>% #  only those with ptsd dx in dad before scz dx in index case
  arrange(dad.id,desc(dad.ptsd)) %>%
  group_by(dad.id) %>%
  slice(1) %>%
  select(dad.ptsd,dad.id)

## KIDS
kids.ptsd <- inner_join(kid, ptsd, by = c("kid.id" = "id"))
kids.ptsd <- kids.ptsd %>%
  mutate(kids.ptsd = ifelse(first.ptsd.date < first_scz_date, 1, 0)) %>% #  only those with ptsd dx in dad before scz dx in index case
  arrange(kid.id,desc(kids.ptsd)) %>%
  group_by(kid.id) %>%
  slice(1) %>%
  select(kids.ptsd,kid.id)

## SIBS
ptsd.sibs <- inner_join(sibs.scz, ptsd, by = c("sib.id"="id"))
ptsd.sibs <- ptsd.sibs %>%
  group_by(id) %>%
  mutate(sibs.ptsd = ifelse(first.ptsd.date < first_scz_date, 1, 0)) %>% #  only those with ptsd dx in sibs before scz dx in index case)
  ungroup() %>%
  select(id, sibs.ptsd)

# join back with kids and parents file 
ptsd.family <- left_join(scz.family, mother.ptsd, by = "mom.id")
ptsd.family <- left_join(ptsd.family, father.ptsd, by = "dad.id")
ptsd.kid <- left_join(kid, kids.ptsd, by = "kid.id")

ptsd.kid <- ptsd.kid %>%
  select(id, kid.id, kids.ptsd) %>%
  arrange(id, desc(kids.ptsd)) %>% # remove multiple entries in kids (i.e. >1 kid)
  group_by(id) %>%
  slice(1)

ptsd.family$mom.ptsd <- as.integer(ptsd.family$mom.ptsd)
ptsd.family$dad.ptsd <- as.integer(ptsd.family$dad.ptsd)

# add variable that indicates family history yes or no
ptsd.family <- ptsd.family %>%
  select(id,mom.ptsd,dad.ptsd) %>%
  mutate(fhx = case_when(mom.ptsd==1 | dad.ptsd==1 ~ 1,
                         mom.ptsd==0 | dad.ptsd==0 ~ 0,
                         TRUE ~ NA_real_)) %>%
  select(-mom.ptsd, -dad.ptsd)

ptsd.kid <- ptsd.kid %>%
  rename(fhx=kids.ptsd)

# merge parents and kids file
ptsd.fhx <- full_join(ptsd.family, ptsd.kid)
ptsd.fhx <- full_join(ptsd.fhx, ptsd.sibs)

ptsd.fhx <- ptsd.fhx %>%
  mutate(fhx.ptsd = case_when((fhx == 1 | sibs.ptsd == 1) ~ 1, 
                              (fhx == 0 | sibs.ptsd == 0)~ 0,
                              TRUE ~ NA_real_)) %>%
  select(id,fhx.ptsd) %>%
  arrange(id,desc(fhx.ptsd)) %>%
  group_by(id) %>%
  slice(1)

# join scz with fhx
scz <- left_join(scz, ptsd.fhx, by = "id")

# remove unneeded dataframes
rm(ptsd,ptsd.family,ptsd.kid,father.ptsd,ptsd.fhx,kids.ptsd,mother.ptsd,ptsd.sibs)


#######################
# Substance abuse #####
#######################
suds <- hdr %>%
  filter(  (icd.version==8 & grepl("^303|^3030|^3031|^3032|^3039|^304|^3040|^3041|^3042|^3043|^3044|^3045|^3046|^3047|^3048|^3049", diagnosis))
           | (icd.version==9 & grepl("^303|^303A|^303X|^304|^304A|^304B|^304C|^304D|^304D|^304E|^304F|^304G|^304H|^304W|^304X|^305A|^305X", diagnosis))
           | (icd.version==10 & grepl("^F10|^F100|^F101|^F102|^F103|^F104|^F105|^F106|^F107|^F108|^F109|^F11|^F110|^F111|^F112|^F113|^F114|^F115|^F116|^F117|^F118|^F119|
                               F19|^F190|^F191|^F192|^F192|^F194|^F195|^F196|^F197|^F198|^F199", diagnosis)) 
  ) %>%
  mutate(suds = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.suds.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,suds,first.suds.date)

### MOM
mother.suds <- inner_join(mother, suds, by = c("mom.id"="id"))
mother.suds <- mother.suds %>%
  mutate(mom.suds = ifelse(first.suds.date < first_scz_date, 1, 0)) %>% #  only those with suds dx in mom before scz dx in index case
  arrange(mom.id,desc(mom.suds)) %>%
  group_by(mom.id) %>%
  slice(1) %>%
  select(mom.suds,mom.id)

### DAD
father.suds <- inner_join(father, suds, by = c("dad.id"="id"))
father.suds <- father.suds %>%
  mutate(dad.suds = ifelse(first.suds.date < first_scz_date, 1, 0)) %>% #  only those with suds dx in dad before scz dx in index case
  arrange(dad.id,desc(dad.suds)) %>%
  group_by(dad.id) %>%
  slice(1) %>%
  select(dad.suds,dad.id)

## KIDS
kids.suds <- inner_join(kid, suds, by = c("kid.id" = "id"))
kids.suds <- kids.suds %>%
  mutate(kids.suds = ifelse(first.suds.date < first_scz_date, 1, 0)) %>% #  only those with suds dx in dad before scz dx in index case
  arrange(kid.id,desc(kids.suds)) %>%
  group_by(kid.id) %>%
  slice(1) %>%
  select(kids.suds,kid.id)

## SIBS
suds.sibs <- inner_join(sibs.scz, suds, by = c("sib.id"="id"))
suds.sibs <- suds.sibs %>%
  group_by(id) %>%
  mutate(sibs.suds = ifelse(first.suds.date < first_scz_date, 1, 0)) %>% #  only those with suds dx in sibs before scz dx in index case)
  ungroup() %>%
  select(id, sibs.suds)

# join back with kids and parents file 
suds.family <- left_join(scz.family, mother.suds, by = "mom.id")
suds.family <- left_join(suds.family, father.suds, by = "dad.id")
suds.kid <- left_join(kid, kids.suds, by = "kid.id")

suds.kid <- suds.kid %>%
  select(id, kid.id, kids.suds) %>%
  arrange(id, desc(kids.suds)) %>% # remove multiple entries in kids (i.e. >1 kid)
  group_by(id) %>%
  slice(1)

suds.family$mom.suds <- as.integer(suds.family$mom.suds)
suds.family$dad.suds <- as.integer(suds.family$dad.suds)

# add variable that indicates family history yes or no
suds.family <- suds.family %>%
  select(id,mom.suds,dad.suds) %>%
  mutate(fhx = case_when(mom.suds==1 | dad.suds==1 ~ 1,
                         mom.suds==0 | dad.suds==0 ~ 0,
                         TRUE ~ NA_real_)) %>%
  select(-mom.suds, -dad.suds)

suds.kid <- suds.kid %>%
  rename(fhx=kids.suds)

# merge parents and kids file
suds.fhx <- full_join(suds.family, suds.kid)
suds.fhx <- full_join(suds.fhx, suds.sibs)

suds.fhx <- suds.fhx %>%
  mutate(fhx.suds = case_when((fhx == 1 | sibs.suds == 1) ~ 1, 
                              (fhx == 0 | sibs.suds == 0)~ 0,
                              TRUE ~ NA_real_)) %>%
  select(id,fhx.suds) %>%
  arrange(id,desc(fhx.suds)) %>%
  group_by(id) %>%
  slice(1)

# join scz with fhx
scz <- left_join(scz, suds.fhx, by = "id")

# remove unneeded dataframes
rm(suds,suds.family,suds.kid,father.suds,suds.fhx,kids.suds,mother.suds,suds.sibs)


#######################
# Suicides ############
#######################
sui <- hdr %>%
   filter((icd.version==8 & grepl("^E950|^E951|^E952$|^E9520|^E9521|^E9529|^E953|^E954|^E955|^E956|^E957|^E958|^E959", diagnosis))
      | (icd.version==9 & grepl("^E95A|^E95B|^E95C|^E95D|^E95E|^E95F|^E95G|^E95H|^E95W|^E95X", diagnosis))
      | (icd.version==10)) %>%
  mutate(sui = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.sui.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,sui,first.sui.date)

### MOM
mother.sui <- inner_join(mother, sui, by = c("mom.id"="id"))
mother.sui <- mother.sui %>%
  mutate(mom.sui = ifelse(first.sui.date < first_scz_date, 1, 0)) %>% #  only those with sui dx in mom before scz dx in index case
  arrange(mom.id,desc(mom.sui)) %>%
  group_by(mom.id) %>%
  slice(1) %>%
  select(mom.sui,mom.id)

### DAD
father.sui <- inner_join(father, sui, by = c("dad.id"="id"))
father.sui <- father.sui %>%
  mutate(dad.sui = ifelse(first.sui.date < first_scz_date, 1, 0)) %>% #  only those with sui dx in dad before scz dx in index case
  arrange(dad.id,desc(dad.sui)) %>%
  group_by(dad.id) %>%
  slice(1) %>%
  select(dad.sui,dad.id)

## KIDS
kids.sui <- inner_join(kid, sui, by = c("kid.id" = "id"))
kids.sui <- kids.sui %>%
  mutate(kids.sui = ifelse(first.sui.date < first_scz_date, 1, 0)) %>% #  only those with sui dx in dad before scz dx in index case
  arrange(kid.id,desc(kids.sui)) %>%
  group_by(kid.id) %>%
  slice(1) %>%
  select(kids.sui,kid.id)

## SIBS
sui.sibs <- inner_join(sibs.scz, sui, by = c("sib.id"="id"))
sui.sibs <- sui.sibs %>%
  group_by(id) %>%
  mutate(sibs.sui = ifelse(first.sui.date < first_scz_date, 1, 0)) %>% #  only those with sui dx in sibs before scz dx in index case)
  ungroup() %>%
  select(id, sibs.sui)

# join back with kids and parents file 
sui.family <- left_join(scz.family, mother.sui, by = "mom.id")
sui.family <- left_join(sui.family, father.sui, by = "dad.id")
sui.kid <- left_join(kid, kids.sui, by = "kid.id")

sui.kid <- sui.kid %>%
  select(id, kid.id, kids.sui) %>%
  arrange(id, desc(kids.sui)) %>% # remove multiple entries in kids (i.e. >1 kid)
  group_by(id) %>%
  slice(1)

sui.family$mom.sui <- as.integer(sui.family$mom.sui)
sui.family$dad.sui <- as.integer(sui.family$dad.sui)

# add variable that indicates family history yes or no
sui.family <- sui.family %>%
  select(id,mom.sui,dad.sui) %>%
  mutate(fhx = case_when(mom.sui==1 | dad.sui==1 ~ 1,
                         mom.sui==0 | dad.sui==0 ~ 0,
                         TRUE ~ NA_real_)) %>%
  select(-mom.sui, -dad.sui)

sui.kid <- sui.kid %>%
  rename(fhx=kids.sui)

# merge parents and kids file
sui.fhx <- full_join(sui.family, sui.kid)
sui.fhx <- full_join(sui.fhx, sui.sibs)

sui.fhx <- sui.fhx %>%
  mutate(fhx.sui = case_when((fhx == 1 | sibs.sui == 1) ~ 1, 
                              (fhx == 0 | sibs.sui == 0)~ 0,
                              TRUE ~ NA_real_)) %>%
  select(id,fhx.sui) %>%
  arrange(id,desc(fhx.sui)) %>%
  group_by(id) %>%
  slice(1)

# join scz with fhx
scz <- left_join(scz, sui.fhx, by = "id")

# remove unneeded dataframes
rm(sui,sui.family,sui.kid,father.sui,sui.fhx,kids.sui,mother.sui,sui.sibs)


### END OF FAMILY HX

rm(father,mother,kid,scz.kid,sibs,sibs.scz,scz.family,scz.list)

rm(child,family)

```


### Step 6: Exposure-Prior medical history
We are only selecting those with a date (1) before first treatment contact and (2) before age 15 years (indicating childhood):
+ comorbid somatic disorders: autoimmune, infections, cancer, CVD
+ medication use (wait on this-not sure which drugs to focus on)
```{r}
# now that we have the list of SCZ cases, export this list and import to SAS and match up with full HDR for only those SCZ cases (to limit data size)
scz.list <- scz %>%
  select(id)
write.table(scz.list, file="/Volumes/Common/projects/S3/S3_Research/tryggve2/scz.cases.txt",sep="\t",row.names = F)

# import complete HDR for only those SCZ cases

scz.hdr <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/tryggve_psych_v_hdr_scz.tsv", sep="\t") %>%
  as_tibble() %>%
  rename(id=LOPNR,
         admit.date=DATUM, # date of admission
         inpatient=SOURCE, # inpatient (1) or outpatient (2)
         icd.version=ICD, # icd version - 8, 9 or 10
         diagnosis=DIA,# ICD diagnostic code
         #id.admit.num=X_PID, # per individual ID per admission
         dx.code.num=X_DID) %>% # primary vs. other ICD dx codes, i.e. H01 = primary, all others = secondary
         #care.code=MVO) %>%  # MVO of 9 = psych care, all others will not be used
  mutate(inpatient = ifelse(inpatient == 2, 0, 1)) #recode such that inpatient == 1 and outpatient == 0


# make list of scz cases and date of first contact for comparison with psych disorders below
first.scz.date <- scz %>%
  select(id, first_scz_date, dob)

#########################################
## Prior personal medical history (somatic disorders)
#########################################

# Autoimmune diseases from Song et al JAMA 2018 (autoimmune / PTSD overlap; Unnurs icelandic group)

# Diabetes mellitus, insulin dependent icd-8: 250, icd-9: 250, 242A, icd-10: E10
# Autoimmune thyroid disease ICD8: 242.00, 242.09, 244, 245.03; ICD9: 242A, 242X, 244W, 244X, 245C, 245W; ICD10: E03.5, E03.9, E05.0, E05.5, E05.9, E06.3, E06.5
# Addison’s disease ICD8: 255.1	ICD9: 255E, ICD10:	E27.1, E27.2
# Autoimmune polyglandular syndrome	8: 258.10; 9:	258B; 10:	E31.0
# Reactive arthritis (Reiter’s syndrome)	8: 099.3, 711.1	9: 099D, 711A	10: M02.3, M02.8, M02.9
# Rheumatoid arthritis	8: 712.00, 712.10, 712,20, 712,39, 712.38	9: 714A-D, 719D; 10: M05, M06, M08.0, M08.1, M08.2, M08.3, M08.4
# Ankylosing spondylitis	8: 712.4; 9: 720A; 10:	M45
# Polyarteritis nodosa and related condition (Incl. Kawasaki, Churg-Strauss syndrome, etc.)	8: 446, 9:	446A, 446B, 10:	M30
# Thrombotic microangiopathy	8: -; 9:	446G	10: M31.1
# Granulomatosis with polyangiitis (Wegeners’s granulomatosis) 8:	446.20, 9:	446E, 10:	M31.3
# Microscopic polyangiitis 8/9:none; 10:	M31.7
# Henoch-Schonlein purpura	8:287.00; 9:	287A, 10:	D69.0
# Giant cell arteritis/ Polymyalgia rheumatica	8: 446.30,446.31,446.39,717.9	9: 446F, 725	10: M35.3, M31.5, M31.6
# Systemic lupus erythematosus	8: 734.1; 9:	710A; 10:	M32
# Polymyositis/dermatomyositis	8: 716.10, 716.00	,9: 710E, 710D; 10: M33.0, M33.1, M33.2, M33.9
# Systematic sclerosis (scleroderma)	8: 710.0, 734.0; 9:710B; 10:	M34
# Sjögren’s syndrome	8:734.90,9:	710C, 10:	M35.0
# Mixed connective tissue disease	8:734.98, 734.91, 734.99;	9: 710X, 710W;	10: M35.1
# Behcet’s syndrome	8:136.0;9:	136B; 10:	M35.2
# Pemphigus vulgaris	8: 694.00; 9:	694E; 10:	L10.0
# Bullous pemphigoid	8:694.05; 9:	694F;	10: L12
# Dermatitis herpetiformis	8:693.99; 9:	694A; 10:	L13.0
# Psoriasis	8:696	696A, 9:696B; 10: L40
# Alopecia areata 8:704.00	; 9:704A; 10:	L64
# Vitiligo	8:709.05	709A	L80
# Pernicious anemia	281.0	281A	D51.0
# Autoimmune hemolytic anemia	283.90, 283.92	283A	D59.0, D59.1
# Idiopathic thrombocytopenic purpura	287.3	287D	D69.3
# Acute disseminated encephalomyelitis	-	-	G04
# Anti-NMDA receptor encephalitis	-	-	G13.1
# Multiple sclerosis	340	340	G35
# Neuromyelitis optica and ADEM	341	341A	G36
# Guillain-Barré syndrome	354.01	357A	G61.0, G61.1, G61.8, G61.9
# Myasthenia gravis	733.0	358A	G70.0
# Primary biliary cirrhosis	-	571G	K74.3
# Crohn’s disease	563.00	555	K50
# Ulcerative colitis	563.10, 569.02	556	K51
# Coeliac disease	269.00, 269.98	579A	K90.0
# Acute rheumatic fever and chorea	390-392	390-392	I00, I01.0, I01.1, I01.2, I01.8, I01.9, I02.0, I02.9
# Sarcoidosis	135	135	D86
# IgA nephropathy	580,582	580,582	N00,N01,N03,N05

autoimmune <- scz.hdr %>%
   filter((icd.version==8 & grepl("^250|^242|^244|^245|^255|^099|^711|^712|^446|^287|^717|^734|^716|^710|
                                  ^734|^136|^694|^693|^696|^696A|^704|^709|^281|^283|^287|^340|^341|^354|^733
                                  |^563|^569|^269|^390|^391|^392|^135|^580|^582", diagnosis))
      | (icd.version==9 & grepl("^250|^242A|^242X|^244W|^244X|^245C|^245W|^255E|^099D|^711A|^714A|^714B|
                                ^714C|^714D|^720A|^725|^446A|^446B|^446G|^446E|^446F|^287A|^710A|^710B|
                                ^710E|^710D|^710C|^710X|^710W|^136B|^694E|^694F|^694A|^696B|^704A|709A|^281A|^283A
                                |^287D|^340|^341A|^357A|^358A|^517G|^555|^556|^579A|^390|^391|^392|^135|^580|^582", diagnosis))
      | (icd.version==10 & grepl("^E10|^E03|^E05|^E06|^E27|^M02|^M05|^M06|^M08|^M45|^M30|^M31|^M32|^M33|
                                 ^M34|^M35|^D69|^L10|^L12|^L13|^L40|^L64|^L80|^D51|^D59|^D69|^G04|^G13|^G35|^G36|^G61
                                 |^G70|^K74|^K50|^K51|^K90|^I00|^I01|^I02|^D86|^N00|^N01|^N03|^N05", diagnosis))) %>%
  mutate(ai = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.ai.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,ai,first.ai.date)

autoimmune <- inner_join(first.scz.date,autoimmune) # inner join together with scz cases

autoimmune <- autoimmune %>%
  mutate(age.first.ai = as.numeric((first.ai.date - dob)/365.25),
         ai.childhood = ifelse(age.first.ai <= 15, 1, 0), # only those with ai dx before age 15 years
         ai.pre.scz = ifelse(first.ai.date < first_scz_date, 1, 0)  #  only those with ai dx before scz contact
  ) %>%
  select(id,ai.childhood,ai.pre.scz)

scz <- left_join(scz, autoimmune)
rm(autoimmune)
scz <- scz %>%
  mutate(ai.childhood=ifelse(is.na(ai.childhood),0,ai.childhood),
         ai.pre.scz=ifelse(is.na(ai.pre.scz),0,ai.pre.scz))
         
# Cardiovascular disease 
#for icd 8 and 9: https://dimdi.de/static/de/klassi/icd-10-who/historie/icd-vorgaenger/icd-8/ICD-8-Systematik.htm
#for icd 10: use Death register publication by Brooke et al 2017 (J. Ludvigsson is co-author)
cvd <- scz.hdr %>%
   filter((icd.version==8 & grepl("^390|^391|^392|^393|^394|^395|^396|^397|^398|^400|^401|^402|^403|^404|^410|^411|^412|^413|^414|^42|^440|^441|
                                  ^442|^443|^444|^445|^446|^447|^448|^450|^451|^452|^453|^454|^455|^456|^457|^458", diagnosis))
      | (icd.version==9 & grepl("^390|^391|^392|^393|^394|^395|^396|^397|^398|^401|^402|^403|^404|^405|^410|^411|^412|^413|^414|^415|^416|^417|
                                ^418|^419|^42|^44|^451|^452|^453|^454|^455|^456|^457|^458|^459", diagnosis))
      | (icd.version==10 & grepl("^I0|^I1|^I2|^I3|^I4|^I5|^I7|^I8|^I9", diagnosis))) %>%
  mutate(cvd = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.cvd.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,cvd,first.cvd.date)

cvd <- inner_join(first.scz.date,cvd) # inner join together with scz cases

cvd <- cvd %>%
  mutate(age.first.cvd = as.numeric((first.cvd.date - dob)/365.25),
         cvd.childhood = ifelse(age.first.cvd <= 15, 1, 0), # only those with cvd dx before age 15 years
         cvd.pre.scz = ifelse(first.cvd.date < first_scz_date, 1, 0)  #  only those with cvd dx before scz contact
  ) %>%
  select(id,cvd.childhood,cvd.pre.scz)

scz <- left_join(scz, cvd)
rm(cvd)
scz <- scz %>%
  mutate(cvd.childhood=ifelse(is.na(cvd.childhood),0,cvd.childhood),
         cvd.pre.scz=ifelse(is.na(cvd.pre.scz),0,cvd.pre.scz))

# Stroke	I60-I69 
stroke <- scz.hdr %>%
   filter((icd.version==8 & grepl("^430|^431|^432|^433|^434|^435|^436|^437|^438", diagnosis))
      | (icd.version==9 & grepl("^430|^431|^432|^433|^434|^435|^436|^437|^438", diagnosis))
      | (icd.version==10 & grepl("^I6", diagnosis))) %>%
  mutate(strk = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.strk.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,strk,first.strk.date)

stroke <- inner_join(first.scz.date,stroke) # inner join together with scz cases

stroke <- stroke %>%
  mutate(age.first.strk = as.numeric((first.strk.date - dob)/365.25),
         strk.childhood = ifelse(age.first.strk <= 15, 1, 0), # only those with stroke dx before age 15 years
         strk.pre.scz = ifelse(first.strk.date < first_scz_date, 1, 0)  #  only those with stroke dx before scz contact
  ) %>%
  select(id,strk.childhood,strk.pre.scz)

scz <- left_join(scz, stroke)
rm(stroke)
scz <- scz %>%
  mutate(strk.childhood=ifelse(is.na(strk.childhood),0,strk.childhood),
         strk.pre.scz=ifelse(is.na(strk.pre.scz),0,strk.pre.scz))

# Cancer/(broad codes)	ICD8: 140-163,170-174,180-228,230-239
# Cancer/(broad codes)	ICD9: 140-165,170-176,179-208,210-239
# Cancer  ICD10 C00-C97 
cancer <- scz.hdr %>%
   filter((icd.version==8 & grepl("^14|^15|^160|^161|^162|^163|^170|^171|^172|^173|^174|^18|^19|
                                  ^20|^21|^220|^221|^222|^223|^224|^225|^226|^227|^228|^23", diagnosis))
      | (icd.version==9 & grepl("^14|^15|^160|^161|^162|^163|^164|^165|^170|^171|^172|^173|^174|^175|^176|^179|^18|^19|
                                  ^200|^201|^202|^203|^204|^205|^206|^207|^208|^21|^22|^230|^231|^232|^233|^234|^235|^236|^237|^238|^239", diagnosis))
      | (icd.version==10 & grepl("^C0|^C1|^C2|^C3|^C4|^C5|^C6|^C7|^C8|^C9", diagnosis))) %>%
  mutate(cancer = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.cancer.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,cancer,first.cancer.date)

cancer <- inner_join(first.scz.date,cancer) # inner join together with scz cases

cancer <- cancer %>%
  mutate(age.first.cancer = as.numeric((first.cancer.date - dob)/365.25),
         cancer.childhood = ifelse(age.first.cancer <= 15, 1, 0), # only those with cancer dx before age 15 years
         cancer.pre.scz = ifelse(first.cancer.date < first_scz_date, 1, 0)  #  only those with cancer dx before scz contact
  ) %>%
  select(id,cancer.childhood,cancer.pre.scz)

scz <- left_join(scz, cancer)
rm(cancer)
scz <- scz %>%
  mutate(cancer.childhood=ifelse(is.na(cancer.childhood),0,cancer.childhood),
         cancer.pre.scz=ifelse(is.na(cancer.pre.scz),0,cancer.pre.scz))

# Infections and parasitic conditions	
# ICD8: 000-027,030-046,050-057,060-068,070-104, 110-117,120-136
# ICD9: 001-018,020-027,030-042,045-057,060-066,070-088,090-104,110-118,120-139
# ICD10 A00-B99
infect <- scz.hdr %>%
   filter((icd.version==8 & grepl("^00|^01|^020|^021|^022|^023|^024|^025|^026|^027|^03|
                                  ^040|^041|^042|^043|^044|^045|^046|^050|^051|^052|^053|
                                  ^054|^055|^056|^057|^060|^061|^062|^063|^064|^065|^066|^067|
                                  ^068|^07|^08|^09|^100|^101|^102|^103|^104|^110|^111|^112|^113|
                                  ^114|^115|^116|^117|^12|^130|^131|^132|^133|^134|^135|^136", diagnosis))
      | (icd.version==9 & grepl("^001|^010|^011|^012|^013|^014|^015|^016|^017|^018|^020|^021|
                                  ^022|^023|^024|^025|^026|^027|^03|^040|^041|^042|^045|^046|^047|^048|^049|^050|^051|^052|^053|
                                  ^054|^055|^056|^057|^060|^061|^062|^063|^064|^065|^066|^07|^080|^081|^082|^083|^084|^085|^086|
                                  ^087|^088|^09|^100|^101|^102|^103|^104|^110|^111|^112|^113|
                                  ^114|^115|^116|^117|^118|^12|^130|^131|^132|^133|^134|^135|^136|^137|^138|^139", diagnosis))
      | (icd.version==10 & grepl("^A|^B", diagnosis))) %>%
  mutate(infect = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(first.infect.date = first(admit.date)) %>%
  slice(1) %>%
  select(id,infect,first.infect.date)

infect <- inner_join(first.scz.date,infect) # inner join together with scz cases

infect <- infect %>%
  mutate(age.first.infect = as.numeric((first.infect.date - dob)/365.25),
         infect.childhood = ifelse(age.first.infect <= 15, 1, 0), # only those with infect dx before age 15 years
         infect.pre.scz = ifelse(first.infect.date < first_scz_date, 1, 0)  #  only those with infect dx before scz contact
  ) %>%
  select(id,infect.childhood,infect.pre.scz)

scz <- left_join(scz, infect)
rm(infect)
scz <- scz %>%
  mutate(infect.childhood=ifelse(is.na(infect.childhood),0,infect.childhood),
         infect.pre.scz=ifelse(is.na(infect.pre.scz),0,infect.pre.scz))



# Medication use  --- wait until decide on specific drugs
```

### Step 7: Exposure-Education
+ Year 9 grades
+ Educational attainment
```{r}
#################
#### GRADES #####
#################

# find if any duplicates - i.e. individuals had > 1 assessment - select first assessment only
grades.dup <- grades %>%
  group_by(id) %>%
  filter(n()>1)

# in duplicates, select first assesment for any that have >1 assessment and assign a dummy variable to indicate ones to keep
grades.dup <- grades.dup %>%
  arrange(id) %>%
  group_by(id) %>%
  slice(1) %>%
  mutate(dup = 1)

grades.dup.1 <- grades %>%
  group_by(id) %>%
  filter(n()>1)

join <- left_join(grades.dup.1, grades.dup)

join <- join %>%
  mutate(dup = ifelse(is.na(dup), 0, dup))

grades <- left_join(grades, join)

grades <- grades %>%
  filter(is.na(dup) | dup == 1) %>%
  select(-dup)

rm(join,grades.dup,grades.dup.1)

# split between two measure of grading
# note: if avg.grade is = 0 this could be either failed or not participated
# ie you CANNOT tell the difference bw failed and not participated

########## 1988 to 1997 ##################
# standardize avg.grade.fr88 grades88to97
# (1988-1997, min-max=0,0-5,0)

grades88to97 <- grades %>%
  filter(grad.yr >= 1988 & grad.yr <= 1997) %>%
  group_by(grad.yr) %>%
  mutate(z_avg.grade88to97 = scale(avg.grade.fr88, center = TRUE, scale = TRUE)) %>%
  ungroup()

# check distribution
hist(grades88to97$z_avg.grade88to97)
# note: there are 5068 (see below) with "0" recorded for grade
# these are either did not participate or failed

grades88to97 %>% filter(avg.grade.fr88 == 0) %>% summarise(n=n())

########################################

########## 1998 ####################
# Standardize avg.grade.fr98 grades98
# Merit rating (1998-, best 16 subjects, min-max=0-320)

grades98toCurrent <- grades %>%
  filter(grad.yr >= 1998) %>%
  group_by(grad.yr) %>%
  mutate(z_avg.grade98toCur = scale(merit.grade.fr98, center = TRUE, scale = TRUE)) %>%
  ungroup()

# check distribution
hist(grades98toCurrent$z_avg.grade98toCur)
########################################

# merge back together
grades.2 <- full_join(grades88to97, grades98toCurrent)

# any duplicates? there are 9 dups - leave in
grades.2 %>% distinct(id) %>% summarise(n=n())

# how many missing grade year? none 
grades.2 %>% filter(is.na(grad.yr)) %>% summarise(n=n())

# combine two measures into one
grades.2 <- grades.2 %>%
  rename(std.grade = z_avg.grade98toCur) %>%
  mutate(std.grade = ifelse(is.na(std.grade),z_avg.grade88to97, std.grade)) %>%
  select(id, grad.yr, std.grade)

# how many missing grades? 0 missing
grades.2 %>% filter(is.na(std.grade)) %>% summarise(n=n())
hist(grades.2$std.grade)

### merge with studypop file
scz <- left_join(scz, grades.2)

rm(grades,grades88to97,grades98toCurrent,grades.2)

#############################

############################# 
######## Education ##########
############################# 
# standardize by birth year? do later
# specific coding for education level in Sweden: www.scb.se/statistik/UF/UF0506/SUN2000English%20version.doc

# The first digit gives a broad indication of the level of education and corresponds to ‘level’ in ISCED 97. 
# 6	Postgraduate education
# 5	Post-secondary education, two years or longer
# 4	Post-secondary education, less than two years
# 3	Upper secondary education
# 2	Primary and lower secondary education, 9 (or 10) years
# 1	Primary and lower secondary education, less than 9 years
# (0	Pre-primary education)
# 999 empty/missing

education <- education %>%
  mutate(edlevel = case_when(str_detect(edlevel, "^0") ~ "Pre-primary education",
                               str_detect(edlevel, "^1|^2") ~ "Primary and lower secondary education",
                               str_detect(edlevel, "^3") ~ "Upper secondary education",
                               str_detect(edlevel, "^4|^5") ~ "Post-secondary education",
                               str_detect(edlevel, "^6") ~ "Postgraduate education",
                               TRUE ~ NA_character_)
         )


education <- education %>%
  select(id, edlevel)
scz <- left_join(scz,education)

rm(education)

```



### Step : Exposure set: Family of origin
+ parental ages at birth
+ parental SES-education-occupation-income
+ birth order
+ family composition (number of parents/partners, parental criminality-incarceration, divorce, siblings, out-of-home placement)
```{r}
# parental ages at birth
# mother from Medical Birth Register:+ MALDER	Mothers age at child birth (years).
maternal.age <-
# father from link to multigeneration register to get father then link to population register
paternal.age <-

# parental SES-education-occupation-income
# parent education - linking from above education register + multigen reg

# birth order
# family composition (number of parents/partners, divorce, siblings, out-of-home placement)
# multigeneration register for number of parents, kids
  
  
  
```


### Step : Exposure set: Childhood trauma
+ parental death: cause of death + multigen that occurs < 18 year
+ severe medical illness = "severe" would be? age < 18 year?
```{r}


```



### Step 11: Outcomes

need dates for all of these:
+ the number of psychiatric hospitalizations and outpatient contacts - have from hospital discharge register
+ the number of somatic healthcare utilization - have from hospital discharge register:
  + somatic defintion?
+ indication of 2nd or 3rd line treatments (pharmacological augmentation, ECT, TMS, DBS, clozapine) - drug reg for clozapine not sure about ECT/TMS/DBS
+ suicide completed (death reg) or suicide attempt (HDR)

```{r}
# after the first-Scz-date (month/year)-use 15 as 

```


### Step : Cohort selection and observation period
```{r}
# need start date of observation period -- ?
# add flag for died or emigrated during observation time and time of follow up

scz <- scz %>%
  mutate(fu.indicator = case_when(!is.na(dod) ~ "died",
                                  !is.na(first.emigrate.date) ~ "emigrated"),
         obs.time = )


# basic descriptives
scz %>%
  summarise(n=n(), 
            median_obstime = median(obs.time), 
            mean_obstime = mean(obs.time))
scz %>%
  group_by(fu.indicator) %>% # by follow up group
  summarise(n=n(),
            median_obstime = median(obs.time), 
            mean_obstime = mean(obs.time))



# study cohort selection
study <- scz %>% 
  filter(birth.country == "Sweden" &
           substr(as.character(first_scz_date),1,4) %in% c(1973:2014) & age.second.scz >= 18)

```



### Step 12: Analysis: Logistic regression
```{r}
# make sure all variables have no missing data (i.e., assign 'na' if not known)

# Build the clozapine model - univariable
clo_uni_model <- glm(clo ~ relevel(sex,2), data = scz, family = "binomial")
summary(clo_uni_model)
sex<-as.data.frame(exp(cbind("Odds ratio" = coef(clo_uni_model), confint.default(clo_uni_model, level = 0.95)))) # if want to do Forest Plot


# Build the clozapine model - multivariable
clo_model_adj <- glm(clo ~ relevel(sex,2) + age.cohort.entry + sun2000niva.3 + psychinpt + psychoutpt + mdd + bip + sui + sud + weeksAdequate + INJweeks + POLYPHARMweeks + relevel(totalINJ_c,5) + totalORAL_c2 + fhx.scz.any + fhx.bip, data = scz, family = "binomial")
summary(clo_model_adj)
clo_multi_model <-as.data.frame(exp(cbind("Odds ratio" = coef(clo_model_adj), confint.default(clo_model_adj, level = 0.95))))


write.table(clo_multi_model,file="~/x.tsv") #edit in excel

         
#################################################
### FOREST PLOT FOR LOGISTIC REGRESSION #########
#################################################
       
# plot OR and 95%CI
# Create labels for plot
clo_uni <- fread("~/Documents/Projects/1.Tx.Resistance/CRIME3/data/crime3-univar-OR-clovsno.csv") %>%
  as_tibble()

clo_multi <- fread("~/Documents/Projects/1.Tx.Resistance/CRIME3/data/crime3-multivar-OR-clovsno.csv") %>%
  as_tibble() %>%
  rename(aOR=OR,
         aupper=upper,
         alower=lower)

clo <- full_join(clo_uni,clo_multi)

clo$Variable <- factor(clo$Variable, levels=rev(clo$Variable))
clo$adjust <- as.factor(clo$adjust)

dodge <- position_dodge(width=.5)

fp <- ggplot(data=clo) +
        geom_pointrange(aes(x=Variable, y=OR, ymin=lower, ymax=upper, color="Unadjusted"), position=dodge) + 
        geom_pointrange(aes(x=Variable, y=aOR, ymin=alower, ymax=aupper, color="Adjusted"), position=dodge) + 
        geom_hline(yintercept=1, lty=3) +  # add a dotted line at x=1 after flip
        coord_flip() +  # flip coordinates (puts labels on y axis)
        xlab("") + 
        ylab("Odds Ratio (95% CI)") +
        theme_bw() 

print(fp)
ggsave(file="~/CRIME3-logreg-clo.png",fp, width=8, height=5)


```
  
  
  