---
title: "Tryggve2 - Schizophrenia - Sweden - Part 5"
author: "Kaarina Kowalec, kaarina.kowalec@ki.se"
date: "May 2019"
output: 
  html_document: 
    fig_caption: yes
    theme: simplex
    highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Motivation and Overview

Examine variables associated with poor outcomes in schizophrenia, using first,
the Swedish National Registers. In this report, I will present some initial
coding for the analyses.

**Overview**

1. Import data
2. Create a dataframe with all necessary variables including exposures and 
outcomes
3. Analyses (Covered by this RMD)

***

### Step 0: Load R packages
```{r R packages}
library(tidyverse)
library(psych)
set.seed(1)
Sys.setenv(TZ="Europe/Stockholm")
```
## Factor Analysis
### Step 1: Exploratory Factor Analysis
```{r}
sub <- analysis %>%
  select(suicide_yn,pd_yn,hsp_yn,trs_yn)

# Conduct a single-factor EFA
EFA_model <- fa(sub)

# View the results
# loadings range from -1 to +1
EFA_model$loadings

fa.diagram(EFA_model)

head(EFA_model$scores)

plot(density(EFA_model$scores, na.rm = TRUE), 
    main = "Factor Scores")

summary(EFA_model$scores)

cor.test(rowSums(sub), EFA_model$scores)   
#	Pearson's product-moment correlation

#data:  rowSums(sub) and EFA_model$scores
#t = 115.97, df = 13804, p-value < 2.2e-16
#alternative hypothesis: true correlation is not equal to 0
#95 percent confidence interval:
# 0.6939486 0.7108487
#sample estimates:
 #     cor 
#0.7024977 
# correlation matrix using lowerCor() function in psych package

# How many dimensions to use
# first compute the correlation matrix
efa.cor <- cor(sub, use="pairwise.complete.obs")

# calculate eigenvalues based on the correlation matrix
eigenvals <- eigen(efa.cor)

# check eigenvalues 
eigenvals$values # see how many >1  (2 are >1)

# can create the scree plot based on the correlation matrix
scree(efa.cor, factors=F)

# Tetrachoric correlation

tetrachoric(sub)

#Call: tetrachoric(x = sub, y = NULL, correct = 0.5, smooth = TRUE, 
 #   global = TRUE, weight = NULL, na.rm = TRUE, delete = TRUE)
#tetrachoric correlation 
#           scd_y pd_yn hsp_y trs_y
#suicide_yn  1.00                  
#pd_yn       0.40  1.00            
#hsp_yn      0.10  0.14  1.00      
#trs_yn      0.02 -0.25  0.43  1.00

# with tau of 
#suicide_yn      pd_yn     hsp_yn     trs_yn 
 #     1.13       1.32       0.50       0.83 


```
### Step 2: Split into EFA and CFA
```{r}
# Establish two sets of indices to split the dataset
N <- nrow(sub)
indices <- seq(1, N)
indices_EFA <- sample(indices, floor((.5*N)))
indices_CFA <- indices[!(indices %in% indices_EFA)]

# Use those indices to split the dataset into halves for your EFA and CFA
sub_EFA <- sub[indices_EFA, ]
sub_CFA <- sub[indices_CFA, ]

# Use the indices from the previous exercise to create a grouping variable
group_var <- vector("numeric", nrow(sub))
group_var[indices_EFA] <- 1
group_var[indices_CFA] <- 2

# Bind that grouping variable onto the gcbs dataset
sub_grouped <- cbind(sub, group_var)

# Compare stats across groups
describeBy(sub_grouped, group = group_var)
statsBy(sub_grouped, group = "group_var")
```
### Step 3: Correlations & Internal consistency
```{r}
# Take a look at some correlation data
lowerCor(sub, use = "pairwise.complete.obs")
corr.test(sub)$p
corr.test(sub)$ci

# internal consistency
# Estimate coefficient alpha
alpha(sub)
# alpha coefficient = -0.06 (v poor)

# Calculate split-half reliability
splitHalf(sub)
# split-half reliability = 0.5 (v poor) 

```


### Step 4: Other method of EFA
```{r}
pca <- princomp(sub)
summary(pca)
plot(pca)
```


## Cox Regression
### Step 1: Generate AUC 
```{r}
library(survAUC)

sub <- analysis %>%
  filter(!is.na(EduYr.id)) %>%
  filter(!is.na(large_munc_sczdx))

TR <- sub[1:6578,]
TE <- sub[6579:13157,]

# all covariates
train.fit <- coxph(Surv(time_any, outcome_yn) ~ sex + age.first.scz +
                             bip.pre.scz + anx.pre.scz + ptsd.pre.scz + adhd.pre.scz +
                             ed.pre.scz + sui.pre.scz + suds.pre.scz + ocd.pre.scz +
                             asd.pre.scz + mdd.pre.scz + ai.pre.scz + cancer.pre.scz +
                             congen.malform + intdis.pre.scz + large_munc_sczdx +
                             relevel(infect.type,5) + EduYr.id + fhx.scz + birthyear,
                             method="breslow", data=TR)
lp <- predict(train.fit)
lpnew <- predict(train.fit, newdata=TE)
Surv.rsp <- Surv(TR$time_any, TR$outcome_yn)
Surv.rsp.new <- Surv(TE$time_any, TE$outcome_yn)
times <- seq(10, 1000, 10)
AUC_CD <- AUC.cd(Surv.rsp, Surv.rsp.new, lp, lpnew, times)

# specific outcome
train.fit.2 <- coxph(Surv(time_hsp, hsp_yn) ~ sex + age.first.scz +
                             bip.pre.scz + anx.pre.scz + ptsd.pre.scz + adhd.pre.scz +
                             ed.pre.scz + sui.pre.scz + suds.pre.scz + ocd.pre.scz +
                             asd.pre.scz + mdd.pre.scz + congen.malform + intdis.pre.scz +
                             relevel(infect.type,5) + EduYr.id +
                             fhx.scz + birthyear,
                             method="breslow", data=TR)
lp.2 <- predict(train.fit.2)
lpnew.2 <- predict(train.fit.2, newdata=TE)
Surv.rsp.2 <- Surv(TR$time_any, TR$outcome_yn)
Surv.rsp.new.2 <- Surv(TE$time_any, TE$outcome_yn)
times <- seq(10, 1000, 10)

AUC_CD.2 <- AUC.cd(Surv.rsp.2, Surv.rsp.new.2, lp.2, lpnew.2, times)

#PLOT

plot(AUC_CD.2, col="red", type="l", ylim=c(0,1),
     xlab="Time",ylab="AUC",
     main="AUC", add=FALSE)
lines(AUC_CD,col="green")

```

