---
title: "Tryggve2 - Schizophrenia - Sweden - Part III"
author: "Kaarina Kowalec, kaarina.kowalec@ki.se"
date: "November 2018"
output: 
  html_document: 
    fig_caption: yes
    theme: simplex
    highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Motivation and Overview

Examine variables associated with poor outcomes in schizophrenia, using first, the Swedish National Registers. In this report, I will present some initial coding for the analyses.

**Overview**

1. Import data to create list of SCZ cases and other exposures/outcomes
2. Generate list of SCZ cases then export list of id's to pull out full hospital discharge register for only those cases (limit file sizes)
3. Generate exposures
4. Generate outcomes
5. Analyses

**Exposures or Predictors**
1) Family of origin:
+ parental ages at birth ##LY
+ parental SES-education-occupation-income
+ birth order
+ family composition (number of parents/partners, parental criminality-incarceration, divorce, siblings, out-of-home placement)

2) Childhood trauma: ##LY
+ parental death
+ severe medical illness of parents/sibs

3) Education: done-kk
+ educational attainment
+ standardized testing in grade 9

4) Prior personal medical history (kk done, need childhood hospitalisation):
+ comorbid disorders and medication use
+ prior personal medical history (somatic disorders)
+ childhood hospitalization

5) Family history of psychiatric disorder ##kk
+ MD, BIP, psychosis, suicide in first-degree relatives

6) Genetic risk factors  NOT NOW
+ GRS of MDD, SCZ, BIP, IQ etc.

**Outcomes**
+ the number of psychiatric hospitalizations and outpatient contacts 
+ the number of somatic healthcare utilization
+ indication of 2nd or 3rd line treatments (pharmacological augmentation, ECT, TMS, DBS, clozapine) 
+ suicide or suicide attempt

**TO DOs**

cut down this doc into 4 separate RMD (eventually):
outcome
exposure/risk factors
basics
analysis

***


### Step 0: Load R packages
```{r R packages}
library(tidyverse)
library(data.table)
library(lubridate)
library(skimr)

Sys.setenv(TZ="Europe/Stockholm")
```

### Step 1: Outcome 1 and 2: Suicide - attempts or completed
Suicide attempt must have occurred after the first SCZ date and only need first one (for now)
```{r}
# attempts from hospital discharge register

#################################
# Suicides attempts #############
#################################

sui.a <- scz.hdr %>%
   filter((icd.version==8 & grepl("^E950|^E951|^E952$|^E9520|^E9521|^E9529|^E953|^E954|^E955|^E956|^E957|^E958|^E959", diagnosis))
      | (icd.version==9 & grepl("^E95A|^E95B|^E95C|^E95D|^E95E|^E95F|^E95G|^E95H|^E95W|^E95X", diagnosis))
      | (icd.version==10 & grepl("^X60|^X61|^X62|^X63|^X64|^X65|^X66|^X67|^X68|^X69|^X70|^X71|^X72|^X73|^X74|^X75|^X76|^X77|^X78|^X79|^X80|^X81|
                               X82|^X83|^X84", diagnosis)) ) %>%
  mutate(sui = 1,
         admit.date = ymd(as.character(admit.date))) %>%
  arrange(id, admit.date)

sui.a <- inner_join(first.scz.date,sui.a) # inner join together with scz cases to get date of first.scz.admit

sui.a <- sui.a %>%
  filter(admit.date > first_scz_date) %>% # remove sui admission before scz contact   
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(date.outcome1 = first(admit.date),
         outcome1 = 1) %>%
  slice(1) %>%
  select(id, date.outcome1, outcome1)

#################################
# Suicides completed ############
#################################

# completes from death register - need to go back and get this from COD - get entire COD for SCZ (or MDD)

scz.cod <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/tryggve_psych_v_cod_scz.tsv", sep="\t") %>%
  as_tibble() %>%
  rename(id=LOPNR,
         dod=X_DODSDAT,
         icd.version=ICD_NR,
         death.cause=CAUSE,
         death.cause.nr=CAUSENR)

sui.c <- scz.cod %>%
   filter((icd.version==8 & grepl("^E950|^E951|^E952$|^E9520|^E9521|^E9529|^E953|^E954|^E955|^E956|^E957|^E958|^E959", death.cause))
      | (icd.version==9 & grepl("^E95A|^E95B|^E95C|^E95D|^E95E|^E95F|^E95G|^E95H|^E95W|^E95X", death.cause))
      | (icd.version==10 & grepl("^X60|^X61|^X62|^X63|^X64|^X65|^X66|^X67|^X68|^X69|^X70|^X71|^X72|^X73|^X74|^X75|^X76|^X77|^X78|^X79|^X80|^X81|
                               X82|^X83|^X84", death.cause)) ) %>%
  mutate(sui.d = 1,
         death.date = ymd(as.character(dod))) %>%
  arrange(id, death.date)

sui.c <- inner_join(first.scz.date,sui.c) # inner join together with scz cases to get date of first.scz.admit

sui.c <- sui.c %>%
  filter(death.date > first_scz_date) %>% # remove sui death before scz contact **i know, theoretically, impossible
  arrange(id, death.date) %>%
  group_by(id) %>%
  mutate(date.outcome2 = first(death.date),
         outcome2 = 1) %>%
  slice(1) %>%
  select(id, date.outcome2, outcome2)

# merge outcome 1 and 2 with original SCZ dataframe

scz <- left_join(scz, sui.a)
scz <- left_join(scz, sui.c)

rm(sui.a, sui.c)
```


### Step 2: Outcome 3: Premature mortality
+ Mortality which occurs by age 56 years or earlier (any causes). Premature mortality was defined as death before the age of 56 years, as defined by Fazel, Lichtenstein, et al, Lancet Psychiatry 2014.
+ Could also take life expectancy in Sweden for both males and females in 2016 (82.2 years), less 25 years (expected number of years less for someone w SCZ) = 82-25 = 57 years
```{r}
premature <- scz.cod %>%
  arrange(id, dod) %>%
  mutate(death.date = ymd(as.character(dod))) %>%
  group_by(id) %>%
  slice(1) %>%
  select(id,death.date)

premature <- inner_join(first.scz.date,premature) # inner join together with scz cases to get date of first.scz.admit

premature <- premature %>%
  filter(death.date > first_scz_date) %>% # remove deaths before scz contact **i know, theoretically, impossible
  arrange(id, death.date) %>%
  group_by(id) %>%
  filter(as.numeric((death.date - dob)/365.25) <= 56) %>%
  mutate(date.outcome3 = first(death.date),
        outcome3 = 1) %>%
  slice(1) %>%
  select(id, date.outcome3, outcome3)

# merge outcome 3 with original SCZ dataframe 

scz <- left_join(scz, premature)
rm(premature)
```

### Step 3: Outcome 4: Hospitalisations for SCZ

```{r}
scz.hdr.ex <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/tryggve_psych_v_hdr_ex_scz.tsv", sep="\t") %>%
  select(-PSVARD) %>%
  as_tibble() %>%
  rename(admit.num=X_PID,
         id=LOPNR,
         admit.date=DATUM,
         ward.code=MVO,
         inpatient=SOURCE,
         admit.days=VTID,
         #"psych.care"="PSVARD", # quotations bc character variable
         icd.version=ICD) %>%
  mutate(inpatient = ifelse(inpatient == 2, 0, 1), #recode such that inpatient == 1 and outpatient == 0
         admit.date = ymd(as.character(admit.date))) %>%
  filter(grepl("^9", ward.code)) 

# only want to look at SCZ specific admissions but want time in hospital
scz.hdr.in <- scz.hdr %>%
  filter((icd.version != 10 & substr(diagnosis,1,3) == "295") |
          (icd.version == 10 & substr(diagnosis,1,3) == "F20") |
           (icd.version == 10 & substr(diagnosis,1,3) == "F25")) %>%
  filter(inpatient==1) %>%
  arrange(admit.num, id)

scz.hdr.ex <- scz.hdr.ex %>%
  select(id,admit.num,ward.code,admit.days) %>%
  arrange(admit.num, id)

scz.hdr.new <- left_join(scz.hdr.in, scz.hdr.ex) %>% # join to get # days admitted
  filter(!is.na(ward.code)) # keeps only MVO=9 admissions

scz.hdr.new <- inner_join(first.scz.date,scz.hdr.new) # inner join together with scz cases to get date of first.scz.admit

# select only those in the analysis cohort (birth cohort) - want to get an idea
# of number of hospitlisations but only for those in birth cohort
cohort <- analysis %>% # generated in *_part4.rmd
  select(id) %>%
  mutate(studypop=1)
xscz.hdr.new <- inner_join(cohort,scz.hdr.new) %>%
  mutate(admit.days_tiles = ntile(admit.days, 10))

# median value in highest decile is 229 days when selected to only study cohort
# how about defining as first instance that someone has admission for 100 days?
# not specific enough
# how about defining as first instance that someone has admission for 200 days?

outcome4 <- scz.hdr.new %>%
  mutate(admit.date = ymd(as.character(admit.date))) %>%
  filter(admit.date > first_scz_date) %>% # remove hospitalisations before first scz contact or equal to first contact
  filter(admit.days > 200) %>% # keep any hospitalisations > 200 days
  arrange(id, admit.date) %>%
  group_by(id) %>%
  mutate(date.outcome4.new = first(admit.date),
         outcome4.new = 1) %>%
  slice(1) %>%
  select(id, date.outcome4.new, outcome4.new)

scz <- left_join(scz, outcome4)
rm(outcome4)

```
### Step 4: Outcome 5: Treatment resistance - Clozapine only
```{r}
# combine yearly drug data
y05 <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/drug/tryggve_psych_cloz_y05.tsv", sep="\t") %>%
  as_tibble() 
y06 <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/drug/tryggve_psych_cloz_y06.tsv", sep="\t") %>%
  as_tibble() 
y07 <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/drug/tryggve_psych_cloz_y07.tsv", sep="\t") %>%
  as_tibble() 
y08 <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/drug/tryggve_psych_cloz_y08.tsv", sep="\t") %>%
  as_tibble() 
y09 <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/drug/tryggve_psych_cloz_y09.tsv", sep="\t") %>%
  as_tibble() 
y10 <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/drug/tryggve_psych_cloz_y10.tsv", sep="\t") %>%
  as_tibble() 
y11 <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/drug/tryggve_psych_cloz_y11.tsv", sep="\t") %>%
  as_tibble() 
y12 <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/drug/tryggve_psych_cloz_y12.tsv", sep="\t") %>%
  as_tibble() 
y13 <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/drug/tryggve_psych_cloz_y13.tsv", sep="\t") %>%
  as_tibble() 

#dont use 2014, b/c other data only goes up to 12-31-2013
#y14 <- fread("/Volumes/Common/projects/S3/S3_Research/tryggve2/drug/tryggve_psych_cloz_y14.tsv", sep="\t") %>%
 # as_tibble() 

outcome5 <- rbind(y05,y06,y07,y08,y09,y10,y11,y12,y13)
rm(y05,y06,y07,y08,y09,y10,y11,y12,y13)

outcome5 <- outcome5 %>%
  rename(id=LOPNR,
         atc=ATC,
         date.redeem=EDATUM) %>%
  mutate(date.redeem = ymd(as.character(date.redeem)))

xoutcome5 <- inner_join(first.scz.date,outcome5) # inner join together with scz cases to get date of first.scz.admit

# remove anyone w clozapine prescription b4 1st scz contact or at first contact
xoutcome5 <- xoutcome5 %>% 
  arrange(id, date.redeem) %>%
  mutate(cloz.b4.scz = ifelse(date.redeem <= first_scz_date, 1, 0)) %>% 
  arrange(desc(cloz.b4.scz)) %>% 
  group_by(id) %>%
  slice(1) %>%
  filter(cloz.b4.scz == 0) %>%
  select(id,cloz.b4.scz) #select only ID because these are people who did not receive any CLOZ before SCZ dx

outcome5 <- left_join(outcome5,xoutcome5) # to get back the individual prescription level

outcome5 <- outcome5 %>%
  filter(cloz.b4.scz == 0)

outcome5 <- inner_join(first.scz.date,outcome5) %>% # inner join together with scz cases to get date of first.scz.admit
  arrange(id, date.redeem) %>%
  group_by(id) %>%
  mutate(date.outcome5 = first(date.redeem),
         outcome5 = 1) %>%
  slice(1) %>%
  select(id, date.outcome5, outcome5)

rm(xoutcome5)

scz <- left_join(scz, outcome5)

```


### Step 5: Outcome 6: Hospitalisation for somatic disease
Do only cardiovascular disease for now.
```{r}

# Cardiovascular disease 
#for icd 8 and 9: https://dimdi.de/static/de/klassi/icd-10-who/historie/icd-vorgaenger/icd-8/ICD-8-Systematik.htm
#for icd 10: use Death register publication by Brooke et al 2017 (J. Ludvigsson is co-author)
cvd <- scz.hdr %>%
   filter((icd.version==8 & grepl("^390|^391|^392|^393|^394|^395|^396|^397|^398|^400|^401|^402|^403|^404|^410|^411|^412|^413|^414|^42|^440|^441|
                                  ^442|^443|^444|^445|^446|^447|^448|^450|^451|^452|^453|^454|^455|^456|^457|^458", diagnosis))
      | (icd.version==9 & grepl("^390|^391|^392|^393|^394|^395|^396|^397|^398|^401|^402|^403|^404|^405|^410|^411|^412|^413|^414|^415|^416|^417|
                                ^418|^419|^42|^44|^451|^452|^453|^454|^455|^456|^457|^458|^459", diagnosis))
      | (icd.version==10 & grepl("^I0|^I1|^I2|^I3|^I4|^I5|^I7|^I8|^I9", diagnosis))) %>%
  mutate(cvd = 1,
         cvd.admit.date = ymd(as.character(admit.date))) 

cvd <- inner_join(first.scz.date,cvd) # inner join together with scz cases

outcome6 <- cvd %>%
  filter(cvd.admit.date > first_scz_date) %>% # remove hospitalisations before first scz contact or equal to first contact
  arrange(id, cvd.admit.date) %>%
  group_by(id) %>%
  mutate(date.outcome6 = first(cvd.admit.date),
         outcome6 = 1) %>%
  slice(1) %>%
  select(id, date.outcome6, outcome6)

scz <- left_join(scz, outcome6)
rm(outcome6)

```
### Step 6: Outcome 7-9: income support, sickness pension, divorce
1.9 Income and civil status: from LISA register, on a per-year basis
+ Required Variables:
  + unique ID - rename as "id"  
  + civil status
  + income 
  + sickness pension
  + income support
  + year = year of assessment
  + divorce
```{r read in income and civil levels}
lisa <- fread("~/scz.lisa.income.tsv") %>%
  as_tibble() %>%
  mutate(income.sup=ifelse(!is.na(income.sup.yr),1,0)) %>%
  rename(income.sup.date=income.sup.yr,
         sick.pension.date=sick.pension.yr)

lisa$income.sup.date <- sapply(lisa$income.sup.date, paste0, "0101")
lisa$sick.pension.date <- sapply(lisa$sick.pension.date, paste0, "0101")
lisa <- lisa %>%
  mutate(income.sup.date = gsub("NA0101", NA, income.sup.date),
         sick.pension.date = gsub("NA0101", NA, sick.pension.date)
         )

lisa <- lisa %>%
  mutate(income.sup.date = ymd(as.character(income.sup.date)),
         sick.pension.date = ymd(as.character(sick.pension.date))
         ) 

lisa <- inner_join(first.scz.date,lisa) # inner join together with scz cases

outcome7 <- lisa %>%
  filter(income.sup.date > first_scz_date) %>% # remove income support before first scz contact or equal to first contact
  arrange(id, income.sup.date) %>%
  group_by(id) %>%
  mutate(date.outcome7 = first(income.sup.date),
         outcome7 = 1) %>%
  slice(1) %>%
  select(id, date.outcome7, outcome7)

scz <- left_join(scz, outcome7)
rm(outcome7)

outcome8 <- lisa %>%
  filter(sick.pension.date > first_scz_date) %>% # remove sickness pension before first scz contact or equal to first contact
  arrange(id, sick.pension.date) %>%
  group_by(id) %>%
  mutate(date.outcome8 = first(sick.pension.date),
         outcome8 = 1) %>%
  slice(1) %>%
  select(id, date.outcome8, outcome8)

scz <- left_join(scz, outcome8)
rm(outcome8)

##### divorce ######

div <- fread("~/scz.lisa.div.tsv") %>%
  as_tibble()

div$divorce.yr <- sapply(div$divorce.yr , paste0, "0101")
div <- div %>%
  mutate(divorce.yr = gsub("NA0101", NA, divorce.yr)
         )

div <- div %>%
  mutate(divorce.date = ymd(as.character(divorce.yr))) 

div <- inner_join(first.scz.date,div) # inner join together with scz cases

outcome9 <- div %>%
  filter(divorce.date > first_scz_date) %>% # remove divorce before first scz contact or equal to first contact
  arrange(id, divorce.date) %>%
  group_by(id) %>%
  mutate(date.outcome9 = first(divorce.date),
         outcome9 = 1) %>%
  slice(1) %>%
  select(id, date.outcome9, outcome9)

scz <- left_join(scz, outcome9)
rm(outcome9,div)
```
