---
title: "Tryggve2 - Major depressive disorder - Sweden (part 1) "
author: "Lu Yi, lu.yi@ki.se"
date: "October 2018"
output: 
  html_document: 
    fig_caption: yes
    theme: simplex
    highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Motivation and Overview

NOTE: NOT RUN

Examine variables associated with poor outcomes in schizophrenia, using first, the Swedish National Registers. In this report, I will present some initial coding for the analyses.

**Overview**

1. Import data
2. Create a dataframe with all necessary variables including exposures and outcomes
3. Analyses

**Exposures or Predictors**
1) Family of origin:
+ parental ages at birth ##LY
+ parental SES-education-occupation-income
+ birth order
+ family composition (number of parents/partners, parental criminality-incarceration, divorce, siblings, out-of-home placement)

2) Childhood trauma: ##LY
+ parental death
+ severe medical illness of parents/sibs

3) Education: done-kk
+ educational attainment
+ standardized testing in grade 9

4) Prior personal medical history:
+ comorbid disorders and medication use ##kk
+ prior personal medical history (somatic disorders) #kk
+ childhood hospitalization

5) Family history of psychiatric disorder ##kk
+ MD, BIP, psychosis, suicide in first-degree relatives

6) Genetic risk factors  NOT NOW
+ GRS of MDD, SCZ, BIP, IQ etc.

**Outcomes**
+ the number of psychiatric hospitalizations and outpatient contacts 
+ the number of somatic healthcare utilization
+ indication of 2nd or 3rd line treatments (pharmacological augmentation, ECT, TMS, DBS, clozapine) 
+ suicide or suicide attempt

**TO DOs**

cut down this doc into 4 separate RMD (eventually):
outcome
exposure/risk factors
basics
analysis

***

### Step 0: Load R packages
```{r R packages}
library(dplyr)
library(data.table)
library(ggplot2)
library(lubridate)

Sys.setenv(TZ="Europe/Stockholm")
```

### Step 1: Load data

Essential registers:
1. Hospital discharge data (hdr): all in-/out-patient hospitalisation codes, multiple lines per person, contains date of admission/discharge
2. Population register (pop): birth date, sex, birth country and Swedish birth county 
3. Death register (death): date of death, causes of death
4. Migration register (migrate): date of emigration/immigration
5. Multi-generation register (family): personnummer of index case and personnummer of child/parents
   - SWE: two files, parent & child
6. Medical birth register (birth): containing birth outcomes
7. Education (education): from LISA register; Svensk utbildningsnomenklatur (SUN) or Swedish education nomenclature

Other registers:
8. Grade 9 grades (grades): need?
9. Drug register (drug): All prescriptions redeemed between 7/2005-12/2014, with ATC codes/drug names)
10. Civil register (civil) - need

Add full path for each dataset <- REQUIRE EDITING on the file path
```{r specify file path}
# Required:
hdr.f <- "/Volumes/tryggve_psych_v_npr.tsv"
pop.f <- "/Volumes/tryggve_psych_v_indiv.tsv"
death.f <- "/Volumes/tryggve_psych_v_death.tsv"
migrate.f <- "/Volumes/tryggve_psych_v_migrate.tsv"
parent.f <- "/Volumes/tryggve_psych_v_parent.tsv"
child.f <- "/Volumes/tryggve_psych_v_child.tsv"
birth.f <- "/Volumes/tryggve_psych_v_mbr.tsv"
education.f <- "/Volumes/tryggve_psych_v_education.tsv"

# Other:
grades.f <- "/Volumes/tryggve_psych_v_grades.tsv"

```

Read in register data & Recode variables <- REQUIRE EDITING on the original variable names!

1.1 Hospital discharge data (hdr)
+ Required Variables:  
    + unique ID - rename as "id"  
    + date of admission - rename as "admit.date"
    + SOURCE: inpatient (=1) / outpatient (=2) - recode & rename as "inpatient" (inpatient=1; outpatient=0)
    + ICD version - rename as "icd.version"
    + ICD code - rename as "diagnosis"
    + primary vs secondary diagnosis - rename as "dx.code.num"
n.b. might also add "MVO" code (=9) indicating dx from psychiatric care
```{r Read in hospital discharge register}
hdr <- fread(hdr.f, sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "admit.date"="DATUM", # date of admission
         "inpatient"="SOURCE", # inpatient (1) or outpatient (2)
         "icd.version"="ICD", # ICD version - 8, 9 or 10
         "diagnosis"="DIA", # ICD diagnostic code
         "dx.code.num"="X_DID") %>% # primary vs. other ICD dx codes, i.e. H01 = primary, all others = secondary 
  mutate(inpatient = ifelse(inpatient == 2, 0, 1)) #recode such that inpatient == 1 and outpatient == 0
```

1.2 Population register (pop)
+ Required Variables:  
  + unique ID - rename as "id"  
  + Sex (1 = male, 2 = female) - rename as "sex"  
  + Birth country - rename as "birth.country"
    + Sweden (0=SWE)
    + Nordic not Sweden (numbers 1-9 for the followings)
    + EU28 not Nordic
    + Europe not EU28 not Nordic
    + Africa
    + North America
    + South America
    + Asia
    + Oceania
    + Soviet union
  + Birth county - rename as "birth.county" <-  drop this variable 
  + Birth year and month (yyyymm) - add 15th as day of the month & rename as "dob"
```{r read in population register}
pop <- fread(pop.f, sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "sex"="KON", # 1 = male, 2 = female
         "birth.country"="FODELSEGRUPP", # country of birth
         "birth.county"="FODELSELAN", # county of birth in each country
         "dob"="FODELSEDATUM") %>%
  select(-birth.county) %>% # drop birth county
  mutate(swe.born = case_when(birth.country == 0 ~ 1,
                              birth.country != 0 ~ 0)) %>% # add an indicator of Sweden born (1=SWE, 0=non-SWE, NAs=unknown)
  mutate(dob = paste(as.character(dob),"15",sep=""))  # add 15th as day of the month

```

1.3 Death register (death)
+ Required Variables:  
  + unique ID - rename as "id"  
  + date of death (yyyymmdd) - rename as "dod"
```{r read in cause of death register}
death <- fread(death.f, sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "dod"="X_DODSDATUM")

# "death.cause"="CAUSE",
# "icd.version"="ICD_NR",
```

1.4 Migration register (migrate): date of emigration/immigration
+ Required Variables:
  + unique ID - rename as "id"  
  + Migration date (yyyymm) - add 15th as day of the month & rename as  "migration date"
  + Migration type (E=emigration, I=immigration) - rename as "migrate.type"
```{r read in migration register}
migrate <- fread(migrate.f, sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "migrate.date"="MDATUM",
         "migrate.type"="MTYP") %>% # (E=emigration, I=immigration)
  mutate(migrate.date = paste(as.character(migrate.date),"15",sep=""))
```

1.5 Multi-generation register (parent + child): personnummer of index case and personnummer of child/parents
+ Required Variables:
  + unique ID - rename as "id"  
  + ID of mother - rename as "mom.id"  
  + ID of father - rename as "dad.id"  
  + ID of child - rename as "kid.id"  
```{r read in multigeneration register}
parent <- fread(parent.f, sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "mom.id"="LOPNRMOR",
         "dad.id"="LOPNRFAR")

child <- fread(child.f, sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "kid.id"="LOPNRBARN")
```

1.6. Medical birth register (birth): containing birth outcomes
+ Required Variables:
  + Individual number (child) - rename as "kid.id"  
  + Individual number (mother) - rename as "mom.id"  
  + Mothers age at child birth (years) - rename as "mom.age.at.birth"  
```{r read in medical birth register}
birth <- fread(birth.f, sep="\t") %>%
  as_tibble() %>%
  rename("mom.id"="LOPNRMOR",
         "kid.id"="LOPNRBARN",
         "mom.age.at.birth"="MALDER")
```

1.7 Education (education): from LISA register
+ Required Variables:
  + unique ID - rename as "id"  
  + highest education level - rename as "edlevel"  
```{r read in education levels}
education <- fread(education.f, sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "edlevel"="SUN2000NIVA")
```

1.8 Grade 9 grades (grades)
+ Required Variables:
  + unique ID - rename as "id"  
  + graduation year - rename as "grad.yr"
  + Average grade (1988-1997) - rename as "avg.grade.fr88"
  + Merit rating (1998-, best 16 subjects) - rename as "merit.grade.fr98"
```{r read in 9th grade scores} 
grades <- fread(grades.f, sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "grad.yr"="AVGAR",
         "avg.grade.fr88"="MEDELBETYG", # MEDELBETYG	Average grade (1988-1997, min-max=0,0-5,0)
         "merit.grade.fr98"="MERITVARDE") #  MERITVARDE	Merit rating (1998-, best 16 subjects, min-max=0-320)
```

Other registers <- SKIP FOR NOW 
```{r read in other}
#civil <- fread("~/civil.tsv",sep="\t") %>%
#  as_tibble() %>%
#  rename("id"="LOPNR",
#         "year"="AR",
#         "civil.status"="CIVIL") %>%
#  mutate(civil.status == case_when(civil.status == 1 ~ "Unmarried",
#                                   civil.status == 2 ~ "Married man",
#                                   civil.status == 3 ~ "Married woman, not cohabiting",
#                                   civil.status == 4 | civil.status == "RP" ~ "Divorced",
#                                   civil.status == 5 ~ "Widow/Widower",
#                                   civil.status == 7 ~ "Married woman, cohabiting",
#                                   civil.status == 8 ~ "Child, <18 years",
#                                   civil.status == 9 ~ "Fosterchild, <18 years",
#                                   civil.status == "EP" ~ "Surviving partner",
#                                   civil.status == "G" ~ "Married",
#                                   civil.status == "OG" ~ "Registered partner",
#                                   civil.status == "SP" ~ "Divorced partner",
#                                   civil.status == "─" ~ "Widow/Widower",
#                                   TRUE ~ NA))

# ONLY AVAILABLE IN SWEDEN
# drug register not need by other countries 
#drug <- fread("~/drug.tsv", sep="\t") %>%
#  as_tibble() %>%
#  rename("id"="LOPNR",
#         "num.pkg"="ANTAL",
#         "atc"="ATC",
#         "ddd"="DDDFORP",
#         "dose"="DOSER", #free text
#         "disp.date"="EDATUM", # (yyyy-mm-dd)
#         "pkg.size"="FORPSTL",
#         "admin.route"="LFORMGRUPP", #  +formula of medication, grouped: 7=cutaneous/transdermal, 14=rectal, 12=unspecified, 9=oral/fast
#         "drug.prod"="PRODUKTNAMN", # product name
#         "strength_alf"="STYRKA_ALF",
#         "strength_enh"="STYRKA_ENH",
#         "strength_num"="STYRKNUMERIC")

```

### Step 2: Select MDD cohort

2.1 Apply ICD codes to select MDD cases 
Note: we start with broad code, will refine the code later
```{r extract MDD cases}
mdd <- hdr %>%
  filter((icd.version == 8 & substr(diagnosis,1,4) == "2960") |
           (icd.version == 8 & substr(diagnosis,1,4) == "3004") |
           (icd.version == 9 & substr(diagnosis,1,4) == "296B") |
           (icd.version == 9 & substr(diagnosis,1,3) == "311") |
           (icd.version == 10 & substr(diagnosis,1,3) == "F32") |
           (icd.version == 10 & substr(diagnosis,1,3) == "F33")) %>%
  mutate(mdd = 1, # add disease status
         mdd_date = ymd(admit.date)) %>% # turn the date (character now) into R date object
  add_count(id) %>%  # record the number of diagnoses per individual into variable "n"
  mutate(nmddcontacts = n, n=NULL) # n=null removes the original n variable
```

2.2 Record dates of first and second MDD admission
```{r record first/second dx dates}
mdd <- mdd %>%
  arrange(id, mdd_date) %>% # sort by ID, and dx date
  group_by(id) %>%  # group by ID
  mutate(first_mdd_date = first(mdd_date),  # record the date of first dx
         second_mdd_date = nth(mdd_date, 2)) %>%  # record the date of second dx
  ungroup()  # ungroup by ID

# if no second_mdd_date, exclude them
```

2.3 Divide inpatient/outpatient dx
```{r inpatient vs outpatient}
# mdd hospitalisations
mddinpt <- mdd %>%
  filter(inpatient == 1) %>% # filter for inpatient record
  add_count(id) %>% # add number of inpatient records per person
  rename(nmddinpt = n) %>% # rename to indicate the number of mdd inpatient records
  arrange(id) %>% # order by id
  group_by(id) %>% # for each id
  slice(1) %>% # select the first row
  select(id, nmddinpt) # need only the variables of ID and the number of mdd inpatient records

# mdd outpatient visits
mddoutpt <- mdd %>%
  filter(inpatient == 0) %>% # filter for outpatient record
  add_count(id) %>% # add number of outpatient records per person
  rename(nmddoutpt = n) %>% # rename to indicate the number of mdd outpatient records
  arrange(id) %>% # order by id
  group_by(id) %>% # for each id
  slice(1) %>%  # select the first row
  select(id, nmddoutpt)# need only the variables of ID and the number of mdd outpatient records

# Q for KK: collapsing to per-indv record will make it difficult to apply filters (e.g. refine ICD codes) later. Skip for now?
#mdd <- mdd %>%
#  arrange(id,mdd_date) %>% # collapse to per-individual record and not per-admission
#  group_by(id) %>%
#  slice(1) %>%
#  select(id, mdd, first_mdd_date, second_mdd_date, nmddcontacts) %>%
#  ungroup()

# add back together
mdd <- left_join(mdd, mddinpt)
mdd <- left_join(mdd, mddoutpt, by = "id")

rm(mddinpt,mddoutpt) # remove un-needed dataframes 

# change the NAs in inp/outp counts into 0
# some of the n_inpatient / n_outpatient = NA because the individual has all inpatient records and no outpatients records or vice-versa for mdd
mdd <- mdd %>%
  mutate(nmddoutpt = ifelse(is.na(nmddoutpt), 0, nmddoutpt),
         nmddinpt = ifelse(is.na(nmddinpt), 0, nmddinpt))

```

### Step 3: Demographics: sex, death, emigration

3.1 Merge MDD cohort with population register 
Add sex and birthdate to the MDD cohort, also derive age at first & second diagnoses
```{r add sex and birth info}
mdd <- left_join(mdd, pop, by="id") %>% 
  mutate(dob = ymd(dob)) %>% # birth date as date object 
  mutate(age.first.mdd = as.integer((first_mdd_date - dob)/365.25), # age at the first & second dx (as integer)
         age.second.mdd = as.integer((second_mdd_date - dob)/365.25)) 

rm(pop) # remove pop register

```

3.2 Merge MDD cohort with Cause of Death register 
Add death date to the MDD cohort, also derive age at death (if deceased)
```{r add death info}
mdd <- left_join(mdd, death, by="id") %>% 
  mutate(dod = ymd(dod)) %>% # death date as date object
  mutate(age.death = as.integer((dod - dob)/365.25)) # age at death, NAs if not deceased

rm(death) # remove death reg
```

3.3 Merge MDD cohort with Emigration register
First restrict the cohort to those born in the respective country
Then add the date of first emigration (since then lost FU)
```{r add emigration info}
# check how many filtered due to foreign born
nrow(mdd) - nrow(mdd %>% filter(swe.born == 1))

# select only those born in respective country
mdd <- mdd %>% 
  filter(swe.born == 1) # apply filter on the indicator variable

emigrate <- migrate %>%
  filter(migrate.type=="E") %>% # select only emigration
  arrange(id, migrate.date) %>% # sort by ID and migration date
  group_by(id) %>% # for each unique ID
  mutate(first.emig.date = ymd(first(migrate.date))) %>% # record the first emigration date
  slice(1) %>% # and select just one record
  ungroup() %>% 
  select(id, first.emig.date)

mdd <- left_join(mdd, emigrate,by="id")

rm(migrate, emigrate) # remove migration register
```

### SAVE ALL THE DATA 
```{r save data}
save.image(file = "../DATA/mdd.part1.RData",compress = TRUE)
```
