---
title: "Tryggve2 - Schizophrenia - Sweden - "
author: "Kaarina Kowalec, kaarina.kowalec@ki.se"
date: "October 2018"
output: 
  html_document: 
    fig_caption: yes
    theme: simplex
    highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Motivation and Overview

NOTE: NOT RUN

Examine variables associated with poor outcomes in schizophrenia, using first, the Swedish National Registers. In this report, I will present some initial coding for the analyses.

**Overview**

1. Import data
2. Create a dataframe with all necessary variables including exposures and outcomes
3. Analyses

**Exposures or Predictors**
1) Family of origin:
+ parental ages at birth ##LY
+ parental SES-education-occupation-income
+ birth order
+ family composition (number of parents/partners, parental criminality-incarceration, divorce, siblings, out-of-home placement)

2) Childhood trauma: ##LY
+ parental death
+ severe medical illness of parents/sibs

3) Education: done-kk
+ educational attainment
+ standardized testing in grade 9

4) Prior personal medical history:
+ comorbid disorders and medication use ##kk
+ prior personal medical history (somatic disorders) #kk
+ childhood hospitalization

5) Family history of psychiatric disorder ##kk
+ MD, BIP, psychosis, suicide in first-degree relatives

6) Genetic risk factors  NOT NOW
+ GRS of MDD, SCZ, BIP, IQ etc.

**Outcomes**
+ the number of psychiatric hospitalizations and outpatient contacts 
+ the number of somatic healthcare utilization
+ indication of 2nd or 3rd line treatments (pharmacological augmentation, ECT, TMS, DBS, clozapine) 
+ suicide or suicide attempt

**TO DOs**

cut down this doc into 4 separate RMD (eventually):
outcome
exposure/risk factors
basics
analysis

***

### Step 1: Load data & packages

1. Hospital discharge data (hdr): all in-/out-patient hospitalisation codes, multiple lines per person, contains date of admission/discharge
2. Population register (pop): birth date, sex, birth country and Swedish birth county 
3. Death register (death): date of death, causes of death
4. Migration register (migrate): date of emigration/immigration
5. Multi-generation register (family): personnummer of index case and personnummer of child/parents
6. Medical birth register (birth): containing birth outcomes
7. Drug register (drug): All prescriptions redeemed between 7/2005-12/2014, with ATC codes/drug names)
8. Education (education): from LISA register; Svensk utbildningsnomenklatur (SUN) or Swedish education nomenclature
9. Grade 9 grades (grades): need?
10. Civil register (civil) - need

```{r warning=FALSE, message=FALSE}
library(dplyr)
library(data.table)
library(ggplot2)
library(lubridate)

Sys.setenv(TZ="Europe/Stockholm")

hdr <- fread("~/xx.tsv", sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "admit.date"="DATUM", # date of admission
         "inpatient"="SOURCE", # inpatient (1) or outpatient (2)
         "icd.version"="ICD",
         "diagnosis"="DIA") %>% # ICD diagnostic code
  mutate(inpatient = ifelse(inpatient == 2, 0, 1)) %>% #recode such that inpatient == 1 and outpatient == 0
  select(-s3cnt) # extra variable from previous study, dont use

pop <- fread("~/pop.tsv", sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "sex"="KON", # 1 = male, 2 = female
         "birth.country"="FODELSEGRUPP",
         "swe.birth.county"="FODELSELAN") %>%
  mutate(swe.birth.county = ifelse(swe.birth.county == 2, 1, swe.birth.county), # account for recategorisation as per Amir
    swe.birth.county = ifelse(swe.birth.county == 11, 12, swe.birth.county),
    swe.birth.county = ifelse(swe.birth.county %in% c(15, 16), 14, swe.birth.county))

pop$birth.country <- as.factor(pop$birth.country)
levels(pop$birth.country) <-
  c(
    "Sweden",
    "Nordic not Sweden",
    "EU28 not Nordic",
    "Europe not EU28 not Nordic",
    "Africa",
    "North America",
    "South America",
    "Asia",
    "Oceania",
    "Soviet union"
  )

# urban vs rural birth?
levels(pop$birth.county) <-
  c("Stockholm County",
    "Uppsala",
    "Sodermanland",
    "Vastergotland",
    "Jonkoping",
    "Kronoberg",
    "Kalmar",
    "Gotland",
    "Blekinge",
    "Skane",
    "Halland",
    "Vastra Gotaland",
    "Varmland",
    "Orebro",
    "Vastmanland",
    "Dalarna",
    "Gavleborg",
    "Vasternorrland",
    "Jamtland",
    "Vasterbotten",
    "Norrbotten"
  )

  
death <- fread("~/death.tsv", sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "death.cause"="CAUSE",
         "icd.version"="ICD_NR",
         "dod"="X_DODSDAT")

migrate <- fread("~/migrate.tsv", sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "migrate.data"="MDATUM",
         "migrate.type"="MTYP") # (E=emigration, I=immigration)

family <- fread("~/family.tsv", sep="\t") %>%
  as_tibble() %>%
  rename("id"=="LOPNR",
         "mom.id"="LOPNRMOR",
         "dad.id"="LOPNRFAR",
         "kid.id"="LOPNRBARN")

birth <- fread("~/birth.tsv", sep="\t") %>%
  as_tibble() %>%
  rename("mom.id"="LOPNRMOR",
         "kid.id"="LOPNRBARN",
         "mom.age.at.birth"="MALDER")

education <- fread("~/education.tsv", sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "edlevel"="SUN2000NIVA")

civil <- fread("~/civil.tsv",sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "year"="AR",
         "civil.status"="CIVIL") %>%
  mutate(civil.status == case_when(civil.status == 1 ~ "Unmarried",
                                   civil.status == 2 ~ "Married man",
                                   civil.status == 3 ~ "Married woman, not cohabiting",
                                   civil.status == 4 | civil.status == "RP" ~ "Divorced",
                                   civil.status == 5 ~ "Widow/Widower",
                                   civil.status == 7 ~ "Married woman, cohabiting",
                                   civil.status == 8 ~ "Child, <18 years",
                                   civil.status == 9 ~ "Fosterchild, <18 years",
                                   civil.status == "EP" ~ "Surviving partner",
                                   civil.status == "G" ~ "Married",
                                   civil.status == "OG" ~ "Registered partner",
                                   civil.status == "SP" ~ "Divorced partner",
                                   civil.status == "─" ~ "Widow/Widower",
                                   TRUE ~ NA))

######################################
######### ONLY AVAILABLE IN SWEDEN ###
######################################

# drug register not need by other countries 
drug <- fread("~/drug.tsv", sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "num.pkg"="ANTAL",
         "atc"="ATC",
         "ddd"="DDDFORP",
         "dose"="DOSER", #free text
         "disp.date"="EDATUM", # (yyyy-mm-dd)
         "pkg.size"="FORPSTL",
         "admin.route"="LFORMGRUPP", #  +formula of medication, grouped: 7=cutaneous/transdermal, 14=rectal, 12=unspecified, 9=oral/fast
         "drug.prod"="PRODUKTNAMN", # product name
         "strength_alf"="STYRKA_ALF",
         "strength_enh"="STYRKA_ENH",
         "strength_num"="STYRKNUMERIC")

grades <- fread("~/grades.tsv", sep="\t") %>%
  as_tibble() %>%
  rename("id"="LOPNR",
         "grad.yr"="AVGAR",
         "avg.grade.fr88"="MEDELBETYG", # MEDELBETYG	Average grade (1988-1997, min-max=0,0-5,0)
         "merit.grade.fr98"="MERITVARDE") #  MERITVARDE	Merit rating (1998-, best 16 subjects, min-max=0-320)
```
### Step 2: Create data frame with SCZ cases, date of first and second SCZ admission and age at these dates

```{r message=FALSE}

scz <- hdr %>%
  filter((ICD != 10 & substr(DIA,1,3) == "295") | # use grep and refine codes over time
          (ICD == 10 & substr(DIA,1,3) == "F20") |
           (ICD == 10 & substr(DIA,1,3) == "F25")) %>%
  mutate(scz = 1,
         scz_date = ymd(as.character(DATUM))) %>% # date of admission with SCZ as discharge dx
  add_count(LOPNR) %>% 
  mutate(nsczcontacts = n, n=NULL) # n=null removes the original n variable

scz <- scz %>%
  arrange(LOPNR, scz_date) %>%
  group_by(LOPNR) %>%
  mutate(first_scz_date = first(scz_date),
         second_scz_date = nth(scz_date, 2),
         age.first.scz = as.numeric((first_scz_date - dob)/365.25),
         mutate(age.second.scz = as.numeric((second_scz_date - dob)/365.25))) %>%
  ungroup()

# if no second_scz_date, exclude them

## scz hospitalisations
## need to add MVO code == 9 (for psychiatric in/outpatient)

sczinpt <- scz %>%
  filter(SOURCE == 1) %>%
  add_count(LOPNR) %>%
  rename(nsczinpt = n) %>% 
  arrange(LOPNR) %>%
  group_by(LOPNR) %>%
  slice(1) %>%
  select(-X_PID, -X_DID, -DIA, -SOURCE, -DATUM, -ICD, -scz_date, -first_scz_date)

sczoutpt <- scz %>%
  filter(SOURCE == 2) %>%
  add_count(LOPNR) %>%
  rename(nsczoutpt = n) %>% 
  arrange(LOPNR) %>%
  group_by(LOPNR) %>%
  slice(1) %>%
  select(-X_PID, -X_DID, -DIA, -SOURCE, -DATUM, -ICD, -scz_date, -first_scz_date)

scz <- left_join(scz, sczinpt)

scz <- left_join(scz, sczoutpt, by = "LOPNR")

scz <- scz %>%
  mutate(nsczcontacts.x = ifelse(is.na(nsczcontacts.x), nsczcontacts.y, nsczcontacts.x)) %>%
  select(-nsczcontacts.y) %>%
  rename(nsczcontacts = nsczcontacts.x) %>%
  mutate(nsczoutpt = ifelse(is.na(nsczoutpt), 0, nsczoutpt)) %>%
  mutate(nsczinpt = ifelse(is.na(nsczinpt), 0, nsczinpt))


```



### Step 3: Demographics: sex, death, emigration

```{r}
###############################
# DOB #########################
###############################

# add 15 to date (14 to those in Feb, birth.yr = yyyymm)
scz$birth.yr <- as.character(scz$birth.yr)

scz <- scz %>%
  mutate(feb.birth = str_extract(birth.yr, "[:digit:]{4}02"))

scz$feb.birth <- sapply(scz$feb.birth, paste0, "14")

scz <- scz %>%
  mutate(feb.birth = gsub("NA14", NA, feb.birth))

scz$birth.yr <- sapply(scz$birth.yr, paste0, "15")

scz <- scz %>%
  mutate(birth.yr = ifelse(!is.na(feb.birth), feb.birth, birth.yr))
 
###############################
# Death #######################
###############################

death <- death %>%
  mutate(deathdate = ymd(as.character(X_DODSDAT))) %>%
  select(-X_DODSDAT)

scz <- left_join(scz, death) # join in data with main population

###############################
# Date of Emigration ##########
###############################

# select only those born in respective countries?

migrate <- migrate %>%
  filter(MTYP == "E")
  
migrate$MDATUM <- as.character(migrate$MDATUM)

migrate <- migrate %>%
  mutate(feb.birth = str_extract(MDATUM, "[:digit:]{4}02"))

migrate$feb.birth <- sapply(migrate$feb.birth, paste0, "14")

migrate <- migrate %>%
  mutate(feb.birth = gsub("NA14", NA, feb.birth))

migrate$MDATUM <- sapply(migrate$MDATUM, paste0, "15")

migrate <- migrate %>%
  mutate(MDATUM = ifelse(!is.na(feb.birth), feb.birth, MDATUM))

migrate <- migrate %>%
  mutate(emig.date = ymd(as.character(MDATUM))) %>%
  select(-MDATUM, -feb.birth)

migrate <- migrate %>%
  arrange(LOPNR, emig.date) %>%
  group_by(LOPNR) %>%
  mutate(emig.date = first(emig.date)) %>%
  ungroup()

migrate <- migrate %>%
  arrange(LOPNR) %>%
  group_by(LOPNR) %>%
  slice(1)

scz <- left_join(scz, migrate)
```
