---
title: "Tryggve2 - Schizophrenia - Sweden - Part 4"
author: "Kaarina Kowalec, kaarina.kowalec@ki.se"
date: "March 2019"
output: 
  html_document: 
    fig_caption: yes
    theme: simplex
    highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Motivation and Overview

NOTE: NOT RUN

Examine variables associated with poor outcomes in schizophrenia, using first, the Swedish National Registers. In this report, I will present some initial coding for the analyses.

**Overview**

1. Import data
2. Create a dataframe with all necessary variables including exposures and outcomes
3. Analyses

***

### Step 0: Load R packages
```{r R packages}
library(tidyverse)
library(survival)
library(broom)
library(kableExtra)
library(lubridate)

Sys.setenv(TZ="Europe/Stockholm")
```

### Step 1: Limit to study cohort
```{r}
# birth cohort & SCZ 2x dx case definition
scz <- scz %>%
  mutate(birthyear = str_extract(dob,"[:digit:]{4}"))

scz$birthyear <- as.numeric(scz$birthyear)

studypop <- scz %>% 
  filter(!is.na(second_scz_date) &
           birth.country == "Sweden" &
           !is.na(swe.birth.county) &
           birthyear %in% c(1958:1993) &
           case.exclu == 0)

# need a variable indicating which outcome was first reached & time to reach 
# need to reshape data such that each individual has multiple rows
# outcomes dataframe

outcomes <- studypop %>%
  select(-date.outcome6, -date.outcome8, -outcome6, -outcome8) %>% 
  #removed date.outcome6-CVD, outcome=sicknesspension-for now
  arrange(id) %>%
  mutate(end.follow = ymd("2014-12-31")) %>%
  gather("date.outcome1","date.outcome2", "date.outcome3", "date.outcome4", 
         "date.outcome5", "date.outcome7", "date.outcome9","dod",
         "first.emigrate.date","end.follow",
         key = "outcome",
         value = "date") %>%
  select(id, outcome1:date, scz:z_edlevel) %>%
  arrange(id,date) %>%
  group_by(id) %>%
  mutate(date.outcome.reached = first(date)) %>%
  select(id, date.outcome.reached, outcome1:outcome9,outcome:first_scz_date) %>%
  slice(1) 

outcomes <- outcomes %>%
  mutate(time = as.numeric(date - first_scz_date)/365.25) %>% # calculate time 
  filter(time >= 0) # remove people who emigrated before reached outcome

outcomes$outcome <- as.factor(outcomes$outcome)

outcomes <- outcomes %>%
  mutate(outcome_di = ifelse(outcome!="end.follow" & outcome!="dod" &
                            outcome!="first.emigrate.date", 1, 0),
         outcome_cat = case_when(outcome=="date.outcome1"~"suicide.attempt",
                             outcome=="date.outcome2"~"suicide.complete",
                             outcome=="date.outcome3"~"premature.mortality",
                             outcome=="date.outcome4"~"hospitalised.100days",
                             outcome=="date.outcome5"~"clozapine.prescrip",
                             outcome=="date.outcome7"~"income.support",
                             outcome=="date.outcome9"~"divorce",             
                             outcome=="dod"~"died(not premature",
                             outcome=="end.follow"~"end.of.follow",
                             outcome=="first.emigrate.date"~"emigrated",
                             TRUE ~ NA_character_)
  ) %>%
  select(id, date.outcome.reached, time, outcome_di,outcome_cat) %>%
  rename("fu.time"="time")
#rm outcome=="date.outcome8"~"sickness.pension",

# merge back w exposures

analysis <- left_join(outcomes,scz)

        
###############################
#CONTINUOUS VARIABLES
studypop.sum.continuous <- analysis %>% 
  group_by(outcome_cat) %>% 
  summarise(
    N=n(), 
    Follow.up.time = round(mean(fu.time),digits=2),
    Number.of.SCZ.contacts = round(mean(nsczcontacts, na.rm=T), digits = 2),
    Number.of.SCZ.inpatient = round(mean(nsczinpt, na.rm=T), digits = 2),
    Number.of.SCZ.outpatient = round(mean(nsczoutpt, na.rm=T), digits = 2),
    Age.at.SCZ.dx = round(mean(age.first.scz, na.rm=T), digits = 2),
    Year9.Grade = round(mean(std.grade, na.rm=T), digits = 2),
    EducationalAttain = round(mean(z_edlevel, na.rm=T), digits = 2),
    Paternal.age = round(mean(paternal_age, na.rm=T), digits = 2),
    Maternal.age = round(mean(maternal_age, na.rm=T), digits = 2)
  ) %>%
  arrange(desc(N)) %>%
  mutate("%" = round((N/13561)*100,digits=2)) %>%
  select(outcome_cat,N,"%","Follow.up.time","Number.of.SCZ.contacts":"Maternal.age")

studypop.sum.continuous <- as.data.frame(t(studypop.sum.continuous))

studypop.sum.continuous <- studypop.sum.continuous %>%
  rename("End of Follow Up"="V1",
         "Long Hospitalisation"="V2",
         "Income support"="V3",
         "Treatment-Resistance"="V4",
         "Suicide attempt"="V5",
         "Premature death"="V6",
         "Divorce"="V7",
         "Suicide completed"="V8",
         "Emigrated"="V9",
         "Died, not premature"="V10")

studypop.sum.continuous = studypop.sum.continuous[-1, ]

studypop.sum.continuous <- studypop.sum.continuous[,c(2,4,5,6,8,3,7,1,9,10)]

kable(studypop.sum.continuous, align = "c", "html") %>%
  kable_styling("striped", full_width = F) %>%
  add_header_above(c(" " = 1, "Outcome reached" = 7, "Did not reach outcomes" = 3)) %>%
  group_rows(" ", 1, 2) %>%
  group_rows("Continuous variables", 3, 8) %>%
  footnote("Based on 13,561 individuals born in Sweden between 1958 and 1993 with schizophrenia or schizoaffective disorder"
  ) 

#CATEGORICAL VARIABLES - combine psych variables
analysis <- analysis %>%
  mutate(psych_childhood = case_when(bip.childhood == 1 |
                                       anx.childhood == 1 |
                                       ptsd.childhood == 1 |
                                       adhd.childhood == 1 |
                                       ed.childhood == 1 |
                                       sui.childhood == 1 |
                                       ocd.childhood == 1 |
                                       suds.childhood == 1 |
                                       asd.childhood == 1 |
                                       mdd.childhood == 1 ~ 1,
                                     bip.childhood == 0 |
                                       anx.childhood == 0 |
                                       ptsd.childhood == 0 |
                                       adhd.childhood == 0 |
                                       ed.childhood == 0 |
                                       sui.childhood == 0 |
                                       ocd.childhood == 0 |
                                       suds.childhood == 0 |
                                       asd.childhood == 0 |
                                       mdd.childhood == 0 ~ 0,
                                     TRUE ~ 0))

analysis <- analysis %>%
  mutate(psych_pre.scz = case_when(bip.pre.scz == 1 |
                                     anx.pre.scz == 1 |
                                     ptsd.pre.scz == 1 |
                                     adhd.pre.scz == 1 |
                                     ed.pre.scz == 1 |
                                     sui.pre.scz == 1 |
                                     ocd.pre.scz == 1 |
                                     suds.pre.scz == 1 |
                                     asd.pre.scz == 1 |
                                     mdd.pre.scz == 1 ~ 1,
                                   bip.pre.scz == 0 |
                                     anx.pre.scz == 0 |
                                     ptsd.pre.scz == 0 |
                                     adhd.pre.scz == 0 |
                                     ed.pre.scz == 0 |
                                     sui.pre.scz == 0 |
                                     ocd.pre.scz == 0 |
                                     suds.pre.scz == 0 |
                                     asd.pre.scz == 0 |
                                     mdd.pre.scz == 0 ~ 0,
                                   TRUE ~ 0))

analysis <- analysis %>%
  mutate(psych_fhx = case_when(fhx.scz == 1 |
                                 fhx.bp == 1 |
                                 fhx.anx == 1 |
                                 fhx.ptsd == 1 |
                                 fhx.adhd == 1 |
                                 fhx.ed == 1 |
                                 fhx.sui == 1 |
                                 fhx.ocd == 1 |
                                 fhx.suds == 1 |
                                 fhx.asd == 1 |
                                 fhx.mdd == 1 ~ 1,
                               fhx.scz == 0 |
                                 fhx.bp == 0 |
                                 fhx.anx == 0 |
                                 fhx.ptsd == 0 |
                                 fhx.adhd == 0 |
                                 fhx.ed == 0 |
                                 fhx.sui == 0 |
                                 fhx.ocd == 0 |
                                 fhx.suds == 0 |
                                 fhx.asd == 0 |
                                 fhx.mdd == 0 ~ 0,
                               TRUE ~ 0)
  )

outcomeN <- analysis %>% 
  group_by(outcome_cat) %>% 
  summarise(
    grpN=n())

sex <- analysis %>%
  filter(sex==1) %>%
  group_by(outcome_cat) %>%
  summarise(Males=n()) 
sex <- left_join(sex,outcomeN)
sex <- sex %>%
  mutate(Males_percent = round((Males/grpN)*100,digits=2),
         Males_percent = sapply(Males_percent,paste0, "%"),
         Males=str_c(Males,Males_percent,sep=", ",collapse = NULL))
         
ai.childhood <- analysis %>%
  filter(ai.childhood==1) %>%
  group_by(outcome_cat) %>%
  summarise(autoimmune.childhood=n()) 
ai.childhood <- left_join(ai.childhood,outcomeN)
ai.childhood <- ai.childhood %>%
  mutate(autoimmune.childhood_percent = round((autoimmune.childhood/grpN)*100,digits=2),
         autoimmune.childhood_percent = sapply(autoimmune.childhood_percent,paste0, "%"),
         autoimmune.childhood=str_c(autoimmune.childhood,autoimmune.childhood_percent,sep=", ",collapse = NULL)
  )

ai.pre.scz <- analysis %>%
  filter(ai.pre.scz==1) %>%
  group_by(outcome_cat) %>%
  summarise(ai.pre.scz=n())
ai.pre.scz <- left_join(ai.pre.scz,outcomeN)
ai.pre.scz <- ai.pre.scz %>%
  mutate(ai.pre.scz_percent = round((ai.pre.scz/grpN)*100,digits=2),
         ai.pre.scz_percent = sapply(ai.pre.scz_percent,paste0, "%"),
         ai.pre.scz=str_c(ai.pre.scz,ai.pre.scz_percent,sep=", ",collapse = NULL)
  )

cvd.childhood <- analysis %>%
  filter(cvd.childhood==1) %>%
  group_by(outcome_cat) %>%
  summarise(cvd.childhood=n()) 
cvd.childhood <- left_join(cvd.childhood,outcomeN)
cvd.childhood <- cvd.childhood %>%
  mutate(cvd.childhood_percent = round((cvd.childhood/grpN)*100,digits=2),
         cvd.childhood_percent = sapply(cvd.childhood_percent,paste0, "%"),
         cvd.childhood=str_c(cvd.childhood,cvd.childhood_percent,sep=", ",collapse = NULL)
  )

cvd.pre.scz <- analysis %>%
  filter(cvd.pre.scz==1) %>%
  group_by(outcome_cat) %>%
  summarise(cvd.pre.scz=n())
cvd.pre.scz <- left_join(cvd.pre.scz,outcomeN)
cvd.pre.scz <- cvd.pre.scz %>%
  mutate(cvd.pre.scz_percent = round((cvd.pre.scz/grpN)*100,digits=2),
         cvd.pre.scz_percent = sapply(cvd.pre.scz_percent,paste0, "%"),
         cvd.pre.scz=str_c(cvd.pre.scz,cvd.pre.scz_percent,sep=", ",collapse = NULL)
  )

#strk.childhood <- analysis %>%
#  filter(strk.childhood==1) %>%
 # group_by(outcome_cat) %>%
#  summarise(strk.childhood=n()) 
#strk.childhood <- left_join(strk.childhood,outcomeN)
#strk.childhood <- strk.childhood %>%
 # mutate(strk.childhood_percent = round((strk.childhood/grpN)*100,digits=2),
   #      strk.childhood_percent = sapply(strk.childhood_percent,paste0, "%"),
  #       strk.childhood=str_c(strk.childhood,strk.childhood_percent,sep=", ",collapse = NULL)
  #)

#strk.pre.scz <- analysis %>%
 # filter(strk.pre.scz==1) %>%
  #group_by(outcome_cat) %>%
  #summarise(strk.pre.scz=n()) 
#strk.pre.scz <- left_join(strk.pre.scz,outcomeN)
#strk.pre.scz <- strk.pre.scz %>%
 # mutate(strk.pre.scz_percent = round((strk.pre.scz/grpN)*100,digits=2),
  #       strk.pre.scz_percent = sapply(strk.pre.scz_percent,paste0, "%"),
   #      strk.pre.scz=str_c(strk.pre.scz,strk.pre.scz_percent,sep=", ",collapse = NULL)
  #)

cancer.childhood <- analysis %>%
  filter(cancer.childhood==1) %>%
  group_by(outcome_cat) %>%
  summarise(cancer.childhood=n()) 
cancer.childhood <- left_join(cancer.childhood,outcomeN)
cancer.childhood <- cancer.childhood %>%
  mutate(cancer.childhood_percent = round((cancer.childhood/grpN)*100,digits=2),
         cancer.childhood_percent = sapply(cancer.childhood_percent,paste0, "%"),
         cancer.childhood=str_c(cancer.childhood,cancer.childhood_percent,sep=", ",collapse = NULL)
  )

cancer.pre.scz <- analysis %>%
  filter(cancer.pre.scz==1) %>%
  group_by(outcome_cat) %>%
  summarise(cancer.pre.scz=n())
cancer.pre.scz <- left_join(cancer.pre.scz,outcomeN)
cancer.pre.scz <- cancer.pre.scz %>%
  mutate(cancer.pre.scz_percent = round((cancer.pre.scz/grpN)*100,digits=2),
         cancer.pre.scz_percent = sapply(cancer.pre.scz_percent,paste0, "%"),
         cancer.pre.scz=str_c(cancer.pre.scz,cancer.pre.scz_percent,sep=", ",collapse = NULL)
  )

infect.childhood <- analysis %>%
  filter(infect.childhood==1) %>%
  group_by(outcome_cat) %>%
  summarise(infect.childhood=n())
infect.childhood <- left_join(infect.childhood,outcomeN)
infect.childhood <- infect.childhood %>%
  mutate(infect.childhood_percent = round((infect.childhood/grpN)*100,digits=2),
         infect.childhood_percent = sapply(infect.childhood_percent,paste0, "%"),
         infect.childhood=str_c(infect.childhood,infect.childhood_percent,sep=", ",collapse = NULL)
  )

infect.pre.scz <- analysis %>%
  filter(infect.pre.scz==1) %>%
  group_by(outcome_cat) %>%
  summarise(infect.pre.scz=n())
infect.pre.scz <- left_join(infect.pre.scz,outcomeN)
infect.pre.scz <- infect.pre.scz %>%
  mutate(infect.pre.scz_percent = round((infect.pre.scz/grpN)*100,digits=2),
         infect.pre.scz_percent = sapply(infect.pre.scz_percent,paste0, "%"),
         infect.pre.scz=str_c(infect.pre.scz,infect.pre.scz_percent,sep=", ",collapse = NULL)
  )

parental.breavement <- analysis %>%
  filter(parental.breavement==1) %>%
  group_by(outcome_cat) %>%
  summarise(parental.bereavement=n())
parental.breavement <- left_join(parental.breavement,outcomeN)
parental.breavement <- parental.breavement %>%
  mutate(parental.breavement_percent = round((parental.bereavement/grpN)*100,digits=2),
         parental.breavement_percent = sapply(parental.breavement_percent,paste0, "%"),
         parental.bereavement=str_c(parental.bereavement,parental.breavement_percent,sep=", ",collapse = NULL)
  )

psych_childhood <- analysis %>%
  filter(psych_childhood==1) %>%
  group_by(outcome_cat) %>%
  summarise(psych_childhood=n())
psych_childhood <- left_join(psych_childhood,outcomeN)
psych_childhood <- psych_childhood %>%
  mutate(psych_childhood_percent = round((psych_childhood/grpN)*100,digits=2),
         psych_childhood_percent = sapply(psych_childhood_percent,paste0, "%"),
         psych_childhood=str_c(psych_childhood,psych_childhood_percent,sep=", ",collapse = NULL)
  )

psych_pre.scz <- analysis %>%
  filter(psych_pre.scz==1) %>%
  group_by(outcome_cat) %>%
  summarise(psych_pre.scz=n())
psych_pre.scz <- left_join(psych_pre.scz,outcomeN)
psych_pre.scz <- psych_pre.scz %>%
  mutate(psych_pre.scz_percent = round((psych_pre.scz/grpN)*100,digits=2),
         psych_pre.scz_percent = sapply(psych_pre.scz_percent,paste0, "%"),
         psych_pre.scz=str_c(psych_pre.scz,psych_pre.scz_percent,sep=", ",collapse = NULL)
  )

psych_fhx <- analysis %>%
  filter(psych_fhx==1) %>%
  group_by(outcome_cat) %>%
  summarise(psych_fhx=n())
psych_fhx <- left_join(psych_fhx,outcomeN)
psych_fhx <- psych_fhx %>%
  mutate(psych_fhx_percent = round((psych_fhx/grpN)*100,digits=2),
         psych_fhx_percent = sapply(psych_fhx_percent,paste0, "%"),
         psych_fhx=str_c(psych_fhx,psych_fhx_percent,sep=", ",collapse = NULL)
  )

studypop.sum.categorical <- full_join(sex, ai.childhood)
studypop.sum.categorical <- full_join(studypop.sum.categorical, cvd.childhood)
studypop.sum.categorical <- full_join(studypop.sum.categorical, ai.pre.scz)
studypop.sum.categorical <- full_join(studypop.sum.categorical, cvd.pre.scz)
#studypop.sum.categorical <- full_join(studypop.sum.categorical, strk.childhood)
#studypop.sum.categorical <- full_join(studypop.sum.categorical, strk.pre.scz)
studypop.sum.categorical <- full_join(studypop.sum.categorical, cancer.pre.scz)
studypop.sum.categorical <- full_join(studypop.sum.categorical, cancer.childhood)

studypop.sum.categorical <- full_join(studypop.sum.categorical, infect.childhood)
studypop.sum.categorical <- full_join(studypop.sum.categorical, infect.pre.scz)
studypop.sum.categorical <- full_join(studypop.sum.categorical, parental.breavement)
studypop.sum.categorical <- full_join(studypop.sum.categorical, psych_childhood)
studypop.sum.categorical <- full_join(studypop.sum.categorical, psych_pre.scz)
studypop.sum.categorical <- full_join(studypop.sum.categorical, psych_fhx)

studypop.sum.categorical <- studypop.sum.categorical %>%
  select(-contains("_percent"))

rm(sex, ai.childhood, ai.pre.scz, cvd.childhood, cvd.pre.scz, cancer.pre.scz,
   cancer.childhood,infect.childhood,infect.pre.scz,
   parental.breavement, psych_childhood, psych_pre.scz, psych_fhx)  

studypop.sum.categorical <- studypop.sum.categorical %>%
  select(outcome_cat, grpN, Males, psych_pre.scz, psych_fhx, psych_childhood, 
         parental.bereavement, autoimmune.childhood, ai.pre.scz, cvd.childhood,
         cvd.pre.scz, cancer.pre.scz, 
         cancer.childhood, infect.childhood, infect.pre.scz)


studypop.sum.categorical <- as.data.frame(t(studypop.sum.categorical))

studypop.sum.categorical <- studypop.sum.categorical %>%
  select(V5,V9,V6,V7,V10,V1,V8,V3,V4,V2)

studypop.sum.categorical <- studypop.sum.categorical %>%
  rename("End of Follow Up"="V5",
         "Long Hospitalisation"="V6",
         "Income support"="V7",
         "Suicide attempt"="V9",
         "Treatment-Resistance"="V1",
         "Premature death"="V8",
         "Divorce"="V3",
         "Suicide completed"="V10",
         "Emigrated"="V4",
         "Died, not premature"="V2")

studypop.sum.categorical = studypop.sum.categorical[-1, ]

studypop.sum.categorical <- studypop.sum.categorical[,c(2,4,5,6,8,3,7,1,9,10)]

kable(studypop.sum.categorical, align = "c", "html") %>%
  kable_styling("striped", full_width = F) %>%
  add_header_above(c(" " = 1, "Outcome reached (N, %)" = 7, "Did not reach outcomes (N, %)" = 3)) %>%
  group_rows("Psychiatric", 3, 6) %>%
  group_rows("Somatic", 7, 14) %>%
  footnote("Based on 13,561 individuals born in Sweden between 1958 and 1993 with schizophrenia or schizoaffective disorder"
  )
# for meeting 01/2019 - only used to make table summarising # ppl reaching
# each outcome
outcomes$outcome_cat <- as.factor(outcomes$outcome_cat)

df <- outcomes %>%
  group_by(outcome_cat) %>%
  summarise(n=n(),
            median_follow_time = median(fu.time)) %>%
  mutate(percent = (n/13561)*100) %>%
  arrange(desc(n))
```
### Step X: Logistic regression
```{r}
# merge df
analysis <- left_join(analysis,outcomes, by="id")

log.reg <- glm(outcome_di ~ psych_pre.scz + sex + age.first.scz +
                psych_fhx + infect.pre.scz + cancer.pre.scz + ai.pre.scz, family="binomial", data=analysis)
summary(log.reg)
log.reg <- tidy(exp(cbind("Odds ratio" = coef(log.reg), confint.default(log.reg, level = 0.95))))

kable(log.reg, align = "c", "html") %>%
  kable_styling("striped", full_width = F)

library(olsrr)
sub <- analysis %>%
  ungroup() %>%
  select(outcome_di,sex,age.first.scz,bip.pre.scz,anx.pre.scz,ptsd.pre.scz,adhd.pre.scz,ed.pre.scz,sui.pre.scz,suds.pre.scz,ocd.pre.scz,asd.pre.scz,mdd.pre.scz,fhx.scz,ai.pre.scz,cancer.pre.scz,infect.pre.scz,paternal_age,maternal_age,parental.breavement,psych_childhood,psych_pre.scz,psych_fhx,fu.time)

sub$sex <- as.factor(sub$sex)
sub$bip.pre.scz <- as.factor(sub$bip.pre.scz)
sub$anx.pre.scz <- as.factor(sub$anx.pre.scz)
sub$ptsd.pre.scz <- as.factor(sub$ptsd.pre.scz)
sub$adhd.pre.scz <- as.factor(sub$adhd.pre.scz)
sub$ed.pre.scz <- as.factor(sub$ed.pre.scz)
sub$suds.pre.scz <- as.factor(sub$suds.pre.scz)
sub$ocd.pre.scz <- as.factor(sub$ocd.pre.scz)
sub$asd.pre.scz <- as.factor(sub$asd.pre.scz)
sub$sui.pre.scz <- as.factor(sub$sui.pre.scz)
sub$mdd.pre.scz <- as.factor(sub$mdd.pre.scz)
sub$fhx.scz <- as.factor(sub$fhx.scz)
sub$ai.pre.scz <- as.factor(sub$ai.pre.scz)
sub$cancer.pre.scz <- as.factor(sub$cancer.pre.scz)
sub$infect.pre.scz <- as.factor(sub$infect.pre.scz)
sub$parental.breavement <- as.factor(sub$parental.breavement)
sub$psych_childhood <- as.factor(sub$psych_childhood)
sub$psych_pre.scz <- as.factor(sub$psych_pre.scz)
sub$psych_fhx <- as.factor(sub$psych_fhx)

model <- lm(outcome_di ~ ., data = sub)
ols_step_forward_p(model)
```
### Step X: Cox regression
```{r}

# Cox regression - unadjusted

analysis <- analysis %>%
  mutate(sex = ifelse(sex==1, "M", "F"))
analysis$sex <- as.factor(analysis$sex)

cox1 <- coxph(Surv(fu.time, outcome_di) ~ psych_pre.scz + sex + age.first.scz +
                psych_fhx, data=analysis)
cox5 <- coxph(Surv(fu.time, outcome_di) ~ z_edlevel + sex + age.first.scz, data=analysis)
cox6 <- coxph(Surv(fu.time, outcome_di) ~ parental.breavement+ sex + age.first.scz, data=analysis)

summary(cox1)
cox1 <- tidy(exp(cbind("HR" = coef(cox1), confint.default(cox1, level = 0.95))))

summary(cox5)
cox5 <- tidy(exp(cbind("HR" = coef(cox5), confint.default(cox5, level = 0.95))))
summary(cox6)
cox6 <- tidy(exp(cbind("HR" = coef(cox6), confint.default(cox6, level = 0.95))))

kable(cox1, align = "c", "html") %>%
  kable_styling("striped", full_width = F)
kable(cox5, align = "c", "html") %>%
  kable_styling("striped", full_width = F)
kable(cox6, align = "c", "html") %>%
  kable_styling("striped", full_width = F)


library(My.stepwise)

predict.list <- c("age.first.scz", "bip.pre.scz", "anx.pre.scz", "adhd.pre.scz", "ptsd.pre.scz","ed.pre.scz","sui.pre.scz","suds.pre.scz","ocd.pre.scz","asd.pre.scz","mdd.pre.scz","ai.pre.scz","cancer.pre.scz","infect.pre.scz","psych_childhood","psych_fhx")
stepmodel <- My.stepwise.coxph(Time = "fu.time", Status = "outcome_di", variable.list = predict.list, data=sub)
```

```{r save data}
save.image(file = "../DATA/scz.part1.RData",compress = TRUE)
```